---
title: "Bayesian Survival Models for Life Insurance Lapses (Filtered Dataset)"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  error: false
  message: false
  echo: true
  daemon: false
format:
  html:
    light: superhero
    dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
---


```{r import_libraries}
#| echo: false
#| message: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(arrow)
library(fs)
library(qs)
library(purrr)
library(furrr)
library(glue)
library(scales)
library(cmdstanr)
library(brms)
library(posterior)
library(bayesplot)
library(tidybayes)
library(splines2)
library(rstanarm)


source("lib_utils.R")
source("lib_brms_hazard.R")
source("lib_survival_modelling.R")
source("lib_cashflow_projection.R")
source("lib_stan_diagnostics.R")


conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "MASS",
    "fitdistrplus")
  )


options(
  width = 80L,
  warn  = 1,
  mc.cores = parallelly::availableCores()
  )

# Configure Arrow for multi-threaded parquet I/O
set_cpu_count(parallelly::availableCores())
set_io_thread_count(parallelly::availableCores())

message(glue(
  "Arrow configured: {arrow::cpu_count()} CPU threads, ",
  "{io_thread_count()} I/O threads"
))

message("write_parquet_compressed() loaded from lib_utils.R (zstd level 3 default)")

stan_modeldir   <- "stan_model"
stan_codedir    <- "stan_code"
predictions_dir <- "predictions"
data_dir        <- "data"


set.seed(42)
stanfit_seed <- 4000

theme_set(theme_cowplot())

plan(multisession)
```


```{r setup_chunk_timing}
#| echo: false
#| message: false

# Enable chunk timing for all chunks
knitr::opts_chunk$set(time_it = TRUE)

# Create a list to store chunk times
chunk_times <- list()

# Create a hook to capture chunk execution times
knitr::knit_hooks$set(time_it = function(before, options) {
  if (before) {
    # Record start time in global environment
    assign(".chunk_start_time", Sys.time(), envir = .GlobalEnv)
  } else {
    # Calculate elapsed time
    elapsed <- difftime(
        Sys.time(), get(".chunk_start_time", envir = .GlobalEnv),
        units = "secs"
        ) |>
      as.numeric()
    
    # Store in global list
    chunk_times <- get("chunk_times", envir = .GlobalEnv)
    chunk_times[[options$label]] <- elapsed
    assign("chunk_times", chunk_times, envir = .GlobalEnv)
    
    # Print timing message
    message(glue("⏱️  Chunk '{options$label}': {round(elapsed, 2)} seconds"))
  }
})
```



# Load Data

This notebook uses a **filtered** dataset that focuses on more recent policies
with lower premiums. The data has been pre-filtered to include only:

- Policies with `prem_ape < 15,000` (lower annual premiums)
- Policies starting from `2005-01-01` onwards (recent inception dates)

This filtering helps us focus on a more homogeneous subset of the portfolio,
potentially improving model performance and interpretability. We use the **small**
version of the filtered dataset for faster iteration and development.

We first load up the filtered data we are using to train our model, as well as
the 'rolled-forward' data that observes the same policies a year later.

```{r load_modelling_data}
#| echo: true

model_training_tbl <- glue("{data_dir}/model_training_filtered_small_tbl.parquet") |>
  read_parquet()

model_training_tbl |> glimpse()
```

We then load up the 'current' data that we use to validate our model.


```{r load_current_data}
#| echo: true

model_rollforward_tbl <- glue("{data_dir}/model_rollforward_filtered_small_tbl.parquet") |>
  read_parquet()

model_rollforward_tbl |> glimpse()
```


## Create Shared Data Objects

These objects are used across all three model iterations and are created once
here to avoid duplication.

```{r create_monthly_dates}
#| echo: true

monthly_tbl <- tibble(
  date = seq(as.Date("2015-01-01"), as.Date("2016-02-01"), by = "month"),
  weeks = difftime(date, as.Date("2015-01-01"), units = "weeks") |> as.numeric()
  )

monthly_tbl |> glimpse()
```


```{r create_inforce_subset}
#| echo: true

inforce_tbl <- model_training_tbl |>
  filter(
    status_at_reference_date == "inforce",
    policy_startdate         >= as.Date("2010-01-01")
    )

message(glue("Inforce policies: {nrow(inforce_tbl)}"))
```


```{r create_policy_cutdown}
#| echo: true

policy_cutdown_tbl <- model_training_tbl |>
  select(policy_id, policy_startdate, policy_enddate, prem_ape)

policy_cutdown_tbl |> glimpse()
```


```{r extract_cashflow_week_count}
#| echo: true

cashflow_week_count <- monthly_tbl |> pull(weeks)

message(glue("Number of cashflow periods: {length(cashflow_week_count)}"))
```


```{r calculate_actual_cashflows}
#| echo: true

actual_cashflows_tbl <- model_rollforward_tbl |>
  semi_join(inforce_tbl, by = "policy_id") |>
  calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date)

actual_cashflows_tbl |> glimpse()
```


```{r summarise_actual_cashflows}
#| echo: true

actual_monthly_tbl <- actual_cashflows_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )

actual_monthly_tbl |> glimpse()
```


```{r calculate_actual_cashflows_recent}
#| echo: true

actual_cashflows_recent_tbl <- model_rollforward_tbl |>
  filter(policy_startdate >= as.Date("2010-01-01")) |>
  semi_join(inforce_tbl, by = "policy_id") |>
  calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date)

actual_cashflows_recent_tbl |> glimpse()
```


```{r summarise_actual_cashflows_recent}
#| echo: true

actual_monthly_recent_tbl <- actual_cashflows_recent_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )

actual_monthly_recent_tbl |> glimpse()
```

Finally, we save some of this data to disk so it can be used again.

```{r save_actual_data}
#| echo: false

actual_cashflow_file <- glue("{data_dir}/filtered_actual_cashflow_tbl.parquet")

actual_cashflows_tbl |> write_parquet_compressed(actual_cashflow_file)
```




# Build Initial Cox Model


## brms Model

```{r filterlapse1_fit_cox_model_brms}
#| echo: true
#| message: false

filterlapse1_stanmodel_name <- "filterlapse1_coxph_brms"

filterlapse1_coxph_brmsfit <- brm(
  policy_lifetime | cens(1-lapsed) ~ gender_life1 + smoker_life1,
  data            = model_training_tbl,
  family          = cox(link = "log"),
  cores           = 4,
  seed            = 421,
  save_warmup     = FALSE,
  backend         = "cmdstanr",
  save_model      = glue("{stan_codedir}/{filterlapse1_stanmodel_name}.stan"),
  file            = glue("{stan_modeldir}/{filterlapse1_stanmodel_name}_brmsfit"),
  output_dir      = stan_modeldir,
  output_basename = filterlapse1_stanmodel_name,

  stan_model_args = list(
    include_paths   =     "stan_code",
    pedantic        =            TRUE,
    dir             =    "stan_model"
    )
  )
```

```{r filterlapse1_fit_cox_model_brms_summary}
#| echo: true

summary(filterlapse1_coxph_brmsfit)
```



## rstanarm Model


```{r filterlapse1_fit_cox_model_rstanarm}
#| echo: true
#| message: false

filterlapse1_stanmodel_name <- "filterlapse1_coxph_stansurv"
filterlapse1_stanmodel_file <- glue("{stan_modeldir}/{filterlapse1_stanmodel_name}.qs")

# Check if fitted model exists, load it if so, otherwise fit new model
if (file_exists(filterlapse1_stanmodel_file)) {
  message(glue("Loading cached model from {filterlapse1_stanmodel_file}"))
  filterlapse1_coxph_stansurv <- qread(filterlapse1_stanmodel_file)

} else {
  message(glue("Fitting new model, will save to {filterlapse1_stanmodel_file}"))
  
  # Flush output to ensure visibility in Podman
  cat("Starting Stan model fitting...\n")
  flush.console()
  
  filterlapse1_coxph_stansurv <- stan_surv(
    Surv(policy_lifetime, lapsed) ~ gender_life1 + smoker_life1,
    data   = model_training_tbl,
    
    # MCMC sampling parameters
    chains = 4,
    iter   = 2000,
    warmup = 1000,
    seed   = 4000,
    cores  = 4,
    
    # Baseline hazard specification
    basehaz     = "ms",  # M-splines (flexible, default)
    basehaz_ops = list(df = 6),  # Degrees of freedom for baseline hazard
    
    # Prior specifications (explicit is better than implicit)
    prior           = rstanarm::normal(location = 0, scale = 2.5), # Weakly informative for coefficients
    prior_intercept = rstanarm::normal(location = 0, scale = 10),  # For baseline hazard
    
    # Enable verbose output for console monitoring (not captured in document)
    refresh = 200  # Print progress every 200 iterations
    )
  
  cat("Model fitting complete.\n")
  flush.console()
  
  # Save fitted model for future use
  filterlapse1_coxph_stansurv |> qsave(filterlapse1_stanmodel_file, preset = "balanced")
  message(glue("Model saved to {filterlapse1_stanmodel_file}"))
}
```

```{r filterlapse1_fit_cox_model_stansurv_summary}
#| echo: true
#| message: true

summary(filterlapse1_coxph_stansurv)
```


## Model Diagnostics

We examine the MCMC diagnostics to ensure the model has converged properly and
the posterior samples are reliable. Key metrics include:

- **Rhat**: Should be < 1.01 (measures convergence across chains)
- **ESS ratio**: Should be > 0.1, ideally > 0.5 (effective sample size)
- **Divergences**: Should be 0 (indicates sampling issues)
- **Trace plots**: Should show good mixing without trends

```{r filterlapse1_model_diagnostics_convergence}
#| echo: true
#| fig-width: 14
#| fig-height: 5

# Check actual parameter names
write_lines("Model parameters:", stdout())
names(coef(filterlapse1_coxph_stansurv)) |> print()

plot_stan_convergence(
  filterlapse1_coxph_stansurv,
  title = "Model 1 Convergence Diagnostics: Gender + Smoker"
  )
```


```{r filterlapse1_model_diagnostics_mixing}
#| echo: true
#| fig-width: 14
#| fig-height: 10

plot_stan_mixing(
  filterlapse1_coxph_stansurv,
  title = "Model 1 Chain Mixing: Gender + Smoker"
  )
```


# Construct Posterior Survival Estimates

We now want to use our fitted model to predict the lapse for the portfolio.

We model the monthly paid premia in the portfolio for the final year for which
we have data. We can then compare our modelled cashflows against the observed
data.


## Extract Posterior Estimates


```{r filterlapse1_construct_posterior_estimates}
#| echo: true

filterlapse1_post_draws_file <- glue("{stan_modeldir}/filterlapse1_post_draws_lst.qs")

# Check if cached posterior draws exist
if (file_exists(filterlapse1_post_draws_file)) {
  message(glue("Loading cached posterior draws from {filterlapse1_post_draws_file}"))
  filterlapse1_post_draws_lst <- qread(filterlapse1_post_draws_file)
  
} else {
  message(glue("Computing posterior draws, will save to {filterlapse1_post_draws_file}"))
  
  filterlapse1_post_draws_lst <- posterior_survfit(
      filterlapse1_coxph_stansurv,
      newdata       = inforce_tbl,
      last_time     = NULL,
      draws         = 2000,
      condition     = TRUE,
      return_matrix = TRUE
      )
  
  # Save for future use
  filterlapse1_post_draws_lst |> qsave(filterlapse1_post_draws_file, preset = "balanced")
  message(glue("Posterior draws saved to {filterlapse1_post_draws_file}"))
}

filterlapse1_post_draws_lst[[1]] |> glimpse()
```


```{r filterlapse1_extract_posterior_survival_estimates}
#| echo: true
#| message: false

filterlapse1_post_survival_file <- glue("{predictions_dir}/filterlapse1_post_survival_tbl.parquet")

# Check if cached posterior survival data exists
if (file_exists(filterlapse1_post_survival_file)) {
  message(glue("Loading cached posterior survival data from {filterlapse1_post_survival_file}"))
  filterlapse1_post_survival_tbl <- read_parquet(filterlapse1_post_survival_file)
  
} else {
  message(glue("Extracting posterior survival data, will save to {filterlapse1_post_survival_file}"))
  
  filterlapse1_post_survival_tbl <- filterlapse1_post_draws_lst |>
    map_dfr(
      extract_stansurv_posterior_survivals,

      .progress = "extract_post_surv"
      ) |>
    filter(time < 5 * 52) |>
    mutate(
      policy_id = inforce_tbl$policy_id[idx],
      
      .before = 1
      ) |>
    select(policy_id, draw_id, time, post_surv_prob = value) |>
    interpolate_survival_data(
      new_times = cashflow_week_count
      ) |>
    group_nest(policy_id, draw_id)
  
  # Save for future use
  filterlapse1_post_survival_tbl |> write_parquet_compressed(filterlapse1_post_survival_file)
  message(glue("Posterior survival data saved to {filterlapse1_post_survival_file}"))
}

filterlapse1_post_survival_tbl |> glimpse()
```



```{r filterlapse1_calculate_sim_lapse_times}
#| echo: true
#| message: false

filterlapse1_sim_lapse_times_file <- glue("{predictions_dir}/filterlapse1_sim_lapse_times_tbl.parquet")

# Check if cached simulated lapse times exist
if (file_exists(filterlapse1_sim_lapse_times_file)) {
  message(glue("Loading cached simulated lapse times from {filterlapse1_sim_lapse_times_file}"))
  filterlapse1_sim_lapse_times_tbl <- read_parquet(filterlapse1_sim_lapse_times_file)
  
} else {
  message(glue("Computing simulated lapse times, will save to {filterlapse1_sim_lapse_times_file}"))
  
  filterlapse1_sim_lapse_times_tbl <- filterlapse1_post_survival_tbl |>
    mutate(
      rnd_draw   = runif(n(), 0, 1),
      lapse_time = map2_dbl(
        data, rnd_draw,
        \(x, y) extract_survival_time(x, y, min_time = 1),

        .progress = "extract_survival_time"
        )
      ) |>
    select(policy_id, draw_id, lapse_time)
  
  # Save for future use
  filterlapse1_sim_lapse_times_tbl |> write_parquet_compressed(filterlapse1_sim_lapse_times_file)
  message(glue("Simulated lapse times saved to {filterlapse1_sim_lapse_times_file}"))
}

filterlapse1_sim_lapse_times_tbl |> glimpse()
```



```{r filterlapse1_construct_policy_sim_data}
#| echo: true

filterlapse1_sim_policy_data_tbl <- filterlapse1_sim_lapse_times_tbl |>
  inner_join(policy_cutdown_tbl, by = "policy_id") |>
  mutate(
    lapse_date    = as.Date("2015-01-01") + weeks(lapse_time),
    policy_status = case_when(

      # Policy has expired (term ended)
      (policy_enddate <= lapse_date) & (as.Date("2016-01-01") > policy_enddate) ~ "completed",

      # Policy has lapsed and the lapse occurred before or on the reference date
      (lapse_date < policy_enddate) & (lapse_date <= as.Date("2016-01-01")) ~ "lapsed",

      # Policy is within term and either not lapsed or lapsed after reference date
      as.Date("2016-01-01") <= policy_enddate ~ "inforce",

      # Default case (shouldn't occur with proper data)
      TRUE ~ "unknown"
      ),
    policy_statuschangedate = case_when(
      policy_status == "lapsed"    ~ lapse_date,
      policy_status == "completed" ~ policy_enddate,
      policy_status == "inforce"   ~ policy_startdate,
      TRUE                         ~ as.Date(NA)
      )
    )


filterlapse1_sim_policy_data_tbl |> glimpse()
```


## Construct Simulated Cashflow Data


```{r filterlapse1_construct_simulated_cashflows}
#| echo: true
#| message: false

filterlapse1_sim_cashflow_file <- glue("{predictions_dir}/filterlapse1_sim_cashflow_tbl.parquet")

# Check if cached simulated cashflow data exists
if (file_exists(filterlapse1_sim_cashflow_file)) {
  message(glue("Loading cached simulated cashflows from {filterlapse1_sim_cashflow_file}"))
  filterlapse1_sim_cashflow_tbl <- read_parquet(filterlapse1_sim_cashflow_file)
  
} else {
  message(glue("Computing simulated cashflows, will save to {filterlapse1_sim_cashflow_file}"))
  
  filterlapse1_sim_cashflow_tbl <- filterlapse1_sim_policy_data_tbl |>
    group_nest(draw_id) |>
    mutate(
      newdata = map(
        data, \(x) x |> calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date),
                    
        .progress = "joining_data")
        ) |>
    select(-data) |>
    unnest(newdata)
  
  # Save for future use
  filterlapse1_sim_cashflow_tbl |> write_parquet_compressed(filterlapse1_sim_cashflow_file)
  message(glue("Simulated cashflows saved to {filterlapse1_sim_cashflow_file}"))
}

filterlapse1_sim_cashflow_tbl |> glimpse()
```


```{r filterlapse1_construct_simulated_monthly_cashflows}
#| echo: true

filterlapse1_sim_monthly_cashflow_tbl <- filterlapse1_sim_cashflow_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )

filterlapse1_sim_monthly_cashflow_tbl |> glimpse()
```


```{r filterlapse1_construct_comparison_plot}
#| echo: true

ggplot(filterlapse1_sim_monthly_cashflow_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title = glue("Simulated vs Actual Cashflows for Model {filterlapse1_stanmodel_name}")
    )
```

We create a similar plot using ribbons using quantiles of the simulated
cashflows.

```{r filterlapse1_construct_comparison_plot_ribbons}
#| echo: true

filterlapse1_sim_monthly_bands_tbl <- filterlapse1_sim_monthly_cashflow_tbl |>
  summarise(
    p10 = quantile(total_cashflow, 0.10),
    p25 = quantile(total_cashflow, 0.25),
    p50 = quantile(total_cashflow, 0.50),
    p75 = quantile(total_cashflow, 0.75),
    p90 = quantile(total_cashflow, 0.90),
    
    .by = c(cashflow_date)
    )

ggplot(filterlapse1_sim_monthly_bands_tbl) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p10, ymax = p90),
    alpha = 0.5
    ) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p25, ymax = p75),
    alpha = 0.5
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title    = glue("Simulated vs Actual Cashflows for Model {filterlapse1_stanmodel_name}"),
    subtitle = "Showing 50% and 80% bands"
    )
```


### Recent Policies Only (2010+)

We now repeat both visualizations, but restrict the data to only policies that
have a start date from 2010-01-01 onwards, so that we are focusing solely on
more recently incepted policies.

```{r filterlapse1_filter_recent_policies}
#| echo: true

# Filter simulated cashflows for recent policies
filterlapse1_sim_monthly_cashflow_recent_tbl <- filterlapse1_sim_cashflow_tbl |>
  filter(policy_startdate >= as.Date("2010-01-01")) |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )
```

```{r filterlapse1_construct_comparison_plot_recent}
#| echo: true

ggplot(filterlapse1_sim_monthly_cashflow_recent_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_recent_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title = glue("Simulated vs Actual Cashflows for Model {filterlapse1_stanmodel_name}"),
    subtitle = "Recent policies (start date >= 2010-01-01)"
    )
```

```{r filterlapse1_construct_comparison_plot_ribbons_recent}
#| echo: true

filterlapse1_sim_monthly_bands_recent_tbl <- filterlapse1_sim_monthly_cashflow_recent_tbl |>
  summarise(
    p10 = quantile(total_cashflow, 0.10),
    p25 = quantile(total_cashflow, 0.25),
    p50 = quantile(total_cashflow, 0.50),
    p75 = quantile(total_cashflow, 0.75),
    p90 = quantile(total_cashflow, 0.90),
    
    .by = c(cashflow_date)
    )

ggplot(filterlapse1_sim_monthly_bands_recent_tbl) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p10, ymax = p90),
    alpha = 0.5
    ) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p25, ymax = p75),
    alpha = 0.5
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_recent_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title    = glue("Simulated vs Actual Cashflows for Model {filterlapse1_stanmodel_name}"),
    subtitle = "Recent policies (start date >= 2010-01-01), showing 50% and 80% bands"
    )
```


We now clean up the environment before building the next model.

```{r filterlapse1_clear_large_objects}
#| echo: false

# Save tibbles before removing
filterlapse1_post_survival_tbl |> write_parquet(glue("{predictions_dir}/filterlapse1_post_survival_tbl.parquet"))
filterlapse1_sim_cashflow_tbl  |> write_parquet(glue("{predictions_dir}/filterlapse1_sim_cashflow_tbl.parquet"))

# Remove large objects
rm(filterlapse1_post_survival_tbl, filterlapse1_post_draws_lst, filterlapse1_sim_cashflow_tbl)

gc(verbose = TRUE)
```




# Build Expanded Cox Model

We now want to add a number of covariates to the model to see if we can improve
it.

## rstanarm Model


```{r filterlapse2_fit_cox_model_rstanarm}
#| echo: true
#| results: hide
#| message: false

filterlapse2_stanmodel_name <- "filterlapse2_coxph_stansurv"
filterlapse2_stanmodel_file <- glue("{stan_modeldir}/{filterlapse2_stanmodel_name}.qs")

# Check if fitted model exists, load it if so, otherwise fit new model
if (file_exists(filterlapse2_stanmodel_file)) {
  message(glue("Loading cached model from {filterlapse2_stanmodel_file}"))
  filterlapse2_coxph_stansurv <- qread(filterlapse2_stanmodel_file)

} else {
  message(glue("Fitting new model, will save to {filterlapse2_stanmodel_file}"))

  # Create stan_model directory if it doesn't exist
  dir_create(stan_modeldir, recurse = TRUE)
  
  filterlapse2_coxph_stansurv <- stan_surv(
    Surv(policy_lifetime, lapsed) ~ gender_life1 + smoker_life1 + cluster_id + prem_sa_ratio,
    data   = model_training_tbl,
    
    # MCMC sampling parameters
    chains = 4,
    iter   = 2000,
    warmup = 1000,
    seed   = 4000,
    cores  = 4,
    
    # Baseline hazard specification
    basehaz     = "ms",  # M-splines (flexible, default)
    basehaz_ops = list(df = 6),  # Degrees of freedom for baseline hazard
    
    # Prior specifications (explicit is better than implicit)
    prior           = rstanarm::normal(location = 0, scale = 2.5), # Weakly informative for coefficients
    prior_intercept = rstanarm::normal(location = 0, scale = 10),  # For baseline hazard
    
    # Enable verbose output for console monitoring (not captured in document)
    refresh = 200  # Print progress every 200 iterations
    )
  
  # Save fitted model for future use
  filterlapse2_coxph_stansurv |> qsave(filterlapse2_stanmodel_file, preset = "balanced")
  message(glue("Model saved to {filterlapse2_stanmodel_file}"))
}
```

```{r filterlapse2_fit_cox_model_stansurv_summary}
#| echo: true
#| message: true

summary(filterlapse2_coxph_stansurv)
```


## Model Diagnostics

We examine the MCMC diagnostics for the expanded model that includes cluster
effects and premium-to-sum-assured ratio.

```{r filterlapse2_model_diagnostics_convergence}
#| echo: true
#| fig-width: 14
#| fig-height: 5

plot_stan_convergence(
  filterlapse2_coxph_stansurv,
  title = "Model 2 Convergence Diagnostics: Gender + Smoker + Cluster + Premium Ratio"
  )
```


```{r filterlapse2_model_diagnostics_mixing}
#| echo: true
#| fig-width: 14
#| fig-height: 10

plot_stan_mixing(
  filterlapse2_coxph_stansurv,
  title = "Model 2 Chain Mixing: Gender + Smoker + Cluster + Premium Ratio"
  )
```


## Construct Posterior Survival Estimates

We now want to use our fitted model to predict the lapse for the portfolio.

We model the monthly paid premia in the portfolio for the final year for which
we have data. We can then compare our modelled cashflows against the observed
data.

```{r filterlapse2_construct_rstanarm_posterior_estimates}
#| echo: true

filterlapse2_post_draws_file <- glue("{stan_modeldir}/filterlapse2_post_draws_lst.qs")

# Check if cached posterior draws exist
if (file_exists(filterlapse2_post_draws_file)) {
  message(glue("Loading cached posterior draws from {filterlapse2_post_draws_file}"))
  filterlapse2_post_draws_lst <- qread(filterlapse2_post_draws_file)
  
} else {
  message(glue("Computing posterior draws, will save to {filterlapse2_post_draws_file}"))
  
  filterlapse2_post_draws_lst <- posterior_survfit(
      filterlapse2_coxph_stansurv,
      newdata       = inforce_tbl,
      last_time     = NULL,
      draws         = 2000,
      condition     = TRUE,
      return_matrix = TRUE
      )
  
  # Save for future use
  filterlapse2_post_draws_lst |> qsave(filterlapse2_post_draws_file, preset = "balanced")
  message(glue("Posterior draws saved to {filterlapse2_post_draws_file}"))
}

filterlapse2_post_draws_lst[[1]] |> glimpse()
```


```{r filterlapse2_extract_posterior_survival_estimates}
#| echo: true
#| message: false

filterlapse2_cashflow_week_count <- monthly_tbl |> pull(weeks)

filterlapse2_post_survival_file <- glue("{predictions_dir}/filterlapse2_post_survival_tbl.parquet")

# Check if cached posterior survival data exists
if (file_exists(filterlapse2_post_survival_file)) {
  message(glue("Loading cached posterior survival data from {filterlapse2_post_survival_file}"))
  filterlapse2_post_survival_tbl <- read_parquet(filterlapse2_post_survival_file)
  
} else {
  message(glue("Extracting posterior survival data, will save to {filterlapse2_post_survival_file}"))
  
  filterlapse2_post_survival_tbl <- filterlapse2_post_draws_lst |>
    map_dfr(
      extract_stansurv_posterior_survivals,

      .progress = "extract_post_surv"
      ) |>
    filter(time < 5 * 52) |>
    mutate(
      policy_id = inforce_tbl$policy_id[idx],
      
      .before = 1
      ) |>
    select(policy_id, draw_id, time, post_surv_prob = value) |>
    interpolate_survival_data(
      new_times = filterlapse2_cashflow_week_count
      ) |>
    group_nest(policy_id, draw_id)
  
  # Save for future use
  filterlapse2_post_survival_tbl |> write_parquet_compressed(filterlapse2_post_survival_file)
  message(glue("Posterior survival data saved to {filterlapse2_post_survival_file}"))
}

filterlapse2_post_survival_tbl |> glimpse()
```



```{r filterlapse2_calculate_sim_lapse_times}
#| echo: true
#| message: false

filterlapse2_sim_lapse_times_file <- glue("{predictions_dir}/filterlapse2_sim_lapse_times_tbl.parquet")

# Check if cached simulated lapse times exist
if (file_exists(filterlapse2_sim_lapse_times_file)) {
  message(glue("Loading cached simulated lapse times from {filterlapse2_sim_lapse_times_file}"))
  filterlapse2_sim_lapse_times_tbl <- read_parquet(filterlapse2_sim_lapse_times_file)
  
} else {
  message(glue("Computing simulated lapse times, will save to {filterlapse2_sim_lapse_times_file}"))
  
  filterlapse2_sim_lapse_times_tbl <- filterlapse2_post_survival_tbl |>
    mutate(
      rnd_draw   = runif(n(), 0, 1),
      lapse_time = map2_dbl(
        data, rnd_draw,
        \(x, y) extract_survival_time(x, y, min_time = 1),

        .progress = "extract_survival_time"
        )
      ) |>
    select(policy_id, draw_id, lapse_time)
  
  # Save for future use
  filterlapse2_sim_lapse_times_tbl |> write_parquet_compressed(filterlapse2_sim_lapse_times_file)
  message(glue("Simulated lapse times saved to {filterlapse2_sim_lapse_times_file}"))
}

filterlapse2_sim_lapse_times_tbl |> glimpse()
```



```{r filterlapse2_construct_policy_sim_data}
#| echo: true

filterlapse2_sim_policy_data_tbl <- filterlapse2_sim_lapse_times_tbl |>
  inner_join(policy_cutdown_tbl, by = "policy_id") |>
  mutate(
    lapse_date    = as.Date("2015-01-01") + weeks(lapse_time),
    policy_status = case_when(

      # Policy has expired (term ended)
      (policy_enddate <= lapse_date) & (as.Date("2016-01-01") > policy_enddate) ~ "completed",

      # Policy has lapsed and the lapse occurred before or on the reference date
      (lapse_date < policy_enddate) & (lapse_date <= as.Date("2016-01-01")) ~ "lapsed",

      # Policy is within term and either not lapsed or lapsed after reference date
      as.Date("2016-01-01") <= policy_enddate ~ "inforce",

      # Default case (shouldn't occur with proper data)
      TRUE ~ "unknown"
      ),
    policy_statuschangedate = case_when(
      policy_status == "lapsed"    ~ lapse_date,
      policy_status == "completed" ~ policy_enddate,
      policy_status == "inforce"   ~ policy_startdate,
      TRUE                         ~ as.Date(NA)
      )
    )


filterlapse2_sim_policy_data_tbl |> glimpse()
```



We now construct the simulated cashflows from the simulated data.


```{r filterlapse2_construct_simulated_cashflows}
#| echo: true
#| message: false

filterlapse2_sim_cashflow_file <- glue("{predictions_dir}/filterlapse2_sim_cashflow_tbl.parquet")

# Check if cached simulated cashflow data exists
if (file_exists(filterlapse2_sim_cashflow_file)) {
  message(glue("Loading cached simulated cashflows from {filterlapse2_sim_cashflow_file}"))
  filterlapse2_sim_cashflow_tbl <- read_parquet(filterlapse2_sim_cashflow_file)
  
} else {
  message(glue("Computing simulated cashflows, will save to {filterlapse2_sim_cashflow_file}"))
  
  filterlapse2_sim_cashflow_tbl <- filterlapse2_sim_policy_data_tbl |>
    group_nest(draw_id) |>
    mutate(
      newdata = map(
        data, \(x) x |> calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date),
                    
        .progress = "joining_data")
        ) |>
    select(-data) |>
    unnest(newdata)
  
  # Save for future use
  filterlapse2_sim_cashflow_tbl |> write_parquet_compressed(filterlapse2_sim_cashflow_file)
  message(glue("Simulated cashflows saved to {filterlapse2_sim_cashflow_file}"))
}

filterlapse2_sim_cashflow_tbl |> glimpse()
```


```{r filterlapse2_construct_simulated_monthly_cashflows}
#| echo: true

filterlapse2_sim_monthly_cashflow_tbl <- filterlapse2_sim_cashflow_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )

filterlapse2_sim_monthly_cashflow_tbl |> glimpse()
```


```{r filterlapse2_construct_comparison_plot}
#| echo: true

ggplot(filterlapse2_sim_monthly_cashflow_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title = glue("Simulated vs Actual Cashflows for Model {filterlapse2_stanmodel_name}")
    )
```

We create a similar plot using ribbons using quantiles of the simulated
cashflows.

```{r filterlapse2_construct_comparison_plot_ribbons}
#| echo: true

filterlapse2_sim_monthly_bands_tbl <- filterlapse2_sim_monthly_cashflow_tbl |>
  summarise(
    p10 = quantile(total_cashflow, 0.10),
    p25 = quantile(total_cashflow, 0.25),
    p50 = quantile(total_cashflow, 0.50),
    p75 = quantile(total_cashflow, 0.75),
    p90 = quantile(total_cashflow, 0.90),
    
    .by = c(cashflow_date)
    )

ggplot(filterlapse2_sim_monthly_bands_tbl) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p10, ymax = p90),
    alpha = 0.5
    ) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p25, ymax = p75),
    alpha = 0.5
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title    = glue("Simulated vs Actual Cashflows for Model {filterlapse2_stanmodel_name}"),
    subtitle = "Showing 50% and 80% bands"
    )
```


### Recent Policies Only (2010+)

We now repeat both visualizations, but restrict the data to only policies that
have a start date from 2010-01-01 onwards, so that we are focusing solely on
more recently incepted policies.

```{r filterlapse2_filter_recent_policies}
#| echo: true

# Filter simulated cashflows for recent policies
filterlapse2_sim_monthly_cashflow_recent_tbl <- filterlapse2_sim_cashflow_tbl |>
  filter(policy_startdate >= as.Date("2010-01-01")) |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )
```

```{r filterlapse2_construct_comparison_plot_recent}
#| echo: true

ggplot(filterlapse2_sim_monthly_cashflow_recent_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_recent_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title = glue("Simulated vs Actual Cashflows for Model {filterlapse2_stanmodel_name}"),
    subtitle = "Recent policies (start date >= 2010-01-01)"
    )
```

```{r filterlapse2_construct_comparison_plot_ribbons_recent}
#| echo: true

filterlapse2_sim_monthly_bands_recent_tbl <- filterlapse2_sim_monthly_cashflow_recent_tbl |>
  summarise(
    p10 = quantile(total_cashflow, 0.10),
    p25 = quantile(total_cashflow, 0.25),
    p50 = quantile(total_cashflow, 0.50),
    p75 = quantile(total_cashflow, 0.75),
    p90 = quantile(total_cashflow, 0.90),
    
    .by = c(cashflow_date)
    )

ggplot(filterlapse2_sim_monthly_bands_recent_tbl) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p10, ymax = p90),
    alpha = 0.5
    ) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p25, ymax = p75),
    alpha = 0.5
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_recent_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title    = glue("Simulated vs Actual Cashflows for Model {filterlapse2_stanmodel_name}"),
    subtitle = "Recent policies (start date >= 2010-01-01), showing 50% and 80% bands"
    )
```


We now clean up the environment before building the next model.

```{r filterlapse2_clear_large_objects}
#| echo: false

# Save tibbles before removing
filterlapse2_post_survival_tbl |>
  write_parquet(glue("{predictions_dir}/filterlapse2_post_survival_tbl.parquet"))

filterlapse2_sim_cashflow_tbl  |>
  write_parquet(glue("{predictions_dir}/filterlapse2_sim_cashflow_tbl.parquet"))

# Remove large objects
rm(filterlapse2_post_survival_tbl, filterlapse2_post_draws_lst, filterlapse2_sim_cashflow_tbl)

gc(verbose = TRUE)
```



# Add Time-Dependent Parameters

We now want to add a number of covariates to the model to see if we can improve
it.

## rstanarm Model


```{r filterlapse3_fit_cox_model_rstanarm}
#| echo: true
#| message: false

filterlapse3_stanmodel_name <- "filterlapse3_coxph_stansurv"
filterlapse3_stanmodel_file <- glue("{stan_modeldir}/{filterlapse3_stanmodel_name}.qs")

# Check if fitted model exists, load it if so, otherwise fit new model
if (file_exists(filterlapse3_stanmodel_file)) {
  message(glue("Loading cached model from {filterlapse3_stanmodel_file}"))
  filterlapse3_coxph_stansurv <- qread(filterlapse3_stanmodel_file)

} else {
  message(glue("Fitting new model, will save to {filterlapse3_stanmodel_file}"))

  # Create stan_model directory if it doesn't exist
  dir_create(stan_modeldir, recurse = TRUE)
  
  filterlapse3_coxph_stansurv <- stan_surv(
    Surv(policy_lifetime, lapsed) ~ gender_life1 + tve(smoker_life1, df = 5, degree = 2),
    data   = model_training_tbl,
    
    # MCMC sampling parameters
    chains = 4,
    iter   = 2000,
    warmup = 1000,
    seed   = 4000,
    cores  = 4,
    
    # Baseline hazard specification
    basehaz     = "ms",  # M-splines (flexible, default)
    basehaz_ops = list(df = 6),  # Degrees of freedom for baseline hazard
    
    # Prior specifications (explicit is better than implicit)
    prior           = rstanarm::normal(location = 0, scale = 2.5), # Weakly informative for coefficients
    prior_intercept = rstanarm::normal(location = 0, scale = 10),  # For baseline hazard
    
    # Enable verbose output for console monitoring (not captured in document)
    refresh = 200  # Print progress every 200 iterations
    )
  
  # Save fitted model for future use
  filterlapse3_coxph_stansurv |> qsave(filterlapse3_stanmodel_file, preset = "balanced")
  message(glue("Model saved to {filterlapse3_stanmodel_file}"))
}
```

```{r filterlapse3_fit_cox_model_rstanarm_summary}
#| echo: true
#| message: true

summary(filterlapse3_coxph_stansurv)
```


## Model Diagnostics

We examine the MCMC diagnostics for the time-varying effects model. This model
includes a spline-based time-varying effect for the smoker status, allowing
the hazard ratio to change over policy lifetime.

```{r filterlapse3_model_diagnostics_convergence}
#| echo: true
#| fig-width: 14
#| fig-height: 5

plot_stan_convergence(
  filterlapse3_coxph_stansurv,
  title = "Model 3 Convergence Diagnostics: Gender + Time-Varying Smoker Effect"
  )
```


```{r filterlapse3_model_diagnostics_mixing}
#| echo: true
#| fig-width: 14
#| fig-height: 10

plot_stan_mixing(
  filterlapse3_coxph_stansurv,
  title = "Model 3 Chain Mixing: Gender + Time-Varying Smoker Effect"
  )
```


## Construct Posterior Survival Estimates

We now want to use our fitted model to predict the lapse for the portfolio.

We model the monthly paid premia in the portfolio for the final year for which
we have data. We can then compare our modelled cashflows against the observed
data.

```{r filterlapse3_construct_rstanarm_posterior_estimates}
#| echo: true

filterlapse3_post_draws_file <- glue("{stan_modeldir}/filterlapse3_post_draws_lst.qs")

# Check if cached posterior draws exist
if (file_exists(filterlapse3_post_draws_file)) {
  message(glue("Loading cached posterior draws from {filterlapse3_post_draws_file}"))
  filterlapse3_post_draws_lst <- qread(filterlapse3_post_draws_file)
  
} else {
  message(glue("Computing posterior draws, will save to {filterlapse3_post_draws_file}"))
  
  filterlapse3_post_draws_lst <- posterior_survfit(
      filterlapse3_coxph_stansurv,
      newdata       = inforce_tbl,
      last_time     = NULL,
      draws         = 2000,
      condition     = TRUE,
      return_matrix = TRUE
      )
  
  # Save for future use
  filterlapse3_post_draws_lst |> qsave(filterlapse3_post_draws_file, preset = "balanced")
  message(glue("Posterior draws saved to {filterlapse3_post_draws_file}"))
}

filterlapse3_post_draws_lst[[1]] |> glimpse()
```


```{r filterlapse3_extract_posterior_survival_estimates}
#| echo: true
#| message: false

filterlapse3_cashflow_week_count <- monthly_tbl |> pull(weeks)

filterlapse3_post_survival_file <- glue("{predictions_dir}/filterlapse3_post_survival_tbl.parquet")

# Check if cached posterior survival data exists
if (file_exists(filterlapse3_post_survival_file)) {
  message(glue("Loading cached posterior survival data from {filterlapse3_post_survival_file}"))
  filterlapse3_post_survival_tbl <- read_parquet(filterlapse3_post_survival_file)
  
} else {
  message(glue("Extracting posterior survival data, will save to {filterlapse3_post_survival_file}"))
  
  filterlapse3_post_survival_tbl <- filterlapse3_post_draws_lst |>
    map_dfr(
      extract_stansurv_posterior_survivals,

      .progress = "extract_post_surv"
      ) |>
    filter(time < 5 * 52) |>
    mutate(
      policy_id = inforce_tbl$policy_id[idx],
      
      .before = 1
      ) |>
    select(policy_id, draw_id, time, post_surv_prob = value) |>
    interpolate_survival_data(
      new_times = filterlapse3_cashflow_week_count
      ) |>
    group_nest(policy_id, draw_id)
  
  # Save for future use
  filterlapse3_post_survival_tbl |> write_parquet_compressed(filterlapse3_post_survival_file)
  message(glue("Posterior survival data saved to {filterlapse3_post_survival_file}"))
}

filterlapse3_post_survival_tbl |> glimpse()
```



```{r filterlapse3_calculate_sim_lapse_times}
#| echo: true
#| message: false

filterlapse3_sim_lapse_times_file <- glue("{predictions_dir}/filterlapse3_sim_lapse_times_tbl.parquet")

# Check if cached simulated lapse times exist
if (file_exists(filterlapse3_sim_lapse_times_file)) {
  message(glue("Loading cached simulated lapse times from {filterlapse3_sim_lapse_times_file}"))
  filterlapse3_sim_lapse_times_tbl <- read_parquet(filterlapse3_sim_lapse_times_file)
  
} else {
  message(glue("Computing simulated lapse times, will save to {filterlapse3_sim_lapse_times_file}"))
  
  filterlapse3_sim_lapse_times_tbl <- filterlapse3_post_survival_tbl |>
    mutate(
      rnd_draw   = runif(n(), 0, 1),
      lapse_time = map2_dbl(
        data, rnd_draw,
        \(x, y) extract_survival_time(x, y, min_time = 1),

        .progress = "extract_survival_time"
        )
      ) |>
    select(policy_id, draw_id, lapse_time)
  
  # Save for future use
  filterlapse3_sim_lapse_times_tbl |> write_parquet_compressed(filterlapse3_sim_lapse_times_file)
  message(glue("Simulated lapse times saved to {filterlapse3_sim_lapse_times_file}"))
}

filterlapse3_sim_lapse_times_tbl |> glimpse()
```



```{r filterlapse3_construct_policy_sim_data}
#| echo: true

filterlapse3_sim_policy_data_tbl <- filterlapse3_sim_lapse_times_tbl |>
  inner_join(policy_cutdown_tbl, by = "policy_id") |>
  mutate(
    lapse_date    = as.Date("2015-01-01") + weeks(lapse_time),
    policy_status = case_when(

      # Policy has expired (term ended)
      (policy_enddate <= lapse_date) & (as.Date("2016-01-01") > policy_enddate) ~ "completed",

      # Policy has lapsed and the lapse occurred before or on the reference date
      (lapse_date < policy_enddate) & (lapse_date <= as.Date("2016-01-01")) ~ "lapsed",

      # Policy is within term and either not lapsed or lapsed after reference date
      as.Date("2016-01-01") <= policy_enddate ~ "inforce",

      # Default case (shouldn't occur with proper data)
      TRUE ~ "unknown"
      ),
    policy_statuschangedate = case_when(
      policy_status == "lapsed"    ~ lapse_date,
      policy_status == "completed" ~ policy_enddate,
      policy_status == "inforce"   ~ policy_startdate,
      TRUE                         ~ as.Date(NA)
      )
    )


filterlapse3_sim_policy_data_tbl |> glimpse()
```



We now construct the simulated cashflows from the simulated data.


```{r filterlapse3_construct_simulated_cashflows}
#| echo: true
#| message: false

filterlapse3_sim_cashflow_file <- glue("{predictions_dir}/filterlapse3_sim_cashflow_tbl.parquet")

# Check if cached simulated cashflow data exists
if (file_exists(filterlapse3_sim_cashflow_file)) {
  message(glue("Loading cached simulated cashflows from {filterlapse3_sim_cashflow_file}"))
  filterlapse3_sim_cashflow_tbl <- read_parquet(filterlapse3_sim_cashflow_file)
  
} else {
  message(glue("Computing simulated cashflows, will save to {filterlapse3_sim_cashflow_file}"))
  
  filterlapse3_sim_cashflow_tbl <- filterlapse3_sim_policy_data_tbl |>
    group_nest(draw_id) |>
    mutate(
      newdata = map(
        data, \(x) x |> calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date),

        .progress = "joining_data")
        ) |>
    select(-data) |>
    unnest(newdata)
  
  # Save for future use
  filterlapse3_sim_cashflow_tbl |> write_parquet_compressed(filterlapse3_sim_cashflow_file)
  message(glue("Simulated cashflows saved to {filterlapse3_sim_cashflow_file}"))
}

filterlapse3_sim_cashflow_tbl |> glimpse()
```


```{r filterlapse3_construct_simulated_monthly_cashflows}
#| echo: true

filterlapse3_sim_monthly_cashflow_tbl <- filterlapse3_sim_cashflow_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )

filterlapse3_sim_monthly_cashflow_tbl |> glimpse()
```


```{r filterlapse3_construct_comparison_plot}
#| echo: true

ggplot(filterlapse3_sim_monthly_cashflow_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title = glue("Simulated vs Actual Cashflows for Model {filterlapse3_stanmodel_name}")
    )
```

We create a similar plot using ribbons using quantiles of the simulated
cashflows.

```{r filterlapse3_construct_comparison_plot_ribbons}
#| echo: true

filterlapse3_sim_monthly_bands_tbl <- filterlapse3_sim_monthly_cashflow_tbl |>
  summarise(
    p10 = quantile(total_cashflow, 0.10),
    p25 = quantile(total_cashflow, 0.25),
    p50 = quantile(total_cashflow, 0.50),
    p75 = quantile(total_cashflow, 0.75),
    p90 = quantile(total_cashflow, 0.90),
    
    .by = c(cashflow_date)
    )

ggplot(filterlapse3_sim_monthly_bands_tbl) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p10, ymax = p90),
    alpha = 0.5
    ) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p25, ymax = p75),
    alpha = 0.5
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title    = glue("Simulated vs Actual Cashflows for Model {filterlapse3_stanmodel_name}"),
    subtitle = "Showing 50% and 80% bands"
    )
```


### Recent Policies Only (2010+)

We now repeat both visualizations, but restrict the data to only policies that
have a start date from 2010-01-01 onwards, so that we are focusing solely on
more recently incepted policies.

```{r filterlapse3_filter_recent_policies}
#| echo: true

# Filter simulated cashflows for recent policies
filterlapse3_sim_monthly_cashflow_recent_tbl <- filterlapse3_sim_cashflow_tbl |>
  filter(policy_startdate >= as.Date("2010-01-01")) |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )
```

```{r filterlapse3_construct_comparison_plot_recent}
#| echo: true

ggplot(filterlapse3_sim_monthly_cashflow_recent_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_recent_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title = glue("Simulated vs Actual Cashflows for Model {filterlapse3_stanmodel_name}"),
    subtitle = "Recent policies (start date >= 2010-01-01)"
    )
```

```{r filterlapse3_construct_comparison_plot_ribbons_recent}
#| echo: true

filterlapse3_sim_monthly_bands_recent_tbl <- filterlapse3_sim_monthly_cashflow_recent_tbl |>
  summarise(
    p10 = quantile(total_cashflow, 0.10),
    p25 = quantile(total_cashflow, 0.25),
    p50 = quantile(total_cashflow, 0.50),
    p75 = quantile(total_cashflow, 0.75),
    p90 = quantile(total_cashflow, 0.90),
    
    .by = c(cashflow_date)
    )

ggplot(filterlapse3_sim_monthly_bands_recent_tbl) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p10, ymax = p90),
    alpha = 0.5
    ) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p25, ymax = p75),
    alpha = 0.5
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_recent_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title    = glue("Simulated vs Actual Cashflows for Model {filterlapse3_stanmodel_name}"),
    subtitle = "Recent policies (start date >= 2010-01-01), showing 50% and 80% bands"
    )
```


# Chunk Timing Summary

This section summarizes the execution time for all chunks in this notebook.

```{r chunk_timing_summary}
#| echo: false

# Create timing summary table
timing_summary_tbl <- tibble(
  chunk_label = names(chunk_times),
  seconds = unlist(chunk_times)
  ) |>
  arrange(desc(seconds)) |>
  mutate(
    minutes = round(seconds / 60, 2),
    hours = round(seconds / 3600, 2),
    formatted = case_when(
      seconds < 60 ~ glue("{round(seconds, 1)} sec"),
      seconds < 3600 ~ glue("{minutes} min"),
      TRUE ~ glue("{hours} hr")
      )
    )

# Print summary
message("")
message(strrep("=", 80))
message("CHUNK EXECUTION TIMES (slowest first)")
message(strrep("=", 80))
message("")

timing_summary_tbl |>
  select(chunk_label, formatted) |>
  print(n = Inf)

message("")
message(strrep("-", 80))
total_time <- sum(timing_summary_tbl$seconds)
message(glue("Total execution time: {round(total_time/60, 2)} minutes ({round(total_time, 1)} seconds)"))
message(strrep("=", 80))

# Save to file for future reference
dir_create("chunk_timings", recurse = TRUE)
timing_output_file <- glue("chunk_timings/chunk_timings_{format(Sys.time(), '%Y%m%d_%H%M%S')}.parquet")
timing_summary_tbl |> write_parquet_compressed(timing_output_file)
message(glue("Timing data saved to: {timing_output_file}"))
```


