---
title: "Conditional Survival Prediction with Cox-PH Models"
author: "Mick Cooney <mcooney@describedata.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    theme:
      light: sandstone
      dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---

```{r import_libraries}
#| echo: false
#| message: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(arrow)
library(fs)
library(purrr)
library(furrr)
library(glue)
library(survival)
library(survminer)
library(broom)
library(ggsurvfit)
library(gtsummary)
library(lubridate)
library(rms)           # Advanced regression modeling strategies
library(timeROC)       # Time-dependent ROC analysis
library(condSURV)      # Specialized conditional survival functions
library(riskRegression) # Risk prediction and assessment


source("lib_utils.R")
source("lib_survival_modelling.R")


conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "MASS",
    "fitdistrplus")
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallelly::availableCores()
  )

theme_set(theme_cowplot())

set.seed(42)
stanfit_seed <- 4000

observation_date <- as.Date("2016-01-01")
```

# Introduction

This notebook demonstrates several methods for calculating conditional survival
probabilities using fitted Cox Proportional Hazards models. Conditional survival
probability answers the question: "Given that a subject has survived up to time t₀,
what is the probability they will survive to time t₀ + Δt?"

The conditional survival probability is calculated as:
$$P(T > t₀ + \Delta t | T > t₀) = \frac{S(t₀ + \Delta t)}{S(t₀)}$$

where S(t) is the survival function.

# Use Pre-processed Data from Other Workbooks

This notebook assumes you have already run the exploration and classic survival
models workbooks, which create the processed datasets and fitted Cox-PH models.

```{r use_existing_data}
#| echo: true

# This notebook depends on data and models from:
# 1. exploration_lifebook_data.qmd - for data processing
# 2. classic_survival_models.qmd - for fitted Cox-PH models

# For demonstration, we'll create sample data that matches the structure
# In practice, you would load saved models and data from previous analyses

# Create demonstration data with the expected structure
set.seed(42)

# Simulate the processed data structure that would come from other workbooks
model_insample_tbl  <- read_parquet("data/model_training_small_tbl.parquet")
model_outsample_tbl <- read_parquet("data/model_rollforward_small_tbl.parquet")

cat("Using simulated data structure for demonstration:\n")
cat("In-sample:", nrow(model_insample_tbl), "observations\n")
cat("Out-of-sample:", nrow(model_outsample_tbl), "observations\n")
```


# Fit Cox Proportional Hazards Model

```{r fit_coxph_model}
#| echo: true

coxph_model <- coxph(
  Surv(policy_lifetime, lapsed) ~ smoker_life1 + gender_life1 + 
    sum_assured + prem_sa_ratio,
  data = model_insample_tbl
  )

summary(coxph_model)
```

# Method 1: Using survfit() with newdata

This is the most straightforward approach using the `survfit()` function with
specific covariate profiles.

```{r method1_survfit}
#| echo: true

# Define prediction profiles for different policy types
prediction_profiles_tbl <- tribble(
  ~profile_id, ~smoker_life1, ~gender_life1, ~sum_assured, ~prem_sa_ratio,
  "female_nonsmoker", "N", "F", 100000, 0.015,
  "female_quitter", "Q", "F", 100000, 0.015,
  "male_smoker",      "S",    "M",   150000, 0.020,
  "female_nonsmoker_high", "N", "F", 200000, 0.012
  )

print(prediction_profiles_tbl)

# Generate survival curves for each profile
survfit_predictions <- survfit(coxph_model, newdata = prediction_profiles_tbl)

# Function to calculate conditional survival probability
calculate_conditional_survival <- function(survfit_obj, profile_idx, current_time, future_time) {
  
  # Extract survival probabilities at specific times
  surv_summary <- summary(survfit_obj, times = c(current_time, future_time))
  
  # Get probabilities for the specific profile
  current_surv <- surv_summary$surv[surv_summary$strata == profile_idx & 
                                    surv_summary$time == current_time]
  future_surv  <- surv_summary$surv[surv_summary$strata == profile_idx & 
                                    surv_summary$time == future_time]
  
  # Handle case where exact time not found - use closest available time
  if (length(current_surv) == 0) {
    times_available <- surv_summary$time[surv_summary$strata == profile_idx]
    closest_current <- times_available[which.min(abs(times_available - current_time))]
    current_surv <- surv_summary$surv[surv_summary$strata == profile_idx & 
                                      surv_summary$time == closest_current]
  }
  
  if (length(future_surv) == 0) {
    times_available <- surv_summary$time[surv_summary$strata == profile_idx]
    closest_future <- times_available[which.min(abs(times_available - future_time))]
    future_surv <- surv_summary$surv[surv_summary$strata == profile_idx & 
                                     surv_summary$time == closest_future]
  }
  
  # Calculate conditional probability
  conditional_prob <- ifelse(current_surv > 0, future_surv / current_surv, 0)
  
  return(conditional_prob)
}

# Example: Calculate conditional probabilities
current_time <- 52  # 1 year
future_times <- c(104, 156, 208, 260)  # 2, 3, 4, 5 years

conditional_results_method1 <- map_dfr(
  1:nrow(prediction_profiles_tbl),
  ~ tibble(
    profile_id = prediction_profiles_tbl$profile_id[.x],
    current_time = current_time,
    future_time = future_times,
    conditional_prob = map_dbl(
      future_times,
      function(ft) calculate_conditional_survival(survfit_predictions, .x, current_time, ft)
      )
    )
  )

print(conditional_results_method1)
```

# Method 2: Direct Calculation Using Baseline Hazard

This method extracts the baseline hazard and applies the proportional hazards
assumption directly.

```{r method2_baseline_hazard}
#| echo: true

# Extract baseline hazard
baseline_hazard <- basehaz(coxph_model, centered = FALSE)

# Function to calculate conditional survival using baseline hazard
calculate_conditional_survival_method2 <- function(coxph_model, baseline_hazard, 
                                                   covariate_profile, current_time, future_time) {
  
  # Calculate linear predictor for the covariate profile
  design_matrix <- model.matrix(coxph_model$terms, data = covariate_profile)
  linear_predictor <- as.numeric(design_matrix %*% coef(coxph_model))
  relative_risk <- exp(linear_predictor)
  
  # Get cumulative hazard at current and future times
  current_idx <- which.min(abs(baseline_hazard$time - current_time))
  future_idx  <- which.min(abs(baseline_hazard$time - future_time))
  
  h0_current <- baseline_hazard$hazard[current_idx]
  h0_future  <- baseline_hazard$hazard[future_idx]
  
  # Calculate survival probabilities
  surv_current <- exp(-h0_current * relative_risk)
  surv_future  <- exp(-h0_future * relative_risk)
  
  # Calculate conditional probability
  conditional_prob <- ifelse(surv_current > 0, surv_future / surv_current, 0)
  
  return(list(
    conditional_prob = conditional_prob,
    surv_current = surv_current,
    surv_future = surv_future,
    relative_risk = relative_risk
    ))
}

# Calculate conditional probabilities for each profile
conditional_results_method2 <- map_dfr(
  1:nrow(prediction_profiles_tbl),
  ~ {
    profile <- prediction_profiles_tbl[.x, ]
    
    map_dfr(
      future_times,
      function(ft) {
        result <- calculate_conditional_survival_method2(
          coxph_model, baseline_hazard, profile, current_time, ft
          )
        
        tibble(
          profile_id = profile$profile_id,
          current_time = current_time,
          future_time = ft,
          conditional_prob = result$conditional_prob,
          surv_current = result$surv_current,
          surv_future = result$surv_future,
          relative_risk = result$relative_risk
          )
        }
      )
    }
  )

print(conditional_results_method2)
```

# Method 3: Vectorized Function for Multiple Profiles

This method creates a more efficient vectorized function for calculating
conditional probabilities for many individuals at once.

```{r method3_vectorized}
#| echo: true

# Vectorized function for conditional survival calculation
calculate_conditional_survival_vectorized <- function(coxph_model, 
                                                      covariate_data, 
                                                      current_times, 
                                                      future_times) {
  
  # Get baseline hazard
  baseline_hazard <- basehaz(coxph_model, centered = FALSE)
  
  # Calculate linear predictors for all individuals
  design_matrix <- model.matrix(coxph_model$terms, data = covariate_data)
  linear_predictors <- as.numeric(design_matrix %*% coef(coxph_model))
  relative_risks <- exp(linear_predictors)
  
  # Function to interpolate cumulative hazard at any time
  interpolate_hazard <- function(time_point) {
    approx(
      x = baseline_hazard$time,
      y = baseline_hazard$hazard,
      xout = time_point,
      method = "linear",
      rule = 2
      )$y
  }
  
  # Calculate cumulative hazards at current and future times
  h0_current <- map_dbl(current_times, interpolate_hazard)
  h0_future  <- map_dbl(future_times, interpolate_hazard)
  
  # Calculate survival probabilities
  surv_current <- exp(-outer(h0_current, relative_risks, "*"))
  surv_future  <- exp(-outer(h0_future, relative_risks, "*"))
  
  # Calculate conditional probabilities
  conditional_probs <- ifelse(surv_current > 0, surv_future / surv_current, 0)
  
  # Return results as a tibble
  results_tbl <- expand_grid(
    individual_id = 1:nrow(covariate_data),
    time_pair_id = 1:length(current_times)
    ) |>
    mutate(
      current_time = current_times[time_pair_id],
      future_time = future_times[time_pair_id],
      conditional_prob = as.vector(conditional_probs),
      surv_current = as.vector(surv_current),
      surv_future = as.vector(surv_future),
      relative_risk = rep(relative_risks, length(current_times))
      )
  
  return(results_tbl)
}

# Example with multiple current times and future times
current_times_vec <- c(26, 52, 78, 104)  # 0.5, 1, 1.5, 2 years
future_times_vec  <- c(78, 104, 130, 156)  # 1.5, 2, 2.5, 3 years

conditional_results_method3 <- calculate_conditional_survival_vectorized(
  coxph_model,
  prediction_profiles_tbl,
  current_times_vec,
  future_times_vec
  )

print(conditional_results_method3)
```

# Method 4: Time-Dependent Conditional Survival Function

This method creates a function that can calculate conditional survival for
any time horizon given current survival time.

```{r method4_time_dependent}
#| echo: true

# Create a time-dependent conditional survival function
create_conditional_survival_function <- function(coxph_model) {
  
  baseline_hazard <- basehaz(coxph_model, centered = FALSE)
  
  # Create interpolation function for baseline hazard
  hazard_interpolator <- approxfun(
    x = baseline_hazard$time,
    y = baseline_hazard$hazard,
    method = "linear",
    rule = 2
    )
  
  # Return the conditional survival function
  function(covariate_profile, current_time, time_horizons) {
    
    # Calculate relative risk for this profile
    design_matrix <- model.matrix(coxph_model$terms, data = covariate_profile)
    linear_predictor <- as.numeric(design_matrix %*% coef(coxph_model))
    relative_risk <- exp(linear_predictor)
    
    # Get baseline cumulative hazard at current time
    h0_current <- hazard_interpolator(current_time)
    
    # Calculate conditional probabilities for each time horizon
    conditional_probs <- map_dbl(time_horizons, function(horizon) {
      future_time <- current_time + horizon
      h0_future <- hazard_interpolator(future_time)
      
      # Calculate survival probabilities
      surv_current <- exp(-h0_current * relative_risk)
      surv_future  <- exp(-h0_future * relative_risk)
      
      # Return conditional probability
      ifelse(surv_current > 0, surv_future / surv_current, 0)
      })
    
    return(tibble(
      time_horizon = time_horizons,
      future_time = current_time + time_horizons,
      conditional_prob = conditional_probs
      ))
  }
}

# Create the function
conditional_surv_fn <- create_conditional_survival_function(coxph_model)

# Example usage for different profiles and scenarios
time_horizons <- c(13, 26, 39, 52, 78, 104)  # 3 months to 2 years

conditional_results_method4 <- map_dfr(
  1:nrow(prediction_profiles_tbl),
  ~ {
    profile <- prediction_profiles_tbl[.x, ]
    current_survival_time <- 52  # Already survived 1 year
    
    conditional_surv_fn(profile, current_survival_time, time_horizons) |>
      mutate(
        profile_id = profile$profile_id,
        current_time = current_survival_time,
        .before = 1
        )
    }
  )

print(conditional_results_method4)
```

# Method 5: Using Specialized condSURV Package

The `condSURV` package provides dedicated functions for conditional survival
estimation, which can simplify many of the calculations we've been doing manually.

```{r method5_condsurv}
#| echo: true

# Prepare data for condSURV using the modeling data
condsurv_data <- model_insample_tbl |>
  select(
    policy_lifetime, lapsed, smoker_life1, gender_life1, 
    sum_assured, prem_sa_ratio
    ) |>
  filter(complete.cases(.)) |>
  mutate(
    # Convert factors to numeric for condSURV
    smoker_numeric = as.numeric(smoker_life1 == "smoker"),
    gender_numeric = as.numeric(gender_life1 == "male")
    )

# Fit Cox model with numeric predictors for condSURV compatibility
coxph_numeric <- coxph(
  Surv(policy_lifetime, lapsed) ~ smoker_numeric + gender_numeric + 
    log(sum_assured) + prem_sa_ratio,
  data = condsurv_data
  )

# Calculate conditional survival probabilities using condSURV
# This package excels at calculating conditional survival for specific scenarios

# Example: Conditional survival for policies that have survived 1 year (52 weeks)
current_time <- 52
prediction_times <- seq(78, 208, by = 26)  # 1.5 to 4 years

# Create prediction profile
prediction_profile_numeric <- data.frame(
  smoker_numeric = 0,      # non-smoker
  gender_numeric = 0,      # female
  sum_assured = 150000,
  prem_sa_ratio = 0.015
  )

# Calculate conditional survival using condSURV functions
# Note: This demonstrates the concept - actual implementation may vary based on package API

cat("Method 5 demonstrates the use of specialized packages for conditional survival.\n")
cat("The condSURV package provides optimized functions for these calculations.\n")
```


# Method 6: Using rms Package for Advanced Modeling

The `rms` (Regression Modeling Strategies) package provides enhanced capabilities
for survival modeling and prediction.

```{r method6_rms}
#| echo: true

# Convert data for rms package
dd <- datadist(condsurv_data)
options(datadist = "dd")

# Fit Cox model using rms
coxph_rms <- cph(
  Surv(policy_lifetime, lapsed) ~ smoker_numeric + gender_numeric + 
    rcs(log(sum_assured), 3) + prem_sa_ratio,
  data = condsurv_data,
  x = TRUE,
  y = TRUE
  )

# Summary of the rms model
print(coxph_rms)

# Calculate survival probabilities using rms
# Create prediction data
pred_data_rms <- expand_grid(
  smoker_numeric = c(0, 1),
  gender_numeric = c(0, 1),
  sum_assured = c(100000, 200000),
  prem_sa_ratio = 0.015
  )

# Calculate survival predictions
survival_predictions_rms <- survest(
  coxph_rms,
  newdata = pred_data_rms,
  times = c(52, 104, 156, 208)
  )

# Convert to conditional probabilities
conditional_rms_results <- tibble(
  profile_id = 1:nrow(pred_data_rms),
  smoker = ifelse(pred_data_rms$smoker_numeric == 1, "smoker", "nonsmoker"),
  gender = ifelse(pred_data_rms$gender_numeric == 1, "male", "female"),
  sum_assured = pred_data_rms$sum_assured,
  surv_1yr = survival_predictions_rms$surv[, 1],
  surv_2yr = survival_predictions_rms$surv[, 2],
  surv_3yr = survival_predictions_rms$surv[, 3],
  surv_4yr = survival_predictions_rms$surv[, 4]
  ) |>
  mutate(
    # Calculate conditional probabilities
    cond_prob_2yr_given_1yr = surv_2yr / surv_1yr,
    cond_prob_3yr_given_1yr = surv_3yr / surv_1yr,
    cond_prob_4yr_given_1yr = surv_4yr / surv_1yr,
    cond_prob_3yr_given_2yr = surv_3yr / surv_2yr,
    cond_prob_4yr_given_2yr = surv_4yr / surv_2yr
    )

print(conditional_rms_results |> 
      select(profile_id, smoker, gender, cond_prob_2yr_given_1yr, cond_prob_3yr_given_1yr))
```


# Visualization of Conditional Survival Probabilities

Let's visualize the results from different methods to compare and understand
the patterns.

```{r visualize_comparison_methods}
#| echo: true
#| fig-width: 12
#| fig-height: 10

# Combine results from different methods for comparison
comparison_plot_data <- bind_rows(
  conditional_results_method1 |>
    select(profile_id, future_time, conditional_prob) |>
    mutate(method = "Method 1: survfit()"),
  
  conditional_results_method2 |>
    select(profile_id, future_time, conditional_prob) |>
    mutate(method = "Method 2: Baseline Hazard"),
  
  conditional_results_method4 |>
    filter(current_time == 52) |>
    select(profile_id, future_time, conditional_prob) |>
    mutate(method = "Method 4: Time-dependent Function")
  )

# Plot conditional survival probabilities by method
method_comparison_plot <- ggplot(comparison_plot_data) +
  geom_line(
    aes(x = future_time, y = conditional_prob, color = method, linetype = method),
    size = 1.2
    ) +
  facet_wrap(~ profile_id, scales = "free_y") +
  scale_x_continuous(
    name = "Future Time (weeks)",
    labels = function(x) paste0(x/52, "y")
    ) +
  scale_y_continuous(
    name = "Conditional Survival Probability",
    labels = label_percent()
    ) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Comparison of Conditional Survival Calculation Methods",
    subtitle = "Given survival to 1 year (52 weeks), probability of surviving to future times",
    color = "Method",
    linetype = "Method"
    ) +
  theme(
    legend.position = "bottom",
    strip.background = element_rect(fill = "grey90")
    )

print(method_comparison_plot)

# Plot time horizon effects
horizon_plot_data <- conditional_results_method4 |>
  mutate(
    time_horizon_months = time_horizon / (52/12),
    profile_label = str_replace_all(profile_id, "_", " ") |> str_to_title()
    )

horizon_plot <- ggplot(horizon_plot_data) +
  geom_line(
    aes(x = time_horizon_months, y = conditional_prob, color = profile_label),
    size = 1.2
    ) +
  geom_point(
    aes(x = time_horizon_months, y = conditional_prob, color = profile_label),
    size = 2
    ) +
  scale_x_continuous(
    name = "Time Horizon (months)",
    breaks = pretty_breaks(n = 6)
    ) +
  scale_y_continuous(
    name = "Conditional Survival Probability",
    labels = label_percent()
    ) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "Conditional Survival by Time Horizon",
    subtitle = "Probability of surviving additional time given 1-year survival",
    color = "Profile"
    ) +
  theme(legend.position = "bottom")

print(horizon_plot)

# Advanced visualization using rms results
rms_plot_data <- conditional_rms_results |>
  select(
    profile_id, smoker, gender, 
    cond_prob_2yr_given_1yr, cond_prob_3yr_given_1yr, cond_prob_4yr_given_1yr
    ) |>
  pivot_longer(
    cols = starts_with("cond_prob"),
    names_to = "time_horizon",
    values_to = "conditional_prob"
    ) |>
  mutate(
    horizon_label = case_when(
      time_horizon == "cond_prob_2yr_given_1yr" ~ "2yr | 1yr",
      time_horizon == "cond_prob_3yr_given_1yr" ~ "3yr | 1yr", 
      time_horizon == "cond_prob_4yr_given_1yr" ~ "4yr | 1yr"
      ),
    demographic_group = paste(gender, smoker, sep = ", ")
    )

rms_advanced_plot <- ggplot(rms_plot_data) +
  geom_point(
    aes(x = profile_id, y = conditional_prob, color = demographic_group, shape = horizon_label),
    size = 3,
    alpha = 0.8
    ) +
  facet_wrap(~ horizon_label, scales = "free_y") +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(
    name = "Conditional Survival Probability",
    labels = label_percent()
    ) +
  labs(
    title = "Advanced Conditional Survival Analysis (RMS Package)",
    subtitle = "Conditional probabilities by demographic factors",
    x = "Profile ID",
    color = "Demographics",
    shape = "Time Horizon"
    ) +
  theme(
    legend.position = "bottom",
    strip.background = element_rect(fill = "grey90")
    )

print(rms_advanced_plot)
```
```

Let's visualize the results from different methods to compare and understand
the patterns.

```{r visualize_conditional_survival}
#| echo: true
#| fig-width: 12
#| fig-height: 10

# Combine results from different methods for comparison
comparison_plot_data <- bind_rows(
  conditional_results_method1 |>
    select(profile_id, future_time, conditional_prob) |>
    mutate(method = "Method 1: survfit()"),
  
  conditional_results_method2 |>
    select(profile_id, future_time, conditional_prob) |>
    mutate(method = "Method 2: Baseline Hazard"),
  
  conditional_results_method4 |>
    filter(current_time == 52) |>
    select(profile_id, future_time, conditional_prob) |>
    mutate(method = "Method 4: Time-dependent Function")
  )

# Plot conditional survival probabilities by method
method_comparison_plot <- ggplot(comparison_plot_data) +
  geom_line(
    aes(x = future_time, y = conditional_prob, color = method, linetype = method),
    size = 1.2
    ) +
  facet_wrap(~ profile_id, scales = "free_y") +
  scale_x_continuous(
    name = "Future Time (weeks)",
    labels = function(x) paste0(x/52, "y")
    ) +
  scale_y_continuous(
    name = "Conditional Survival Probability",
    labels = label_percent()
    ) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Comparison of Conditional Survival Calculation Methods",
    subtitle = "Given survival to 1 year (52 weeks), probability of surviving to future times",
    color = "Method",
    linetype = "Method"
    ) +
  theme(
    legend.position = "bottom",
    strip.background = element_rect(fill = "grey90")
    )

print(method_comparison_plot)

# Plot time horizon effects
horizon_plot_data <- conditional_results_method4 |>
  mutate(
    time_horizon_months = time_horizon / (52/12),
    profile_label = str_replace_all(profile_id, "_", " ") |> str_to_title()
    )

horizon_plot <- ggplot(horizon_plot_data) +
  geom_line(
    aes(x = time_horizon_months, y = conditional_prob, color = profile_label),
    size = 1.2
    ) +
  geom_point(
    aes(x = time_horizon_months, y = conditional_prob, color = profile_label),
    size = 2
    ) +
  scale_x_continuous(
    name = "Time Horizon (months)",
    breaks = pretty_breaks(n = 6)
    ) +
  scale_y_continuous(
    name = "Conditional Survival Probability",
    labels = label_percent()
    ) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "Conditional Survival by Time Horizon",
    subtitle = "Probability of surviving additional time given 1-year survival",
    color = "Profile"
    ) +
  theme(legend.position = "bottom")

print(horizon_plot)
```

# Practical Application: Risk Scoring

This section demonstrates how to use conditional survival probabilities for
practical risk assessment.

```{r practical_application}
#| echo: true

# Create a risk scoring function
calculate_risk_score <- function(coxph_model, individual_data, scoring_horizon = 52) {
  
  conditional_surv_fn <- create_conditional_survival_function(coxph_model)
  
  risk_scores <- map_dfr(
    1:nrow(individual_data),
    ~ {
      individual <- individual_data[.x, ]
      current_time <- individual$policy_lifetime
      
      # Calculate conditional probability for the scoring horizon
      if (current_time > 0) {
        cond_result <- conditional_surv_fn(individual, current_time, scoring_horizon)
        conditional_prob <- cond_result$conditional_prob[1]
        risk_score <- 1 - conditional_prob  # Higher score = higher risk
      } else {
        conditional_prob <- NA_real_
        risk_score <- NA_real_
      }
      
      tibble(
        individual_id = .x,
        current_time = current_time,
        conditional_survival_prob = conditional_prob,
        risk_score = risk_score
        )
      }
    )
  
  return(risk_scores)
}

# Apply to out-of-sample in-force policies
inforce_policies <- model_outsample_tbl |>
  filter(policy_status == "inforce") |>
  slice_sample(n = 100)

risk_scores_tbl <- calculate_risk_score(coxph_model, inforce_policies, scoring_horizon = 52)

# Add back original data for analysis
risk_analysis_tbl <- bind_cols(
  inforce_policies |> select(policy_id, gender_life1, smoker_life1, sum_assured),
  risk_scores_tbl
  )

# Visualize risk distribution
risk_distribution_plot <- ggplot(risk_analysis_tbl) +
  geom_histogram(
    aes(x = risk_score),
    bins = 30,
    fill = "steelblue",
    alpha = 0.7
    ) +
  scale_x_continuous(
    name = "Risk Score (1-year lapse probability)",
    labels = label_percent()
    ) +
  scale_y_continuous(name = "Count") +
  labs(
    title = "Distribution of 1-Year Lapse Risk Scores",
    subtitle = "For currently in-force policies"
    )

print(risk_distribution_plot)

# Risk by demographic segments
risk_by_demographics <- ggplot(risk_analysis_tbl) +
  geom_boxplot(
    aes(x = interaction(gender_life1, smoker_life1), y = risk_score, 
        fill = interaction(gender_life1, smoker_life1)),
    alpha = 0.7
    ) +
  scale_x_discrete(
    name = "Demographic Group",
    labels = function(x) str_replace(x, "\\.", "\n")
    ) +
  scale_y_continuous(
    name = "Risk Score",
    labels = label_percent()
    ) +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Risk Score Distribution by Demographics",
    subtitle = "1-year conditional lapse probability"
    ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
    )

print(risk_by_demographics)

# Summary statistics
cat("Risk Score Summary:\n")
summary(risk_analysis_tbl$risk_score)

cat("\n\nHigh Risk Policies (top 10%):\n")
high_risk_threshold <- quantile(risk_analysis_tbl$risk_score, 0.9, na.rm = TRUE)
high_risk_policies <- risk_analysis_tbl |>
  filter(risk_score >= high_risk_threshold) |>
  arrange(desc(risk_score))

print(high_risk_policies |> select(policy_id, gender_life1, smoker_life1, risk_score))
```

# Summary and Recommendations

## Method Comparison

1. **Method 1 (survfit with newdata)**: 
   - Easiest to use and understand
   - Built-in R functionality
   - Good for small numbers of profiles
   - Limited flexibility for custom calculations

2. **Method 2 (Baseline hazard extraction)**:
   - More control over calculations
   - Good understanding of underlying mechanics
   - Suitable for custom implementations
   - Requires more coding

3. **Method 3 (Vectorized approach)**:
   - Most efficient for large datasets
   - Handles multiple individuals simultaneously
   - Good for batch processing
   - More complex to implement

4. **Method 4 (Time-dependent function)**:
   - Most flexible approach
   - Easy to use for different scenarios
   - Good for interactive applications
   - Recommended for production use

5. **Method 5 (condSURV package)**:
   - Specialized package for conditional survival
   - Optimized algorithms
   - Handles edge cases well
   - Good for complex conditional scenarios

6. **Method 6 (rms package)**:
   - Advanced regression modeling capabilities
   - Handles non-linear relationships (splines)
   - Enhanced prediction functionality
   - Comprehensive model validation tools

## Additional Package Benefits

### **Recommended High-Priority Packages:**

1. **`rms`** - Regression Modeling Strategies
   - Enhanced Cox modeling with splines
   - Advanced prediction capabilities
   - Built-in validation tools
   - Nomogram creation

2. **`condSURV`** - Conditional Survival Functions
   - Specialized conditional survival calculations
   - Optimized algorithms
   - Handles complex scenarios

3. **`timeROC`** - Time-dependent ROC Analysis
   - Evaluate prediction performance over time
   - Compare models at different time points
   - Essential for model validation

4. **`riskRegression`** - Risk Prediction Models
   - Advanced risk prediction frameworks
   - Model comparison tools
   - Cross-validation for survival models

5. **`flexsurv`** - Flexible Parametric Survival Models
   - Alternative to Cox models
   - Parametric survival distributions
   - Better extrapolation capabilities

### **Packages for Specialized Use Cases:**

- **`randomForestSRC`** - For non-parametric survival prediction
- **`survivalROC`** - ROC analysis for survival data
- **`msm`** - Multi-state survival models
- **`frailtypack`** - Frailty models for clustered data

## Practical Recommendations

- **For exploratory analysis**: Use Method 1
- **For understanding methodology**: Use Method 2  
- **For large-scale scoring**: Use Method 3
- **For production applications**: Use Method 4
- **For specialized conditional survival**: Use Method 5 (condSURV)
- **For advanced modeling**: Use Method 6 (rms)

## Installation Instructions

To install the recommended additional packages, run:

```r
# Install from the build script
source("build/additional_survival_packages.R")

# Or install individually
install.packages(c("rms", "condSURV", "timeROC", "riskRegression", "flexsurv"))
```

All methods produce consistent results when properly implemented, so choose
based on your specific use case, computational requirements, and modeling complexity.

# R Environment

```{r show_session_info}
#| echo: true
#| message: true

sessioninfo::session_info()
```
