---
title: "Conditional Survival Prediction with Cox-PH Models"
author: "Mick Cooney <mcooney@describedata.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    theme:
      light: sandstone
      dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---

```{r import_libraries}
#| echo: false
#| message: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(arrow)
library(fs)
library(purrr)
library(furrr)
library(glue)
library(survival)
library(survminer)
library(broom)
library(ggsurvfit)
library(gtsummary)
library(lubridate)
library(rms)           # Advanced regression modeling strategies
library(timeROC)       # Time-dependent ROC analysis
library(condSURV)      # Specialized conditional survival functions
library(riskRegression) # Risk prediction and assessment


source("lib_utils.R")
source("lib_survival_modelling.R")


conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "MASS",
    "fitdistrplus")
  )

options(
  width    = 80L,
  warn     = 1,
  mc.cores = parallelly::availableCores()
  )

theme_set(theme_cowplot())

set.seed(42)
stanfit_seed <- 4000

observation_date <- as.Date("2016-01-01")
```

# Introduction

This notebook demonstrates several methods for calculating conditional survival
probabilities using fitted Cox Proportional Hazards models. 

Conditional survival probability is a powerful concept that extends standard survival analysis by updating survival predictions based on the passage of time. While standard survival probability asks, "What is the probability of surviving from time zero to time $t$?", conditional survival asks a more dynamic question: "Given that an individual has already survived to a specific time $t_0$, what is the probability they will survive for an *additional* period $\Delta t$?"

### The Intuition and Mathematics

The intuition behind conditional survival is rooted in the fact that accumulating "survival time" changes the risk profile. For example, a patient who survives five years post-diagnosis often has a much better prognosis for the next five years than a newly diagnosed patient. The risk of the event happening is not uniformly distributed over time. 

Mathematically, this is expressed using the standard rules of conditional probability:
$$P(T > t_0 + \Delta t \mid T > t_0) = \frac{P(T > t_0 + \Delta t \cap T > t_0)}{P(T > t_0)}$$

Since $T > t_0 + \Delta t$ logically implies $T > t_0$, the intersection is simply $T > t_0 + \Delta t$. Therefore, we can rewrite this using the overall survival function $S(t) = P(T > t)$:
$$P(T > t_0 + \Delta t \mid T > t_0) = \frac{S(t_0 + \Delta t)}{S(t_0)}$$

This ratio intuitively scales the future survival probability by the probability of having reached the current time. 

### Business Applications and Value

In practical business applications, conditional survival is often much more relevant than baseline survival from time zero:
*   **Life Insurance and Actuarial Science:** For in-force policy management, actuaries need to know the risk of a policyholder lapsing or dying *today*, given they have already held the policy for several years. This informs dynamic reserve calculations and premium pricing.
*   **Risk Scoring and Customer Retention:** In subscription models or SaaS businesses, the likelihood of customer churn changes based on tenure. Conditional survival helps create dynamic "risk scores" for existing customers, allowing retention teams to target interventions precisely when the risk spikes.
*   **Medical Prognosis:** Clinicians use conditional survival to provide updated, more optimistic prognoses to patients who have survived the initial high-risk period following a treatment or surgery.

# Use Pre-processed Data from Other Workbooks

This notebook assumes you have already run the exploration and classic survival
models workbooks, which create the processed datasets and fitted Cox-PH models.

```{r use_existing_data}
#| echo: true

# This notebook depends on data and models from:
# 1. exploration_lifebook_data.qmd - for data processing
# 2. classic_survival_models.qmd - for fitted Cox-PH models

# For demonstration, we'll create sample data that matches the structure
# In practice, you would load saved models and data from previous analyses

# Create demonstration data with the expected structure
set.seed(42)

# Simulate the processed data structure that would come from other workbooks
model_insample_tbl  <- read_parquet("data/model_training_small_tbl.parquet")
model_outsample_tbl <- read_parquet("data/model_rollforward_small_tbl.parquet")

cat("Using simulated data structure for demonstration:\n")
cat("In-sample:", nrow(model_insample_tbl), "observations\n")
cat("Out-of-sample:", nrow(model_outsample_tbl), "observations\n")
```


# Fit Cox Proportional Hazards Model

```{r fit_coxph_model}
#| echo: true

coxph_model <- coxph(
  Surv(policy_lifetime, lapsed) ~ smoker_life1 + gender_life1 + 
    sum_assured + prem_sa_ratio,
  data = model_insample_tbl
  )

summary(coxph_model)
```

# Method 1: Using survfit() with newdata

The `survfit()` function from the `survival` package is the most straightforward, built-in method for generating survival curves from a fitted Cox model. By supplying a `newdata` argument, we can predict survival probabilities tailored to specific covariate profiles (e.g., a female non-smoker with a specific sum assured).

### Under the Hood of `survfit`

When `survfit()` receives `newdata`, it performs two primary steps:
1.  **Extracts the Baseline Survival:** It calculates the baseline survival curve $S_0(t)$ corresponding to the mean of the covariates in the original training data (this centering ensures numerical stability).
2.  **Applies the Hazard Ratio:** It calculates the relative risk (hazard ratio) for each new profile in `newdata` based on the fitted model coefficients. It then scales the baseline survival curve to match the specific profile: $S(t|X) = S_0(t)^{\exp(\beta X)}$.

### Pros and Cons

*   **Pros:** This method is robust, handles complex model formulations natively, and correctly calculates standard errors and confidence intervals for the predicted curves. It is the safest, most "standard" approach.
*   **Cons:** It can be computationally slow and cumbersome when predicting for a very large number of profiles simultaneously or when we need to rapidly calculate conditional probabilities for many arbitrary $(t_0, t_0 + \Delta t)$ pairs, as it calculates the entire curve up to the maximum time for each profile before we can extract the specific points we need.

```{r method1_survfit}
#| echo: true

# Define prediction profiles for different policy types
prediction_profiles_tbl <- tribble(
  ~profile_id, ~smoker_life1, ~gender_life1, ~sum_assured, ~prem_sa_ratio,
  "female_nonsmoker", "N", "F", 100000, 0.015,
  "female_quitter", "Q", "F", 100000, 0.015,
  "male_smoker",      "S",    "M",   150000, 0.020,
  "female_nonsmoker_high", "N", "F", 200000, 0.012
  )

print(prediction_profiles_tbl)

# Generate survival curves for each profile
survfit_predictions <- survfit(coxph_model, newdata = prediction_profiles_tbl)

# Function to calculate conditional survival probability
calculate_conditional_survival <- function(survfit_obj, profile_idx, current_time, future_time) {
  
  # Extract survival probabilities at specific times
  surv_summary <- summary(survfit_obj, times = c(current_time, future_time), extend = TRUE)
  
  t1_idx <- (surv_summary$time == current_time) |> which()
  t2_idx <- (surv_summary$time == future_time)  |> which()
  
  if (length(t1_idx) == 0 || length(t2_idx) == 0) {
    return(NA_real_)
  }
  
  if (is.matrix(surv_summary$surv)) {
    current_surv <- surv_summary$surv[t1_idx, profile_idx]
    future_surv  <- surv_summary$surv[t2_idx, profile_idx]
  } else {
    current_surv <- surv_summary$surv[t1_idx]
    future_surv  <- surv_summary$surv[t2_idx]
  }
  
  # Calculate conditional probability
  conditional_prob <- ifelse(current_surv > 0, future_surv / current_surv, 0)
  
  return(conditional_prob)
}

# Example: Calculate conditional probabilities
current_time <- 52  # 1 year
future_times <- c(104, 156, 208, 260)  # 2, 3, 4, 5 years

conditional_results_method1_calc_future <- function(ft, survfit_preds, prof_idx, curr_time) {
  calculate_conditional_survival(survfit_preds, prof_idx, curr_time[1], ft)
}

conditional_results_method1_calc_profile <- function(prof_idx, profiles_tbl, curr_time, fut_times, survfit_preds) {
  cond_probs <- fut_times |>
    map_dbl(
      conditional_results_method1_calc_future,
      survfit_preds = survfit_preds,
      prof_idx      = prof_idx,
      curr_time     = curr_time
      )

  tibble(
    profile_id       = profiles_tbl$profile_id[prof_idx],
    current_time     = curr_time,
    future_time      = fut_times,
    conditional_prob = cond_probs
    )
}

conditional_results_method1 <- 1:nrow(prediction_profiles_tbl) |>
  map_dfr(
    conditional_results_method1_calc_profile,
    profiles_tbl  = prediction_profiles_tbl,
    curr_time     = current_time,
    fut_times     = future_times,
    survfit_preds = survfit_predictions
    )

print(conditional_results_method1)
```

# Method 2: Direct Calculation Using Baseline Hazard

To bypass the overhead of `survfit()`, we can directly calculate conditional survival by explicitly extracting the baseline hazard and leveraging the core assumption of the Cox Proportional Hazards (PH) model.

### The Baseline Hazard and Proportionality

A Cox PH model separates the hazard function $h(t|X)$ into two multiplicative components:
$$h(t|X) = h_0(t) \times \exp(\beta X)$$

Here, $h_0(t)$ is the **baseline hazard**—the hazard for an individual whose covariates $X$ are all zero (or at their centered means). The term $\exp(\beta X)$ is the **linear predictor** or relative risk, which is constant over time (this is the "proportional hazards" assumption).

Because the cumulative hazard $H(t)$ is the integral of the hazard function, the proportionality extends to the cumulative hazard:
$$H(t|X) = H_0(t) \times \exp(\beta X)$$

And since survival probability is related to cumulative hazard by $S(t) = \exp(-H(t))$, we can compute survival for any individual directly if we know $H_0(t)$ and their relative risk.

### Why Direct Calculation Works

By using `basehaz()`, we extract the step-function of the baseline cumulative hazard. For any specific time $t$, we can find the corresponding $H_0(t)$, multiply it by the individual's relative risk, and exponentiate to get their specific survival probability. Because we only compute the survival at the exact times we need ($t_0$ and $t_0 + \Delta t$) rather than constructing the whole object, this method is more direct and provides a deeper understanding of the model's mechanics.

```{r method2_baseline_hazard}
#| echo: true

# Extract baseline hazard
baseline_hazard <- basehaz(coxph_model, centered = FALSE)

# Function to calculate conditional survival using baseline hazard
calculate_conditional_survival_method2 <- function(coxph_model, baseline_hazard, 
                                                   covariate_profile, current_time, future_time) {
  
  # Calculate relative risk for the covariate profile
  relative_risk <- predict(coxph_model, newdata = covariate_profile, type = "risk") |>
    as.numeric()

  # Get cumulative hazard at current and future times
  current_idx <- abs(baseline_hazard$time - current_time) |> which.min()
  future_idx  <- abs(baseline_hazard$time - future_time)  |> which.min()

  h0_current <- baseline_hazard$hazard[current_idx]
  h0_future  <- baseline_hazard$hazard[future_idx]
  
  # Calculate survival probabilities
  surv_current <- exp(-h0_current * relative_risk)
  surv_future  <- exp(-h0_future * relative_risk)
  
  # Calculate conditional probability
  conditional_prob <- ifelse(surv_current > 0, surv_future / surv_current, 0)
  
  return(list(
    conditional_prob = conditional_prob,
    surv_current     = surv_current,
    surv_future      = surv_future,
    relative_risk    = relative_risk
    ))
}

conditional_results_method2_calc_future <- function(ft, cox_mod, base_haz, prof, curr_time) {
  result <- calculate_conditional_survival_method2(
    cox_mod, base_haz, prof, curr_time, ft
    )

  tibble(
    profile_id       = prof$profile_id,
    current_time     = curr_time,
    future_time      = ft,
    conditional_prob = result$conditional_prob,
    surv_current     = result$surv_current,
    surv_future      = result$surv_future,
    relative_risk    = result$relative_risk
    )
}

conditional_results_method2_calc_profile <- function(prof_idx, profiles_tbl, fut_times, cox_mod, base_haz, curr_time) {
  profile <- profiles_tbl[prof_idx, ]

  fut_times |>
    map_dfr(
      conditional_results_method2_calc_future,
      cox_mod   = cox_mod,
      base_haz  = base_haz,
      prof      = profile,
      curr_time = curr_time
      )
}

# Calculate conditional probabilities for each profile
conditional_results_method2 <- 1:nrow(prediction_profiles_tbl) |>
  map_dfr(
    conditional_results_method2_calc_profile,
    profiles_tbl = prediction_profiles_tbl,
    fut_times    = future_times,
    cox_mod      = coxph_model,
    base_haz     = baseline_hazard,
    curr_time    = current_time
    )

print(conditional_results_method2)
```

# Method 3: Vectorized Function for Multiple Profiles

When dealing with large portfolios (e.g., predicting the 1-year conditional lapse probability for a million in-force policies), the loop-based approaches of Methods 1 and 2 become computationally prohibitive. This method leverages R's vectorization capabilities to perform these calculations efficiently.

### The Power of Vectorization in R

R is optimized for vectorized operations, meaning it performs computations on entire arrays or vectors at once in compiled C/C++ code, rather than evaluating an R-level loop iteration by iteration.

In this approach, we:
1.  **Extract All Relative Risks Simultaneously:** We use `predict(type = "risk")` on the entire dataset. Under the hood, this uses a design matrix (similar to `model.matrix`) multiplied by the coefficient vector to yield the relative risk $\exp(\beta X)$ for every individual in one fast operation.
2.  **Interpolate Cumulative Hazards:** We use `approx()` to find the baseline cumulative hazard at all desired current times ($t_0$) and future times ($t_0 + \Delta t$).
3.  **The `outer` Product:** The `outer(X, Y, "*")` function calculates the Cartesian product of the time-specific hazards and the individual specific risks. It creates a matrix where every row is a time point and every column is an individual, calculating $H_0(t) \times \exp(\beta X)$ for all combinations simultaneously.

This dramatically reduces computational time, making it the preferred method for bulk risk scoring or portfolio-wide projections.

```{r method3_vectorized}
#| echo: true

# Vectorized function for conditional survival calculation
calculate_conditional_survival_vectorized <- function(coxph_model, 
                                                      covariate_data, 
                                                      current_times, 
                                                      future_times) {
  
  # Get baseline hazard
  baseline_hazard <- basehaz(coxph_model, centered = FALSE)
  
  # Calculate relative risks for all individuals
  relative_risks <- predict(coxph_model, newdata = covariate_data, type = "risk") |>
    as.numeric()

  # Function to interpolate cumulative hazard at any time
  interpolate_hazard <- function(time_point) {
    approx(
      x      = baseline_hazard$time,
      y      = baseline_hazard$hazard,
      xout   = time_point,
      method = "linear",
      rule   = 2
      )$y
  }

  # Calculate cumulative hazards at current and future times
  h0_current <- current_times |> map_dbl(interpolate_hazard)
  h0_future  <- future_times  |> map_dbl(interpolate_hazard)

  # Calculate survival probabilities
  surv_current <- exp(-outer(h0_current, relative_risks, "*"))
  surv_future  <- exp(-outer(h0_future, relative_risks, "*"))
  
  # Calculate conditional probabilities
  conditional_probs <- ifelse(surv_current > 0, surv_future / surv_current, 0)
  
  # Return results as a tibble
  results_tbl <- expand_grid(
    individual_id = 1:nrow(covariate_data),
    time_pair_id  = 1:length(current_times)
    ) |>
    mutate(
      current_time     = current_times[time_pair_id],
      future_time      = future_times[time_pair_id],
      conditional_prob = as.vector(conditional_probs),
      surv_current     = as.vector(surv_current),
      surv_future      = as.vector(surv_future),
      relative_risk    = relative_risks |> rep(length(current_times))
      )
  
  return(results_tbl)
}

# Example with multiple current times and future times
current_times_vec <- c(26, 52, 78, 104)  # 0.5, 1, 1.5, 2 years
future_times_vec  <- c(78, 104, 130, 156)  # 1.5, 2, 2.5, 3 years

conditional_results_method3 <- calculate_conditional_survival_vectorized(
  coxph_model,
  prediction_profiles_tbl,
  current_times_vec,
  future_times_vec
  )

print(conditional_results_method3)
```

# Method 4: Time-Dependent Conditional Survival Function

In interactive applications (like a Shiny dashboard) or complex simulations, we often don't know the exact future time horizons in advance. This method creates a "factory" function that generates a highly optimized conditional survival calculator.

### Closures in R

A **closure** in R is a function that is created by another function and "remembers" the environment in which it was created.

Here, `create_conditional_survival_function()` does the heavy lifting exactly once: it extracts the baseline hazard and builds an interpolation function (`hazard_interpolator`). It then returns a new inner function. This inner function has access to `hazard_interpolator` and the `coxph_model` without needing to recalculate them.

### Flexible, On-the-Fly Predictions

By interpolating the cumulative hazard (using `approxfun`), the returned closure can calculate survival for any arbitrary continuous time point, not just the discrete event times observed in the training data. This allows users to query the function dynamically: "If this policyholder has survived 52.5 weeks, what is the probability they survive the next 13.2 weeks?" The function can calculate this on-the-fly instantly, making it ideal for deployment in production environments.

```{r method4_time_dependent}
#| echo: true

# Create a time-dependent conditional survival function
create_conditional_survival_function <- function(coxph_model) {
  
  baseline_hazard <- basehaz(coxph_model, centered = FALSE)
  
  # Create interpolation function for baseline hazard
  hazard_interpolator <- approxfun(
    x      = baseline_hazard$time,
    y      = baseline_hazard$hazard,
    method = "linear",
    rule   = 2
    )
  
  # Return the conditional survival function
  function(covariate_profile, current_time, time_horizons) {
    
    # Calculate relative risk for this profile
    relative_risk <- predict(coxph_model, newdata = covariate_profile, type = "risk") |>
      as.numeric()

    # Get baseline cumulative hazard at current time
    h0_current <- hazard_interpolator(current_time)
    # Function to calculate horizon probability
    calc_horizon_prob <- function(horizon, curr_time, haz_interp, h0_curr, rel_risk) {
      future_time <- curr_time + horizon
      h0_future <- haz_interp(future_time)

      # Calculate survival probabilities
      surv_current <- exp(-h0_curr * rel_risk)
      surv_future  <- exp(-h0_future * rel_risk)

      # Return conditional probability
      ifelse(surv_current > 0, surv_future / surv_current, 0)
    }

    # Calculate conditional probabilities for each time horizon
    conditional_probs <- time_horizons |>
      map_dbl(
        calc_horizon_prob,
        curr_time  = current_time,
        haz_interp = hazard_interpolator,
        h0_curr    = h0_current,
        rel_risk   = relative_risk
        )
    
    return(tibble(
      time_horizon     = time_horizons,
      future_time      = current_time + time_horizons,
      conditional_prob = conditional_probs
      ))
  }
}

# Create the function
conditional_surv_fn <- create_conditional_survival_function(coxph_model)

# Example usage for different profiles and scenarios
time_horizons <- c(13, 26, 39, 52, 78, 104)  # 3 months to 2 years

conditional_results_method4_calc_profile <- function(prof_idx, profiles_tbl, curr_survival_time, cond_fn, horizons) {
  profile <- profiles_tbl[prof_idx, ]

  cond_fn(profile, curr_survival_time, horizons) |>
    mutate(
      profile_id   = profile$profile_id,
      current_time = curr_survival_time,
      .before      = 1
      )
}

conditional_results_method4 <- 1:nrow(prediction_profiles_tbl) |>
  map_dfr(
    conditional_results_method4_calc_profile,
    profiles_tbl       = prediction_profiles_tbl,
    curr_survival_time = 52, # Already survived 1 year
    cond_fn            = conditional_surv_fn,
    horizons           = time_horizons
    )

print(conditional_results_method4)
```

# Method 5: Using Specialized condSURV Package

While the manual calculations shown above build a strong foundational understanding and offer high performance, they can become cumbersome when dealing with more complex survival scenarios (e.g., competing risks, multi-state models, or confidence interval estimation for conditional probabilities).

The `condSURV` package is part of a broader ecosystem of specialized R packages designed to handle these nuances natively.

### Advantages of Specialized Packages

*   **Robustness to Edge Cases:** Manual implementations might fail or produce unexpected results if asked to predict beyond the maximum follow-up time or handle ties incorrectly. Specialized packages have robust error handling for these scenarios.
*   **Statistical Rigor:** While we calculated point estimates manually, `condSURV` provides built-in methods for calculating standard errors and confidence bands for conditional survival estimates using techniques like bootstrapping or the delta method.
*   **Landmark Analysis:** `condSURV` also supports advanced techniques like landmark analysis, where models are dynamically re-fitted at specific time points ($t_0$), which relaxes the proportional hazards assumption over long horizons.

*(Note: The `condSURV` code here demonstrates the data preparation required for strict numerical compatibility often needed by such specialized packages.)*

```{r method5_condsurv}
#| echo: true

# Prepare data for condSURV using the modeling data
condsurv_data <- model_insample_tbl |>
  select(
    policy_lifetime, lapsed, smoker_life1, gender_life1, 
    sum_assured, prem_sa_ratio
    ) |>
  drop_na() |>
  mutate(
    # Convert factors to numeric for condSURV
    smoker_numeric = (smoker_life1 == "S") |> as.numeric(),
    gender_numeric = (gender_life1 == "M") |> as.numeric()
    )

# Fit Cox model with numeric predictors for condSURV compatibility
coxph_numeric <- coxph(
  Surv(policy_lifetime, lapsed) ~ smoker_numeric + gender_numeric + 
    log(sum_assured) + prem_sa_ratio,
  data = condsurv_data
  )

# Calculate conditional survival probabilities using condSURV
# This package excels at calculating conditional survival for specific scenarios

# Example: Conditional survival for policies that have survived 1 year (52 weeks)
current_time <- 52
prediction_times <- seq(78, 208, by = 26)  # 1.5 to 4 years

# Create prediction profile
prediction_profile_numeric <- data.frame(
  smoker_numeric = 0,      # non-smoker
  gender_numeric = 0,      # female
  sum_assured    = 150000,
  prem_sa_ratio  = 0.015
  )

# Calculate conditional survival using condSURV functions
# Note: This demonstrates the concept - actual implementation may vary based on package API

cat("Method 5 demonstrates the use of specialized packages for conditional survival.\n")
cat("The condSURV package provides optimized functions for these calculations.\n")
```


# Method 6: Using rms Package for Advanced Modeling

The `rms` (Regression Modeling Strategies) package, authored by Frank Harrell, is widely considered the gold standard in R for developing, validating, and predicting from clinical and epidemiological regression models.

### What `rms` Brings to the Table

1.  **Restricted Cubic Splines (`rcs`):** Standard Cox models assume a linear relationship between continuous predictors and the log-hazard. The `rms` package makes it trivial to include non-linear effects using restricted cubic splines (as seen with `rcs(log(sum_assured), 3)`). This prevents model misspecification and often leads to much better predictions.
2.  **Enhanced Prediction via `survest`:** The `survest()` function is explicitly designed to be faster and more predictable than the base `survfit()`. It allows for easy extraction of survival probabilities at exact requested times without messy indexing.
3.  **Comprehensive Validation:** Beyond simple prediction, `rms` provides functions like `validate()` and `calibrate()` to perform rigorous internal validation using bootstrapping, helping assess how well the model discriminates and whether its predicted probabilities match observed frequencies.
4.  **Nomograms:** `rms` is famous for its ability to generate nomograms, which are visual representations of the model that allow non-technical stakeholders to manually calculate probabilities.

By using `rms`, the process of transitioning from overall survival to conditional survival becomes significantly more reliable, especially when the underlying relationships in the data are complex.

```{r method6_rms}
#| echo: true

# Convert data for rms package
condsurv_data_df <- as.data.frame(condsurv_data)
dd <- datadist(condsurv_data_df)
options(datadist = "dd")

# Fit Cox model using rms
coxph_rms <- cph(
  Surv(policy_lifetime, lapsed) ~ smoker_numeric + gender_numeric + 
    rcs(log(sum_assured), 3) + prem_sa_ratio,
  data = condsurv_data_df,
  x    = TRUE,
  y    = TRUE
  )

# Summary of the rms model
print(coxph_rms)

# Calculate survival probabilities using rms
# Create prediction data
pred_data_rms <- expand_grid(
  smoker_numeric = c(0, 1),
  gender_numeric = c(0, 1),
  sum_assured    = c(100000, 200000),
  prem_sa_ratio  = 0.015
  ) |>
  as.data.frame()

# Calculate survival predictions
survival_predictions_rms <- survest(
  coxph_rms,
  newdata = pred_data_rms,
  times   = c(52, 104, 156, 208)
  )

# Convert to conditional probabilities
conditional_rms_results <- tibble(
  profile_id  = 1:nrow(pred_data_rms),
  smoker      = ifelse(pred_data_rms$smoker_numeric == 1, "smoker", "nonsmoker"),
  gender      = ifelse(pred_data_rms$gender_numeric == 1, "male", "female"),
  sum_assured = pred_data_rms$sum_assured,
  surv_1yr    = survival_predictions_rms$surv[, 1],
  surv_2yr    = survival_predictions_rms$surv[, 2],
  surv_3yr    = survival_predictions_rms$surv[, 3],
  surv_4yr    = survival_predictions_rms$surv[, 4]
  ) |>
  mutate(
    # Calculate conditional probabilities
    cond_prob_2yr_given_1yr = surv_2yr / surv_1yr,
    cond_prob_3yr_given_1yr = surv_3yr / surv_1yr,
    cond_prob_4yr_given_1yr = surv_4yr / surv_1yr,
    cond_prob_3yr_given_2yr = surv_3yr / surv_2yr,
    cond_prob_4yr_given_2yr = surv_4yr / surv_2yr
    )

print(conditional_rms_results |> 
      select(profile_id, smoker, gender, cond_prob_2yr_given_1yr, cond_prob_3yr_given_1yr))
```


# Visualization of Conditional Survival Probabilities

Visualizing conditional survival curves is crucial for communicating the changing risk profile over time to stakeholders. Unlike standard survival curves that start at 100% at time zero and strictly decrease, conditional survival curves highlight how the prognosis "resets" at the conditioning time $t_0$.

### Interpreting the Visualizations

1.  **Method Comparison Plot:** This plot ensures that our manual calculations (Method 2, Method 4) align perfectly with the standard built-in implementations (Method 1). If the lines completely overlap, it confirms our mathematical derivations are correct. The curve shows the probability of surviving to a future time, *given* survival up to week 52.
2.  **Time Horizon Effects Plot:** This plot translates the concept into a practical view: "How does the probability of surviving change depending on how far out we look?" It clearly shows that predicting short horizons (e.g., 3 months) yields much higher probabilities than predicting long horizons (e.g., 2 years) for the exact same patient. It also highlights the separation between high-risk (smokers) and low-risk (non-smokers) profiles over those horizons.
3.  **Advanced `rms` Plot:** This multi-faceted plot demonstrates how multiple risk factors (gender and smoking status) combine to affect conditional survival over different time horizons (2-year given 1-year, 3-year given 1-year, etc.). This type of visualization is highly effective for identifying specific demographic cohorts that exhibit steep drops in conditional survival, signaling a need for targeted interventions.

```{r visualize_comparison_methods}
#| echo: true
#| fig-width: 12
#| fig-height: 10

# Combine results from different methods for comparison
comparison_plot_data <- bind_rows(
  conditional_results_method1 |>
    select(profile_id, future_time, conditional_prob) |>
    mutate(method = "Method 1: survfit()"),
  
  conditional_results_method2 |>
    select(profile_id, future_time, conditional_prob) |>
    mutate(method = "Method 2: Baseline Hazard"),
  
  conditional_results_method4 |>
    filter(current_time == 52) |>
    select(profile_id, future_time, conditional_prob) |>
    mutate(method = "Method 4: Time-dependent Function")
  )

# Plot conditional survival probabilities by method
method_comparison_plot <- ggplot(comparison_plot_data) +
  geom_line(
    aes(x = future_time, y = conditional_prob, color = method, linetype = method),
    size = 1.2
    ) +
  facet_wrap(~ profile_id, scales = "free_y") +
  scale_x_continuous(
    name   = "Future Time (weeks)",
    labels = function(x) paste0(x/52, "y")
    ) +
  scale_y_continuous(
    name   = "Conditional Survival Probability",
    labels = label_percent()
    ) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title    = "Comparison of Conditional Survival Calculation Methods",
    subtitle = "Given survival to 1 year (52 weeks), probability of surviving to future times",
    color    = "Method",
    linetype = "Method"
    ) +
  theme(
    legend.position  = "bottom",
    strip.background = element_rect(fill = "grey90")
    )

print(method_comparison_plot)

# Plot time horizon effects
horizon_plot_data <- conditional_results_method4 |>
  mutate(
    time_horizon_months = time_horizon / (52/12),
    profile_label       = str_replace_all(profile_id, "_", " ") |> str_to_title()
    )

horizon_plot <- ggplot(horizon_plot_data) +
  geom_line(
    aes(x = time_horizon_months, y = conditional_prob, color = profile_label),
    size = 1.2
    ) +
  geom_point(
    aes(x = time_horizon_months, y = conditional_prob, color = profile_label),
    size = 2
    ) +
  scale_x_continuous(
    name   = "Time Horizon (months)",
    breaks = pretty_breaks(n = 6)
    ) +
  scale_y_continuous(
    name   = "Conditional Survival Probability",
    labels = label_percent()
    ) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title    = "Conditional Survival by Time Horizon",
    subtitle = "Probability of surviving additional time given 1-year survival",
    color    = "Profile"
    ) +
  theme(legend.position = "bottom")

print(horizon_plot)

# Advanced visualization using rms results
rms_plot_data <- conditional_rms_results |>
  select(
    profile_id, smoker, gender, 
    cond_prob_2yr_given_1yr, cond_prob_3yr_given_1yr, cond_prob_4yr_given_1yr
    ) |>
  pivot_longer(
    cols      = starts_with("cond_prob"),
    names_to  = "time_horizon",
    values_to = "conditional_prob"
    ) |>
  mutate(
    horizon_label     = case_when(
      time_horizon == "cond_prob_2yr_given_1yr" ~ "2yr | 1yr",
      time_horizon == "cond_prob_3yr_given_1yr" ~ "3yr | 1yr", 
      time_horizon == "cond_prob_4yr_given_1yr" ~ "4yr | 1yr"
      ),
    demographic_group = paste(gender, smoker, sep = ", ")
    )

rms_advanced_plot <- ggplot(rms_plot_data) +
  geom_point(
    aes(x = profile_id, y = conditional_prob, color = demographic_group, shape = horizon_label),
    size  = 3,
    alpha = 0.8
    ) +
  facet_wrap(~ horizon_label, scales = "free_y") +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(
    name   = "Conditional Survival Probability",
    labels = label_percent()
    ) +
  labs(
    title    = "Advanced Conditional Survival Analysis (RMS Package)",
    subtitle = "Conditional probabilities by demographic factors",
    x        = "Profile ID",
    color    = "Demographics",
    shape    = "Time Horizon"
    ) +
  theme(
    legend.position  = "bottom",
    strip.background = element_rect(fill = "grey90")
    )

print(rms_advanced_plot)
```


Let's visualize the results from different methods to compare and understand
the patterns.

```{r visualize_conditional_survival}
#| echo: true
#| fig-width: 12
#| fig-height: 10

# Combine results from different methods for comparison
comparison_plot_data <- bind_rows(
  conditional_results_method1 |>
    select(profile_id, future_time, conditional_prob) |>
    mutate(method = "Method 1: survfit()"),
  
  conditional_results_method2 |>
    select(profile_id, future_time, conditional_prob) |>
    mutate(method = "Method 2: Baseline Hazard"),
  
  conditional_results_method4 |>
    filter(current_time == 52) |>
    select(profile_id, future_time, conditional_prob) |>
    mutate(method = "Method 4: Time-dependent Function")
  )

# Plot conditional survival probabilities by method
method_comparison_plot <- ggplot(comparison_plot_data) +
  geom_line(
    aes(x = future_time, y = conditional_prob, color = method, linetype = method),
    size = 1.2
    ) +
  facet_wrap(~ profile_id, scales = "free_y") +
  scale_x_continuous(
    name   = "Future Time (weeks)",
    labels = function(x) paste0(x/52, "y")
    ) +
  scale_y_continuous(
    name   = "Conditional Survival Probability",
    labels = label_percent()
    ) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title    = "Comparison of Conditional Survival Calculation Methods",
    subtitle = "Given survival to 1 year (52 weeks), probability of surviving to future times",
    color    = "Method",
    linetype = "Method"
    ) +
  theme(
    legend.position  = "bottom",
    strip.background = element_rect(fill = "grey90")
    )

print(method_comparison_plot)

# Plot time horizon effects
horizon_plot_data <- conditional_results_method4 |>
  mutate(
    time_horizon_months = time_horizon / (52/12),
    profile_label       = str_replace_all(profile_id, "_", " ") |> str_to_title()
    )

horizon_plot <- ggplot(horizon_plot_data) +
  geom_line(
    aes(x = time_horizon_months, y = conditional_prob, color = profile_label),
    size = 1.2
    ) +
  geom_point(
    aes(x = time_horizon_months, y = conditional_prob, color = profile_label),
    size = 2
    ) +
  scale_x_continuous(
    name   = "Time Horizon (months)",
    breaks = pretty_breaks(n = 6)
    ) +
  scale_y_continuous(
    name   = "Conditional Survival Probability",
    labels = label_percent()
    ) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title    = "Conditional Survival by Time Horizon",
    subtitle = "Probability of surviving additional time given 1-year survival",
    color    = "Profile"
    ) +
  theme(legend.position = "bottom")

print(horizon_plot)
```

# Practical Application: Risk Scoring

Conditional survival isn't just an abstract statistical concept; it is the engine behind dynamic risk scoring systems.

In the context of life insurance or subscription services, a standard Cox model predicts the risk of an event from the start of the policy. However, to manage a portfolio of *in-force* policies, actuaries need to know the risk of lapse *today*. 

By calculating the conditional probability of a policy lapsing within the next year, given that it has survived to its current duration, we generate an actionable **Risk Score**.

*   **Targeted Interventions:** The histogram and boxplots below show the distribution of these scores. A business can set a threshold (e.g., the top 10% highest risk scores) and trigger specific retention campaigns (e.g., a phone call from an agent, a premium discount) targeted exclusively at those policies most likely to lapse in the immediate future.
*   **Dynamic Updating:** Because this score is based on conditional survival, a policy's score will dynamically change every month it remains active, reflecting its maturing risk profile accurately.

```{r practical_application}
#| echo: true

# Helper function for risk calculation
calculate_individual_risk_score <- function(ind_idx, ind_data, cond_fn, score_horizon) {
  individual <- ind_data[ind_idx, ]
  current_time <- individual$policy_lifetime

  # Calculate conditional probability for the scoring horizon
  if (current_time > 0) {
    cond_result <- cond_fn(individual, current_time, score_horizon)
    conditional_prob <- cond_result$conditional_prob[1]
    risk_score <- 1 - conditional_prob  # Higher score = higher risk
  } else {
    conditional_prob <- NA_real_
    risk_score <- NA_real_
  }

  tibble(
    individual_id             = ind_idx,
    current_time              = current_time,
    conditional_survival_prob = conditional_prob,
    risk_score                = risk_score
    )
}

# Create a risk scoring function
calculate_risk_score <- function(coxph_model, individual_data, scoring_horizon = 52) {

  conditional_surv_fn <- create_conditional_survival_function(coxph_model)

  risk_scores <- 1:nrow(individual_data) |>
    map_dfr(
      calculate_individual_risk_score,
      ind_data      = individual_data,
      cond_fn       = conditional_surv_fn,
      score_horizon = scoring_horizon
      )

  return(risk_scores)
}

# Apply to out-of-sample in-force policies
inforce_policies <- model_outsample_tbl |>
  filter(policy_status == "inforce") |>
  slice_sample(n = 100)

risk_scores_tbl <- calculate_risk_score(coxph_model, inforce_policies, scoring_horizon = 52)

# Add back original data for analysis
risk_analysis_tbl <- bind_cols(
  inforce_policies |> select(policy_id, gender_life1, smoker_life1, sum_assured),
  risk_scores_tbl
  )

# Visualize risk distribution
risk_distribution_plot <- ggplot(risk_analysis_tbl) +
  geom_histogram(
    aes(x = risk_score),
    bins  = 30,
    fill  = "steelblue",
    alpha = 0.7
    ) +
  scale_x_continuous(
    name   = "Risk Score (1-year lapse probability)",
    labels = label_percent()
    ) +
  scale_y_continuous(name = "Count") +
  labs(
    title    = "Distribution of 1-Year Lapse Risk Scores",
    subtitle = "For currently in-force policies"
    )

print(risk_distribution_plot)

# Risk by demographic segments
risk_by_demographics <- ggplot(risk_analysis_tbl) +
  geom_boxplot(
    aes(x = interaction(gender_life1, smoker_life1), y = risk_score, 
        fill = interaction(gender_life1, smoker_life1)),
    alpha = 0.7
    ) +
  scale_x_discrete(
    name   = "Demographic Group",
    labels = function(x) str_replace(x, "\\.", "\n")
    ) +
  scale_y_continuous(
    name   = "Risk Score",
    labels = label_percent()
    ) +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title    = "Risk Score Distribution by Demographics",
    subtitle = "1-year conditional lapse probability"
    ) +
  theme(
    legend.position = "none",
    axis.text.x     = element_text(angle = 45, hjust = 1)
    )

print(risk_by_demographics)

# Summary statistics
cat("Risk Score Summary:\n")
risk_analysis_tbl$risk_score |> summary()

cat("\n\nHigh Risk Policies (top 10%):\n")
high_risk_threshold <- risk_analysis_tbl$risk_score |> quantile(0.9, na.rm = TRUE)
high_risk_policies <- risk_analysis_tbl |>
  filter(risk_score >= high_risk_threshold) |>
  arrange(desc(risk_score))

print(high_risk_policies |> select(policy_id, gender_life1, smoker_life1, risk_score))
```

# Summary and Recommendations

## Method Comparison

1. **Method 1 (survfit with newdata)**: 
   - Easiest to use and understand
   - Built-in R functionality
   - Good for small numbers of profiles
   - Limited flexibility for custom calculations

2. **Method 2 (Baseline hazard extraction)**:
   - More control over calculations
   - Good understanding of underlying mechanics
   - Suitable for custom implementations
   - Requires more coding

3. **Method 3 (Vectorized approach)**:
   - Most efficient for large datasets
   - Handles multiple individuals simultaneously
   - Good for batch processing
   - More complex to implement

4. **Method 4 (Time-dependent function)**:
   - Most flexible approach
   - Easy to use for different scenarios
   - Good for interactive applications
   - Recommended for production use

5. **Method 5 (condSURV package)**:
   - Specialized package for conditional survival
   - Optimized algorithms
   - Handles edge cases well
   - Good for complex conditional scenarios

6. **Method 6 (rms package)**:
   - Advanced regression modeling capabilities
   - Handles non-linear relationships (splines)
   - Enhanced prediction functionality
   - Comprehensive model validation tools

## Additional Package Benefits

### **Recommended High-Priority Packages:**

1. **`rms`** - Regression Modeling Strategies
   - Enhanced Cox modeling with splines
   - Advanced prediction capabilities
   - Built-in validation tools
   - Nomogram creation

2. **`condSURV`** - Conditional Survival Functions
   - Specialized conditional survival calculations
   - Optimized algorithms
   - Handles complex scenarios

3. **`timeROC`** - Time-dependent ROC Analysis
   - Evaluate prediction performance over time
   - Compare models at different time points
   - Essential for model validation

4. **`riskRegression`** - Risk Prediction Models
   - Advanced risk prediction frameworks
   - Model comparison tools
   - Cross-validation for survival models

5. **`flexsurv`** - Flexible Parametric Survival Models
   - Alternative to Cox models
   - Parametric survival distributions
   - Better extrapolation capabilities

### **Packages for Specialized Use Cases:**

- **`randomForestSRC`** - For non-parametric survival prediction
- **`survivalROC`** - ROC analysis for survival data
- **`msm`** - Multi-state survival models
- **`frailtypack`** - Frailty models for clustered data

## Practical Recommendations

- **For exploratory analysis**: Use Method 1
- **For understanding methodology**: Use Method 2  
- **For large-scale scoring**: Use Method 3
- **For production applications**: Use Method 4
- **For specialized conditional survival**: Use Method 5 (condSURV)
- **For advanced modeling**: Use Method 6 (rms)

## Installation Instructions

To install the recommended additional packages, run:

```r
# Install from the build script
source("build/additional_survival_packages.R")

# Or install individually
install.packages(c("rms", "condSURV", "timeROC", "riskRegression", "flexsurv"))
```

All methods produce consistent results when properly implemented, so choose
based on your specific use case, computational requirements, and modeling complexity.

# R Environment

```{r show_session_info}
#| echo: true
#| message: true

sessioninfo::session_info()
```
