---
title: "Initial Bayesian Survival Models for Life Insurance Lapses"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  error: false
  message: false
  echo: true
  daemon: false
format:
  html:
    light: superhero
    dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
---


```{r import_libraries}
#| echo: false
#| message: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(arrow)
library(fs)
library(qs)
library(purrr)
library(furrr)
library(glue)
library(scales)
library(cmdstanr)
library(brms)
library(posterior)
library(bayesplot)
library(tidybayes)
library(splines2)
library(rstanarm)


source("lib_utils.R")
source("lib_brms_hazard.R")
source("lib_survival_modelling.R")
source("lib_cashflow_projection.R")
source("lib_stan_diagnostics.R")


conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "MASS",
    "fitdistrplus")
  )


options(
  width = 80L,
  warn  = 1,
  mc.cores = parallelly::availableCores()
  )

# Configure Arrow for multi-threaded parquet I/O
set_cpu_count(parallelly::availableCores())
set_io_thread_count(parallelly::availableCores())

message(glue(
  "Arrow configured: {arrow::cpu_count()} CPU threads, ",
  "{io_thread_count()} I/O threads"
))

message("write_parquet_compressed() loaded from lib_utils.R (zstd level 3 default)")

stan_modeldir <- "stan_model"
stan_codedir  <- "stan_code"
data_dir      <- "data"


set.seed(42)
stanfit_seed <- 4000

theme_set(theme_cowplot())

plan(multisession)
```

The notebook begins by loading all required packages, resolving namespace
conflicts for reproducibility, and configuring multi-threaded Arrow I/O. We
also establish global options (line width, cores) and set a consistent theme.
Parallel and timing infrastructure are enabled up-front so later chunks
automatically benefit from these settings without repeated boilerplate.


```{r setup_chunk_timing}
#| echo: false
#| message: false

# Enable chunk timing for all chunks
knitr::opts_chunk$set(time_it = TRUE)

# Create a list to store chunk times
chunk_times <- list()

# Create a hook to capture chunk execution times
knitr::knit_hooks$set(time_it = function(before, options) {
  if (before) {
    # Record start time in global environment
    assign(".chunk_start_time", Sys.time(), envir = .GlobalEnv)
  } else {
    # Calculate elapsed time
    elapsed <- difftime(
        Sys.time(), get(".chunk_start_time", envir = .GlobalEnv),
        units = "secs"
        ) |>
      as.numeric()
    
    # Store in global list
    chunk_times <- get("chunk_times", envir = .GlobalEnv)
    chunk_times[[options$label]] <- elapsed
    assign("chunk_times", chunk_times, envir = .GlobalEnv)
    
    # Print timing message
    message(glue("⏱️  Chunk '{options$label}': {round(elapsed, 2)} seconds"))
  }
})
```



# Load Data

We first load up the data we are using to train our dataset, as well as the
'rolled-forward' data that observes the same data a year later.

```{r load_modelling_data}
#| echo: true

model_training_tbl <- read_parquet(glue("{data_dir}/model_training_small_tbl.parquet"))

model_training_tbl |> glimpse()
```

We then load up the 'current' data that we use to validate our model.


```{r load_current_data}
#| echo: true

model_rollforward_tbl <- read_parquet(glue("{data_dir}/model_rollforward_small_tbl.parquet"))

model_rollforward_tbl |> glimpse()
```


## Create Shared Data Objects

These objects are used across all three model iterations and are created once
here to avoid duplication.

```{r create_monthly_dates}
#| echo: true

monthly_tbl <- tibble(
  date = seq(as.Date("2015-01-01"), as.Date("2016-02-01"), by = "month"),
  weeks = difftime(date, as.Date("2015-01-01"), units = "weeks") |> as.numeric()
  )

monthly_tbl |> glimpse()
```

The `monthly_tbl` tibble defines the evaluation horizon (monthly anchors and
their week offsets from the origin). All downstream simulation and aggregation
steps reference these dates to ensure temporal alignment across models.


```{r create_inforce_subset}
#| echo: true

inforce_tbl <- model_training_tbl |>
  filter(status_at_reference_date == "inforce")

message(glue("Inforce policies: {nrow(inforce_tbl)}"))
```

```{r create_low_premium_inforce_subset}
#| echo: true

# Pre-compute low-premium inforce policy id set for downstream semi_joins
lowprem_inforce_policy_ids_tbl <- inforce_tbl |>
  filter(prem_ape < 12000) |>
  select(policy_id)

message(glue("Low-premium inforce policies (prem_ape < 12k): {nrow(lowprem_inforce_policy_ids_tbl)}"))
```


```{r create_policy_cutdown}
#| echo: true

policy_cutdown_tbl <- model_training_tbl |>
  select(policy_id, policy_startdate, policy_enddate, prem_ape)

policy_cutdown_tbl |> glimpse()
```


```{r extract_cashflow_week_count}
#| echo: true

cashflow_week_count <- monthly_tbl |> pull(weeks)

message(glue("Number of cashflow periods: {length(cashflow_week_count)}"))
```


```{r calculate_actual_cashflows}
#| echo: true

actual_cashflows_tbl <- model_rollforward_tbl |>
  semi_join(inforce_tbl, by = "policy_id") |>
  calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date)

actual_cashflows_tbl |> glimpse()
```


```{r summarise_actual_cashflows}
#| echo: true

actual_monthly_tbl <- actual_cashflows_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )

actual_monthly_tbl |> glimpse()
```


```{r calculate_actual_cashflows_recent}
#| echo: true

actual_cashflows_recent_tbl <- model_rollforward_tbl |>
  filter(policy_startdate >= as.Date("2010-01-01")) |>
  semi_join(inforce_tbl, by = "policy_id") |>
  calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date)

actual_cashflows_recent_tbl |> glimpse()
```


```{r summarise_actual_cashflows_recent}
#| echo: true

actual_monthly_recent_tbl <- actual_cashflows_recent_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )

actual_monthly_recent_tbl |> glimpse()
```

At this point we have three actual cashflow strata prepared: full portfolio,
recent inceptions (>= 2010), and low-premium (< 12k) which follows below. Each
stratum offers a different calibration lens: structural stability (full),
recency sensitivity (recent), and affordability segment behavior (low-premium).


```{r calculate_actual_cashflows_lowprem}
#| echo: true

# Construct cashflows for low-premium policies (prem_ape < 12k) for assessment only
actual_cashflows_lowprem_tbl <- model_rollforward_tbl |>
  semi_join(lowprem_inforce_policy_ids_tbl, by = "policy_id") |>
  calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date)

actual_cashflows_lowprem_tbl |> glimpse()
```


```{r summarise_actual_cashflows_lowprem}
#| echo: true

actual_monthly_lowprem_tbl <- actual_cashflows_lowprem_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )

actual_monthly_lowprem_tbl |> glimpse()
```

Finally, we save some of this data to disk so it can be used again.

```{r save_actual_data}
#| echo: false

actual_cashflow_file <- glue("{data_dir}/actual_cashflow_tbl.parquet")

actual_cashflows_tbl |> write_parquet_compressed(actual_cashflow_file)
```


# Build Initial Cox Model

Model 1 establishes a baseline proportional hazards specification with two
simple covariates (gender and smoker). We fit using both `brms` (for flexible
extensions later) and `rstanarm` (for direct survival utilities) to compare
parameterization approaches and confirm consistent hazard structure.


## brms Model

```{r lapse1_fit_cox_model_brms}
#| echo: true
#| message: false
#| results: hide

lapse1_stanmodel_name <- "lapse1_coxph_brms"

lapse1_coxph_brmsfit <- brm(
  policy_lifetime | cens(1-lapsed) ~ gender_life1 + smoker_life1,
  data            = model_training_tbl,
  family          = cox(link = "log"),
  cores           = 4,
  seed            = 421,
  save_warmup     = FALSE,
  backend         = "cmdstanr",
  save_model      = glue("{stan_codedir}/{lapse1_stanmodel_name}.stan"),
  file            = glue("{stan_modeldir}/{lapse1_stanmodel_name}_brmsfit"),
  output_dir      = stan_modeldir,
  output_basename = lapse1_stanmodel_name,

  stan_model_args = list(
    include_paths   =     "stan_code",
    pedantic        =            TRUE,
    dir             =    "stan_model"
    )
  )
```

```{r lapse1_fit_cox_model_brms_summary}
#| echo: true

summary(lapse1_coxph_brmsfit)
```

The `brms` fit provides a foundation for potential spline or hierarchical
extensions. For operational projections we proceed with the `rstanarm` version
whose survival extraction helpers we leverage in later sections.



## rstanarm Model


```{r lapse1_fit_cox_model_rstanarm}
#| echo: true
#| message: false
#| results: hide

lapse1_stanmodel_name <- "lapse1_coxph_stansurv"
lapse1_stanmodel_file <- glue("{stan_modeldir}/{lapse1_stanmodel_name}.qs")

# Check if fitted model exists, load it if so, otherwise fit new model
if (file_exists(lapse1_stanmodel_file)) {
  message(glue("Loading cached model from {lapse1_stanmodel_file}"))
  lapse1_coxph_stansurv <- qread(lapse1_stanmodel_file)

} else {
  message(glue("Fitting new model, will save to {lapse1_stanmodel_file}"))
  
  # Flush output to ensure visibility in Podman
  cat("Starting Stan model fitting...\n")
  flush.console()
  
  lapse1_coxph_stansurv <- stan_surv(
    Surv(policy_lifetime, lapsed) ~ gender_life1 + smoker_life1,
    data   = model_training_tbl,
    
    # MCMC sampling parameters
    chains = 4,
    iter   = 2000,
    warmup = 1000,
    seed   = 4000,
    cores  = 4,
    
    # Baseline hazard specification
    basehaz     = "ms",  # M-splines (flexible, default)
    basehaz_ops = list(df = 6),  # Degrees of freedom for baseline hazard
    
    # Prior specifications (explicit is better than implicit)
    prior           = rstanarm::normal(location = 0, scale = 2.5), # Weakly informative for coefficients
    prior_intercept = rstanarm::normal(location = 0, scale = 10),  # For baseline hazard
    
    # Enable verbose output for console monitoring (not captured in document)
    refresh = 200  # Print progress every 200 iterations
    )
  
  cat("Model fitting complete.\n")
  flush.console()
  
  # Save fitted model for future use
  lapse1_coxph_stansurv |> qsave(lapse1_stanmodel_file, preset = "balanced")
  message(glue("Model saved to {lapse1_stanmodel_file}"))
}
```

```{r lapse1_fit_cox_model_stansurv_summary}
#| echo: true
#| message: true

summary(lapse1_coxph_stansurv)
```


## Model Diagnostics

We examine the MCMC diagnostics to ensure the model has converged properly and
the posterior samples are reliable. Key metrics include:

- **Rhat**: Should be < 1.01 (measures convergence across chains)
- **ESS ratio**: Should be > 0.1, ideally > 0.5 (effective sample size)
- **Divergences**: Should be 0 (indicates sampling issues)
- **Trace plots**: Should show good mixing without trends

```{r lapse1_model_diagnostics_convergence}
#| echo: true
#| fig-width: 14
#| fig-height: 5

# Check actual parameter names
write_lines("Model parameters:", stdout())
names(coef(lapse1_coxph_stansurv)) |> print()

plot_stan_convergence(
  lapse1_coxph_stansurv,
  title = "Model 1 Convergence Diagnostics: Gender + Smoker"
  )
```


```{r lapse1_model_diagnostics_mixing}
#| echo: true
#| fig-width: 14
#| fig-height: 10

plot_stan_mixing(
  lapse1_coxph_stansurv,
  title = "Model 1 Chain Mixing: Gender + Smoker"
  )
```


# Construct Posterior Survival Estimates

We now want to use our fitted model to predict the lapse for the portfolio.

We model the monthly paid premia in the portfolio for the final year for which
we have data. We can then compare our modelled cashflows against the observed
data.


## Extract Posterior Estimates


```{r lapse1_construct_posterior_estimates}
#| echo: true

lapse1_post_draws_file <- glue("{stan_modeldir}/lapse1_post_draws_lst.qs")

# Check if cached posterior draws exist
if (file_exists(lapse1_post_draws_file)) {
  message(glue("Loading cached posterior draws from {lapse1_post_draws_file}"))
  lapse1_post_draws_lst <- qread(lapse1_post_draws_file)
  
} else {
  message(glue("Computing posterior draws, will save to {lapse1_post_draws_file}"))
  
  lapse1_post_draws_lst <- posterior_survfit(
      lapse1_coxph_stansurv,
      newdata       = inforce_tbl,
      last_time     = NULL,
      draws         = 2000,
      condition     = TRUE,
      return_matrix = TRUE
      )
  
  # Save for future use
  lapse1_post_draws_lst |> qsave(lapse1_post_draws_file, preset = "balanced")
  message(glue("Posterior draws saved to {lapse1_post_draws_file}"))
}

lapse1_post_draws_lst[[1]] |> glimpse()
```

Posterior survival draws are the raw building blocks for projection. We retain
the full matrix form to support interpolation onto weekly cashflow periods and
policy-level lapse time simulation without re-running MCMC.


```{r lapse1_extract_posterior_survival_estimates}
#| echo: true
#| message: false

lapse1_post_survival_file <- glue("{stan_modeldir}/lapse1_post_survival_tbl.parquet")

# Check if cached posterior survival data exists
if (file_exists(lapse1_post_survival_file)) {
  message(glue("Loading cached posterior survival data from {lapse1_post_survival_file}"))
  lapse1_post_survival_tbl <- read_parquet(lapse1_post_survival_file)
  
} else {
  message(glue("Extracting posterior survival data, will save to {lapse1_post_survival_file}"))
  
  lapse1_post_survival_tbl <- lapse1_post_draws_lst |>
    map_dfr(
      extract_stansurv_posterior_survivals,

      .progress = "extract_post_surv"
      ) |>
    filter(time < 5 * 52) |>
    mutate(
      policy_id = inforce_tbl$policy_id[idx],
      
      .before = 1
      ) |>
    select(policy_id, draw_id, time, post_surv_prob = value) |>
    interpolate_survival_data(
      new_times = cashflow_week_count
      ) |>
    group_nest(policy_id, draw_id)
  
  # Save for future use
  lapse1_post_survival_tbl |> write_parquet_compressed(lapse1_post_survival_file)
  message(glue("Posterior survival data saved to {lapse1_post_survival_file}"))
}

lapse1_post_survival_tbl |> glimpse()
```

Interpolated survival trajectories are nested per policy and draw. Nesting
enables efficient map operations for lapse time extraction while keeping memory
footprint manageable (avoiding repeated joins). We truncate at 5 years for a
focused projection window.



```{r lapse1_calculate_sim_lapse_times}
#| echo: true
#| message: false

lapse1_sim_lapse_times_file <- glue("{stan_modeldir}/lapse1_sim_lapse_times_tbl.parquet")

# Check if cached simulated lapse times exist
if (file_exists(lapse1_sim_lapse_times_file)) {
  message(glue("Loading cached simulated lapse times from {lapse1_sim_lapse_times_file}"))
  lapse1_sim_lapse_times_tbl <- read_parquet(lapse1_sim_lapse_times_file)
  
} else {
  message(glue("Computing simulated lapse times, will save to {lapse1_sim_lapse_times_file}"))
  
  lapse1_sim_lapse_times_tbl <- lapse1_post_survival_tbl |>
    mutate(
      rnd_draw   = runif(n(), 0, 1),
      lapse_time = map2_dbl(
        data, rnd_draw,
        \(x, y) extract_survival_time(x, y, min_time = 1),

        .progress = "extract_survival_time"
        )
      ) |>
    select(policy_id, draw_id, lapse_time)
  
  # Save for future use
  lapse1_sim_lapse_times_tbl |> write_parquet_compressed(lapse1_sim_lapse_times_file)
  message(glue("Simulated lapse times saved to {lapse1_sim_lapse_times_file}"))
}

lapse1_sim_lapse_times_tbl |> glimpse()
```

Simulated lapse times convert survival curves into discrete lapse events by
inverse sampling with a uniform deviate per (policy, draw). This preserves the
posterior uncertainty while producing a single lapse realization per draw for
cashflow simulation.



```{r lapse1_construct_policy_sim_data}
#| echo: true

lapse1_sim_policy_data_tbl <- lapse1_sim_lapse_times_tbl |>
  inner_join(policy_cutdown_tbl, by = "policy_id") |>
  mutate(
    lapse_date    = as.Date("2015-01-01") + weeks(lapse_time),
    policy_status = case_when(

      # Policy has expired (term ended)
      (policy_enddate <= lapse_date) & (as.Date("2016-01-01") > policy_enddate) ~ "completed",

      # Policy has lapsed and the lapse occurred before or on the reference date
      (lapse_date < policy_enddate) & (lapse_date <= as.Date("2016-01-01")) ~ "lapsed",

      # Policy is within term and either not lapsed or lapsed after reference date
      as.Date("2016-01-01") <= policy_enddate ~ "inforce",

      # Default case (shouldn't occur with proper data)
      TRUE ~ "unknown"
      ),
    policy_statuschangedate = case_when(
      policy_status == "lapsed"    ~ lapse_date,
      policy_status == "completed" ~ policy_enddate,
      policy_status == "inforce"   ~ policy_startdate,
      TRUE                         ~ as.Date(NA)
      )
    )


lapse1_sim_policy_data_tbl |> glimpse()
```

Augmenting simulated lapse times with policy metadata allows classification of
end states (lapsed, completed, inforce). These statuses drive premium cessation
logic in subsequent cashflow generation.


## Construct Simulated Cashflow Data


```{r lapse1_construct_simulated_cashflows}
#| echo: true
#| message: false

lapse1_sim_cashflow_file <- glue("{stan_modeldir}/lapse1_sim_cashflow_tbl.parquet")

# Check if cached simulated cashflow data exists
if (file_exists(lapse1_sim_cashflow_file)) {
  message(glue("Loading cached simulated cashflows from {lapse1_sim_cashflow_file}"))
  lapse1_sim_cashflow_tbl <- read_parquet(lapse1_sim_cashflow_file)
  
} else {
  message(glue("Computing simulated cashflows, will save to {lapse1_sim_cashflow_file}"))
  
  lapse1_sim_cashflow_tbl <- lapse1_sim_policy_data_tbl |>
    group_nest(draw_id) |>
    mutate(
      newdata = map(
        data, \(x) x |> calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date),
                    
        .progress = "joining_data")
        ) |>
    select(-data) |>
    unnest(newdata) |>
    # Reattach prem_ape (dropped by calculate_monthly_cashflows) for segment filtering
    left_join(policy_cutdown_tbl |> select(policy_id, prem_ape), by = "policy_id")
  
  # Save for future use
  lapse1_sim_cashflow_tbl |> write_parquet_compressed(lapse1_sim_cashflow_file)
  message(glue("Simulated cashflows saved to {lapse1_sim_cashflow_file}"))
}

lapse1_sim_cashflow_tbl |> glimpse()
```


```{r lapse1_construct_simulated_monthly_cashflows}
#| echo: true

lapse1_sim_monthly_cashflow_tbl <- lapse1_sim_cashflow_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )

lapse1_sim_monthly_cashflow_tbl |> glimpse()
```


```{r lapse1_construct_comparison_plot}
#| echo: true

ggplot(lapse1_sim_monthly_cashflow_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title = glue("Simulated vs Actual Cashflows for Model {lapse1_stanmodel_name}")
    )
```

We create a similar plot using ribbons using quantiles of the simulated
cashflows.

```{r lapse1_construct_comparison_plot_ribbons}
#| echo: true

lapse1_sim_monthly_bands_tbl <- lapse1_sim_monthly_cashflow_tbl |>
  summarise(
    p10 = quantile(total_cashflow, 0.10),
    p25 = quantile(total_cashflow, 0.25),
    p50 = quantile(total_cashflow, 0.50),
    p75 = quantile(total_cashflow, 0.75),
    p90 = quantile(total_cashflow, 0.90),
    
    .by = c(cashflow_date)
    )

ggplot(lapse1_sim_monthly_bands_tbl) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p10, ymax = p90),
    alpha = 0.5
    ) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p25, ymax = p75),
    alpha = 0.5
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title    = glue("Simulated vs Actual Cashflows for Model {lapse1_stanmodel_name}"),
    subtitle = "Showing 50% and 80% bands"
    )
```


### Recent Policies Only (2010+)

We now repeat both visualizations, but restrict the data to only policies that
have a start date from 2010-01-01 onwards, so that we are focusing solely on
more recently incepted policies.

```{r lapse1_filter_recent_policies}
#| echo: true

# Filter simulated cashflows for recent policies
lapse1_sim_monthly_cashflow_recent_tbl <- lapse1_sim_cashflow_tbl |>
  filter(policy_startdate >= as.Date("2010-01-01")) |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )
```

```{r lapse1_construct_comparison_plot_recent}
#| echo: true

ggplot(lapse1_sim_monthly_cashflow_recent_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_recent_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title = glue("Simulated vs Actual Cashflows for Model {lapse1_stanmodel_name}"),
    subtitle = "Recent policies (start date >= 2010-01-01)"
    )
```

```{r lapse1_construct_comparison_plot_ribbons_recent}
#| echo: true

lapse1_sim_monthly_bands_recent_tbl <- lapse1_sim_monthly_cashflow_recent_tbl |>
  summarise(
    p10 = quantile(total_cashflow, 0.10),
    p25 = quantile(total_cashflow, 0.25),
    p50 = quantile(total_cashflow, 0.50),
    p75 = quantile(total_cashflow, 0.75),
    p90 = quantile(total_cashflow, 0.90),
    
    .by = c(cashflow_date)
    )

ggplot(lapse1_sim_monthly_bands_recent_tbl) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p10, ymax = p90),
    alpha = 0.5
    ) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p25, ymax = p75),
    alpha = 0.5
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_recent_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title    = glue("Simulated vs Actual Cashflows for Model {lapse1_stanmodel_name}"),
    subtitle = "Recent policies (start date >= 2010-01-01), showing 50% and 80% bands"
    )
```


### Low Premium Policies (prem_ape < 12k)

We repeat both visualisations focusing only on policies with **prem_ape < 15,000**.
This subset is used purely for assessment; models were trained on the full
small dataset. Differences here can highlight calibration issues for lower-premium
business.

```{r lapse1_filter_low_premium_policies}
#| echo: true

lapse1_sim_monthly_cashflow_lowprem_tbl <- lapse1_sim_cashflow_tbl |>
  semi_join(lowprem_inforce_policy_ids_tbl, by = "policy_id") |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )
```

```{r lapse1_construct_comparison_plot_lowprem}
#| echo: true

ggplot(lapse1_sim_monthly_cashflow_lowprem_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_lowprem_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title = glue("Simulated vs Actual Cashflows (Low Premium) for Model {lapse1_stanmodel_name}"),
    subtitle = "Policies with prem_ape < 12k"
    )
```

```{r lapse1_construct_comparison_plot_ribbons_lowprem}
#| echo: true

lapse1_sim_monthly_bands_lowprem_tbl <- lapse1_sim_monthly_cashflow_lowprem_tbl |>
  summarise(
    p10 = quantile(total_cashflow, 0.10),
    p25 = quantile(total_cashflow, 0.25),
    p50 = quantile(total_cashflow, 0.50),
    p75 = quantile(total_cashflow, 0.75),
    p90 = quantile(total_cashflow, 0.90),
    
    .by = c(cashflow_date)
    )

ggplot(lapse1_sim_monthly_bands_lowprem_tbl) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p10, ymax = p90),
    alpha = 0.5
    ) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p25, ymax = p75),
    alpha = 0.5
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_lowprem_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title    = glue("Simulated vs Actual Cashflows (Low Premium) for Model {lapse1_stanmodel_name}"),
    subtitle = "Policies with prem_ape < 12k, showing 50% and 80% bands"
    )
```


We now clean up the environment before building the next model.

```{r lapse1_clear_large_objects}
#| echo: false

# Save tibbles before removing
lapse1_post_survival_tbl |> write_parquet(glue("{stan_modeldir}/lapse1_post_survival_tbl.parquet"))
lapse1_sim_cashflow_tbl  |> write_parquet(glue("{stan_modeldir}/lapse1_sim_cashflow_tbl.parquet"))

# Remove large objects
rm(lapse1_post_survival_tbl, lapse1_post_draws_lst, lapse1_sim_cashflow_tbl)

gc(verbose = TRUE)
```




# Build Expanded Cox Model

We now want to add a number of covariates to the model to see if we can improve
it.

## rstanarm Model


```{r lapse2_fit_cox_model_rstanarm}
#| echo: true
#| results: hide
#| message: false

lapse2_stanmodel_name <- "lapse2_coxph_stansurv"
lapse2_stanmodel_file <- glue("{stan_modeldir}/{lapse2_stanmodel_name}.qs")

# Check if fitted model exists, load it if so, otherwise fit new model
if (file_exists(lapse2_stanmodel_file)) {
  message(glue("Loading cached model from {lapse2_stanmodel_file}"))
  lapse2_coxph_stansurv <- qread(lapse2_stanmodel_file)

} else {
  message(glue("Fitting new model, will save to {lapse2_stanmodel_file}"))

  # Create stan_model directory if it doesn't exist
  dir_create(stan_modeldir, recurse = TRUE)
  
  lapse2_coxph_stansurv <- stan_surv(
    Surv(policy_lifetime, lapsed) ~ gender_life1 + smoker_life1 + cluster_id + prem_sa_ratio,
    data   = model_training_tbl,
    
    # MCMC sampling parameters
    chains = 4,
    iter   = 2000,
    warmup = 1000,
    seed   = 4000,
    cores  = 4,
    
    # Baseline hazard specification
    basehaz     = "ms",  # M-splines (flexible, default)
    basehaz_ops = list(df = 6),  # Degrees of freedom for baseline hazard
    
    # Prior specifications (explicit is better than implicit)
    prior           = rstanarm::normal(location = 0, scale = 2.5), # Weakly informative for coefficients
    prior_intercept = rstanarm::normal(location = 0, scale = 10),  # For baseline hazard
    
    # Enable verbose output for console monitoring (not captured in document)
    refresh = 200  # Print progress every 200 iterations
    )
  
  # Save fitted model for future use
  lapse2_coxph_stansurv |> qsave(lapse2_stanmodel_file, preset = "balanced")
  message(glue("Model saved to {lapse2_stanmodel_file}"))
}
```

```{r lapse2_fit_cox_model_stansurv_summary}
#| echo: true
#| message: true

summary(lapse2_coxph_stansurv)
```

Model 2 adds structural and financial covariates (`cluster_id`, `prem_sa_ratio`)
to capture distribution channel heterogeneity and pricing relative to sum
assured. Improved explanatory power should narrow predictive bands if these
features materially influence lapse propensity.


## Model Diagnostics

We examine the MCMC diagnostics for the expanded model that includes cluster
effects and premium-to-sum-assured ratio.

```{r lapse2_model_diagnostics_convergence}
#| echo: true
#| fig-width: 14
#| fig-height: 5

plot_stan_convergence(
  lapse2_coxph_stansurv,
  title = "Model 2 Convergence Diagnostics: Gender + Smoker + Cluster + Premium Ratio"
  )
```


```{r lapse2_model_diagnostics_mixing}
#| echo: true
#| fig-width: 14
#| fig-height: 10

plot_stan_mixing(
  lapse2_coxph_stansurv,
  title = "Model 2 Chain Mixing: Gender + Smoker + Cluster + Premium Ratio"
  )
```


## Construct Posterior Survival Estimates

We now want to use our fitted model to predict the lapse for the portfolio.

We model the monthly paid premia in the portfolio for the final year for which
we have data. We can then compare our modelled cashflows against the observed
data.

```{r lapse2_construct_rstanarm_posterior_estimates}
#| echo: true

lapse2_post_draws_file <- glue("{stan_modeldir}/lapse2_post_draws_lst.qs")

# Check if cached posterior draws exist
if (file_exists(lapse2_post_draws_file)) {
  message(glue("Loading cached posterior draws from {lapse2_post_draws_file}"))
  lapse2_post_draws_lst <- qread(lapse2_post_draws_file)
  
} else {
  message(glue("Computing posterior draws, will save to {lapse2_post_draws_file}"))
  
  lapse2_post_draws_lst <- posterior_survfit(
      lapse2_coxph_stansurv,
      newdata       = inforce_tbl,
      last_time     = NULL,
      draws         = 2000,
      condition     = TRUE,
      return_matrix = TRUE
      )
  
  # Save for future use
  lapse2_post_draws_lst |> qsave(lapse2_post_draws_file, preset = "balanced")
  message(glue("Posterior draws saved to {lapse2_post_draws_file}"))
}

lapse2_post_draws_lst[[1]] |> glimpse()
```


```{r lapse2_extract_posterior_survival_estimates}
#| echo: true
#| message: false

lapse2_cashflow_week_count <- monthly_tbl |> pull(weeks)

lapse2_post_survival_file <- glue("{stan_modeldir}/lapse2_post_survival_tbl.parquet")

# Check if cached posterior survival data exists
if (file_exists(lapse2_post_survival_file)) {
  message(glue("Loading cached posterior survival data from {lapse2_post_survival_file}"))
  lapse2_post_survival_tbl <- read_parquet(lapse2_post_survival_file)
  
} else {
  message(glue("Extracting posterior survival data, will save to {lapse2_post_survival_file}"))
  
  lapse2_post_survival_tbl <- lapse2_post_draws_lst |>
    map_dfr(
      extract_stansurv_posterior_survivals,

      .progress = "extract_post_surv"
      ) |>
    filter(time < 5 * 52) |>
    mutate(
      policy_id = inforce_tbl$policy_id[idx],
      
      .before = 1
      ) |>
    select(policy_id, draw_id, time, post_surv_prob = value) |>
    interpolate_survival_data(
      new_times = lapse2_cashflow_week_count
      ) |>
    group_nest(policy_id, draw_id)
  
  # Save for future use
  lapse2_post_survival_tbl |> write_parquet_compressed(lapse2_post_survival_file)
  message(glue("Posterior survival data saved to {lapse2_post_survival_file}"))
}

lapse2_post_survival_tbl |> glimpse()
```

Survival extraction mirrors Model 1: we interpolate onto weekly intervals. Any
shift in survival dispersion versus the baseline model indicates covariate
contribution to reducing residual variance.



```{r lapse2_calculate_sim_lapse_times}
#| echo: true
#| message: false

lapse2_sim_lapse_times_file <- glue("{stan_modeldir}/lapse2_sim_lapse_times_tbl.parquet")

# Check if cached simulated lapse times exist
if (file_exists(lapse2_sim_lapse_times_file)) {
  message(glue("Loading cached simulated lapse times from {lapse2_sim_lapse_times_file}"))
  lapse2_sim_lapse_times_tbl <- read_parquet(lapse2_sim_lapse_times_file)
  
} else {
  message(glue("Computing simulated lapse times, will save to {lapse2_sim_lapse_times_file}"))
  
  lapse2_sim_lapse_times_tbl <- lapse2_post_survival_tbl |>
    mutate(
      rnd_draw   = runif(n(), 0, 1),
      lapse_time = map2_dbl(
        data, rnd_draw,
        \(x, y) extract_survival_time(x, y, min_time = 1),

        .progress = "extract_survival_time"
        )
      ) |>
    select(policy_id, draw_id, lapse_time)
  
  # Save for future use
  lapse2_sim_lapse_times_tbl |> write_parquet_compressed(lapse2_sim_lapse_times_file)
  message(glue("Simulated lapse times saved to {lapse2_sim_lapse_times_file}"))
}

lapse2_sim_lapse_times_tbl |> glimpse()
```

Comparing the distribution of simulated lapse times here versus Model 1 helps
assess whether added covariates shift early versus late lapse concentrations,
informing potential cashflow timing adjustments.



```{r lapse2_construct_policy_sim_data}
#| echo: true

lapse2_sim_policy_data_tbl <- lapse2_sim_lapse_times_tbl |>
  inner_join(policy_cutdown_tbl, by = "policy_id") |>
  mutate(
    lapse_date    = as.Date("2015-01-01") + weeks(lapse_time),
    policy_status = case_when(

      # Policy has expired (term ended)
      (policy_enddate <= lapse_date) & (as.Date("2016-01-01") > policy_enddate) ~ "completed",

      # Policy has lapsed and the lapse occurred before or on the reference date
      (lapse_date < policy_enddate) & (lapse_date <= as.Date("2016-01-01")) ~ "lapsed",

      # Policy is within term and either not lapsed or lapsed after reference date
      as.Date("2016-01-01") <= policy_enddate ~ "inforce",

      # Default case (shouldn't occur with proper data)
      TRUE ~ "unknown"
      ),
    policy_statuschangedate = case_when(
      policy_status == "lapsed"    ~ lapse_date,
      policy_status == "completed" ~ policy_enddate,
      policy_status == "inforce"   ~ policy_startdate,
      TRUE                         ~ as.Date(NA)
      )
    )


lapse2_sim_policy_data_tbl |> glimpse()
```

Policy status derivation is identical; differences in counts per status category
across models highlight portfolio segments where covariates change lapse timing.



We now construct the simulated cashflows from the simulated data.


```{r lapse2_construct_simulated_cashflows}
#| echo: true
#| message: false

lapse2_sim_cashflow_file <- glue("{stan_modeldir}/lapse2_sim_cashflow_tbl.parquet")

# Check if cached simulated cashflow data exists
if (file_exists(lapse2_sim_cashflow_file)) {
  message(glue("Loading cached simulated cashflows from {lapse2_sim_cashflow_file}"))
  lapse2_sim_cashflow_tbl <- read_parquet(lapse2_sim_cashflow_file)
  
} else {
  message(glue("Computing simulated cashflows, will save to {lapse2_sim_cashflow_file}"))
  
  lapse2_sim_cashflow_tbl <- lapse2_sim_policy_data_tbl |>
    group_nest(draw_id) |>
    mutate(
      newdata = map(
        data, \(x) x |> calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date),
                    
        .progress = "joining_data")
        ) |>
    select(-data) |>
    unnest(newdata) |>
    left_join(policy_cutdown_tbl |> select(policy_id, prem_ape), by = "policy_id")
  
  # Save for future use
  lapse2_sim_cashflow_tbl |> write_parquet_compressed(lapse2_sim_cashflow_file)
  message(glue("Simulated cashflows saved to {lapse2_sim_cashflow_file}"))
}

lapse2_sim_cashflow_tbl |> glimpse()
```

Simulated cashflows aggregate premiums until lapse or completion. We later
benchmark aggregated monthly totals against observed strata to evaluate
calibration improvements from the expanded covariate set.


```{r lapse2_construct_simulated_monthly_cashflows}
#| echo: true

lapse2_sim_monthly_cashflow_tbl <- lapse2_sim_cashflow_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )

lapse2_sim_monthly_cashflow_tbl |> glimpse()
```


```{r lapse2_construct_comparison_plot}
#| echo: true

ggplot(lapse2_sim_monthly_cashflow_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title = glue("Simulated vs Actual Cashflows for Model {lapse2_stanmodel_name}")
    )
```

We create a similar plot using ribbons using quantiles of the simulated
cashflows.

```{r lapse2_construct_comparison_plot_ribbons}
#| echo: true

lapse2_sim_monthly_bands_tbl <- lapse2_sim_monthly_cashflow_tbl |>
  summarise(
    p10 = quantile(total_cashflow, 0.10),
    p25 = quantile(total_cashflow, 0.25),
    p50 = quantile(total_cashflow, 0.50),
    p75 = quantile(total_cashflow, 0.75),
    p90 = quantile(total_cashflow, 0.90),
    
    .by = c(cashflow_date)
    )

ggplot(lapse2_sim_monthly_bands_tbl) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p10, ymax = p90),
    alpha = 0.5
    ) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p25, ymax = p75),
    alpha = 0.5
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title    = glue("Simulated vs Actual Cashflows for Model {lapse2_stanmodel_name}"),
    subtitle = "Showing 50% and 80% bands"
    )
```

### Recent Policies Only (2010+)

We now repeat both visualizations, but restrict the data to only policies that
have a start date from 2010-01-01 onwards, so that we are focusing solely on
more recently incepted policies.

```{r lapse2_filter_recent_policies}
#| echo: true

# Filter simulated cashflows for recent policies
lapse2_sim_monthly_cashflow_recent_tbl <- lapse2_sim_cashflow_tbl |>
  filter(policy_startdate >= as.Date("2010-01-01")) |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )
```

```{r lapse2_construct_comparison_plot_recent}
#| echo: true

ggplot(lapse2_sim_monthly_cashflow_recent_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_recent_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title = glue("Simulated vs Actual Cashflows for Model {lapse2_stanmodel_name}"),
    subtitle = "Recent policies (start date >= 2010-01-01)"
    )
```

```{r lapse2_construct_comparison_plot_ribbons_recent}
#| echo: true

lapse2_sim_monthly_bands_recent_tbl <- lapse2_sim_monthly_cashflow_recent_tbl |>
  summarise(
    p10 = quantile(total_cashflow, 0.10),
    p25 = quantile(total_cashflow, 0.25),
    p50 = quantile(total_cashflow, 0.50),
    p75 = quantile(total_cashflow, 0.75),
    p90 = quantile(total_cashflow, 0.90),
    
    .by = c(cashflow_date)
    )

ggplot(lapse2_sim_monthly_bands_recent_tbl) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p10, ymax = p90),
    alpha = 0.5
    ) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p25, ymax = p75),
    alpha = 0.5
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_recent_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title    = glue("Simulated vs Actual Cashflows for Model {lapse2_stanmodel_name}"),
    subtitle = "Recent policies (start date >= 2010-01-01), showing 50% and 80% bands"
    )
```



### Low Premium Policies (prem_ape < 12k)

As with Model 1 we assess performance on low-premium business. This is an
out-of-segment evaluation; the model was fit on the full small dataset.

These policies typically have different lapse dynamics due to payment burden
and distribution channel mix. Adding this slice helps us visually check for
systematic mis-calibration (e.g., persistent divergence between the red actual
line and the central tendency of simulations) without refitting. We retain the
original posterior draws and only re-aggregate.

```{r lapse2_filter_low_premium_policies}
#| echo: true

lapse2_sim_monthly_cashflow_lowprem_tbl <- lapse2_sim_cashflow_tbl |>
  semi_join(lowprem_inforce_policy_ids_tbl, by = "policy_id") |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )
```

```{r lapse2_construct_comparison_plot_lowprem}
#| echo: true

ggplot(lapse2_sim_monthly_cashflow_lowprem_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_lowprem_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title = glue("Simulated vs Actual Cashflows (Low Premium) for Model {lapse2_stanmodel_name}"),
    subtitle = "Policies with prem_ape < 12k"
    )
```

```{r lapse2_construct_comparison_plot_ribbons_lowprem}
#| echo: true

lapse2_sim_monthly_bands_lowprem_tbl <- lapse2_sim_monthly_cashflow_lowprem_tbl |>
  summarise(
    p10 = quantile(total_cashflow, 0.10),
    p25 = quantile(total_cashflow, 0.25),
    p50 = quantile(total_cashflow, 0.50),
    p75 = quantile(total_cashflow, 0.75),
    p90 = quantile(total_cashflow, 0.90),
    
    .by = c(cashflow_date)
    )

ggplot(lapse2_sim_monthly_bands_lowprem_tbl) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p10, ymax = p90),
    alpha = 0.5
    ) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p25, ymax = p75),
    alpha = 0.5
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_lowprem_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title    = glue("Simulated vs Actual Cashflows (Low Premium) for Model {lapse2_stanmodel_name}"),
    subtitle = "Policies with prem_ape < 12k, showing 50% and 80% bands"
    )
```

  The width of the predictive bands shows uncertainty in aggregate lapse-driven
  cashflows for low-premium business. Alignment between the red actual path and
  the median (p50) trajectory suggests calibration is broadly acceptable. If the
  actual line sits near the edge of the 80% band for extended periods we may need
  segment-specific modeling or additional covariates (e.g., channel indicators,
  payment method). Short-term spikes outside the 50% band but inside the 80% band
  are typically tolerable given portfolio volatility.


We now clean up the environment before building the next model.

```{r lapse2_clear_large_objects}
#| echo: false

# Save tibbles before removing
lapse2_post_survival_tbl |>
  write_parquet(glue("{stan_modeldir}/lapse2_post_survival_tbl.parquet"))

lapse2_sim_cashflow_tbl  |>
  write_parquet(glue("{stan_modeldir}/lapse2_sim_cashflow_tbl.parquet"))

# Remove large objects
rm(lapse2_post_survival_tbl, lapse2_post_draws_lst, lapse2_sim_cashflow_tbl)

gc(verbose = TRUE)
```



# Add Time-Dependent Parameters

We now want to add a number of covariates to the model to see if we can improve
it.

## rstanarm Model


```{r lapse3_fit_cox_model_rstanarm}
#| echo: true
#| message: false

lapse3_stanmodel_name <- "lapse3_coxph_stansurv"
lapse3_stanmodel_file <- glue("{stan_modeldir}/{lapse3_stanmodel_name}.qs")

# Check if fitted model exists, load it if so, otherwise fit new model
if (file_exists(lapse3_stanmodel_file)) {
  message(glue("Loading cached model from {lapse3_stanmodel_file}"))
  lapse3_coxph_stansurv <- qread(lapse3_stanmodel_file)

} else {
  message(glue("Fitting new model, will save to {lapse3_stanmodel_file}"))

  # Create stan_model directory if it doesn't exist
  dir_create(stan_modeldir, recurse = TRUE)
  
  lapse3_coxph_stansurv <- stan_surv(
    Surv(policy_lifetime, lapsed) ~ gender_life1 + tve(smoker_life1, df = 3, degree = 1),
    data   = model_training_tbl,
    
    # MCMC sampling parameters
    chains = 4,
    iter   = 2000,
    warmup = 1000,
    seed   = 4000,
    cores  = 4,
    
    # Baseline hazard specification
    basehaz     = "ms",  # M-splines (flexible, default)
    basehaz_ops = list(df = 6),  # Degrees of freedom for baseline hazard
    
    # Prior specifications (explicit is better than implicit)
    prior           = rstanarm::normal(location = 0, scale = 2.5), # Weakly informative for coefficients
    prior_intercept = rstanarm::normal(location = 0, scale = 10),  # For baseline hazard
    
    # Enable verbose output for console monitoring (not captured in document)
    refresh = 200  # Print progress every 200 iterations
    )
  
  # Save fitted model for future use
  lapse3_coxph_stansurv |> qsave(lapse3_stanmodel_file, preset = "balanced")
  message(glue("Model saved to {lapse3_stanmodel_file}"))
}
```

```{r lapse3_fit_cox_model_rstanarm_summary}
#| echo: true
#| message: true

summary(lapse3_coxph_stansurv)
```

Model 3 introduces a time-varying effect for `smoker_life1` via a spline term.
This permits differential hazard evolution over policy lifetime—capturing, for
example, front-loaded or deferred lapse risk among smokers relative to non-smokers.


## Model Diagnostics

We examine the MCMC diagnostics for the time-varying effects model. This model
includes a spline-based time-varying effect for the smoker status, allowing
the hazard ratio to change over policy lifetime.

```{r lapse3_model_diagnostics_convergence}
#| echo: true
#| fig-width: 14
#| fig-height: 5

plot_stan_convergence(
  lapse3_coxph_stansurv,
  title = "Model 3 Convergence Diagnostics: Gender + Time-Varying Smoker Effect"
  )
```


```{r lapse3_model_diagnostics_mixing}
#| echo: true
#| fig-width: 14
#| fig-height: 10

plot_stan_mixing(
  lapse3_coxph_stansurv,
  title = "Model 3 Chain Mixing: Gender + Time-Varying Smoker Effect"
  )
```


## Construct Posterior Survival Estimates

We now want to use our fitted model to predict the lapse for the portfolio.

We model the monthly paid premia in the portfolio for the final year for which
we have data. We can then compare our modelled cashflows against the observed
data.

```{r lapse3_construct_rstanarm_posterior_estimates}
#| echo: true

lapse3_post_draws_file <- glue("{stan_modeldir}/lapse3_post_draws_lst.qs")

# Check if cached posterior draws exist
if (file_exists(lapse3_post_draws_file)) {
  message(glue("Loading cached posterior draws from {lapse3_post_draws_file}"))
  lapse3_post_draws_lst <- qread(lapse3_post_draws_file)
  
} else {
  message(glue("Computing posterior draws, will save to {lapse3_post_draws_file}"))
  
  lapse3_post_draws_lst <- posterior_survfit(
      lapse3_coxph_stansurv,
      newdata       = inforce_tbl,
      last_time     = NULL,
      draws         = 2000,
      condition     = TRUE,
      return_matrix = TRUE
      )
  
  # Save for future use
  lapse3_post_draws_lst |> qsave(lapse3_post_draws_file, preset = "balanced")
  message(glue("Posterior draws saved to {lapse3_post_draws_file}"))
}

lapse3_post_draws_lst[[1]] |> glimpse()
```


```{r lapse3_extract_posterior_survival_estimates}
#| echo: true
#| message: false

lapse3_cashflow_week_count <- monthly_tbl |> pull(weeks)

lapse3_post_survival_file <- glue("{stan_modeldir}/lapse3_post_survival_tbl.parquet")

# Check if cached posterior survival data exists
if (file_exists(lapse3_post_survival_file)) {
  message(glue("Loading cached posterior survival data from {lapse3_post_survival_file}"))
  lapse3_post_survival_tbl <- read_parquet(lapse3_post_survival_file)
  
} else {
  message(glue("Extracting posterior survival data, will save to {lapse3_post_survival_file}"))
  
  lapse3_post_survival_tbl <- lapse3_post_draws_lst |>
    map_dfr(
      extract_stansurv_posterior_survivals,

      .progress = "extract_post_surv"
      ) |>
    filter(time < 5 * 52) |>
    mutate(
      policy_id = inforce_tbl$policy_id[idx],
      
      .before = 1
      ) |>
    select(policy_id, draw_id, time, post_surv_prob = value) |>
    interpolate_survival_data(
      new_times = lapse3_cashflow_week_count
      ) |>
    group_nest(policy_id, draw_id)
  
  # Save for future use
  lapse3_post_survival_tbl |> write_parquet_compressed(lapse3_post_survival_file)
  message(glue("Posterior survival data saved to {lapse3_post_survival_file}"))
}

lapse3_post_survival_tbl |> glimpse()
```

Time-varying effects alter the curvature of survival trajectories. Inspecting
dispersion here versus prior models can reveal whether dynamic covariates
reduce mid-term uncertainty or simply re-allocate variance across time.



```{r lapse3_calculate_sim_lapse_times}
#| echo: true
#| message: false

lapse3_sim_lapse_times_file <- glue("{stan_modeldir}/lapse3_sim_lapse_times_tbl.parquet")

# Check if cached simulated lapse times exist
if (file_exists(lapse3_sim_lapse_times_file)) {
  message(glue("Loading cached simulated lapse times from {lapse3_sim_lapse_times_file}"))
  lapse3_sim_lapse_times_tbl <- read_parquet(lapse3_sim_lapse_times_file)
  
} else {
  message(glue("Computing simulated lapse times, will save to {lapse3_sim_lapse_times_file}"))
  
  lapse3_sim_lapse_times_tbl <- lapse3_post_survival_tbl |>
    mutate(
      rnd_draw   = runif(n(), 0, 1),
      lapse_time = map2_dbl(
        data, rnd_draw,
        \(x, y) extract_survival_time(x, y, min_time = 1),

        .progress = "extract_survival_time"
        )
      ) |>
    select(policy_id, draw_id, lapse_time)
  
  # Save for future use
  lapse3_sim_lapse_times_tbl |> write_parquet_compressed(lapse3_sim_lapse_times_file)
  message(glue("Simulated lapse times saved to {lapse3_sim_lapse_times_file}"))
}

lapse3_sim_lapse_times_tbl |> glimpse()
```

Lapse time simulations now reflect dynamic smoker influence. Any systematic
shift (earlier or later median lapse) compared to Model 2 may justify retaining
the time-varying specification despite added complexity.



```{r lapse3_construct_policy_sim_data}
#| echo: true

lapse3_sim_policy_data_tbl <- lapse3_sim_lapse_times_tbl |>
  inner_join(policy_cutdown_tbl, by = "policy_id") |>
  mutate(
    lapse_date    = as.Date("2015-01-01") + weeks(lapse_time),
    policy_status = case_when(

      # Policy has expired (term ended)
      (policy_enddate <= lapse_date) & (as.Date("2016-01-01") > policy_enddate) ~ "completed",

      # Policy has lapsed and the lapse occurred before or on the reference date
      (lapse_date < policy_enddate) & (lapse_date <= as.Date("2016-01-01")) ~ "lapsed",

      # Policy is within term and either not lapsed or lapsed after reference date
      as.Date("2016-01-01") <= policy_enddate ~ "inforce",

      # Default case (shouldn't occur with proper data)
      TRUE ~ "unknown"
      ),
    policy_statuschangedate = case_when(
      policy_status == "lapsed"    ~ lapse_date,
      policy_status == "completed" ~ policy_enddate,
      policy_status == "inforce"   ~ policy_startdate,
      TRUE                         ~ as.Date(NA)
      )
    )


lapse3_sim_policy_data_tbl |> glimpse()
```

Status classification logic remains unchanged; differences in lapsed vs inforce
counts across models help quantify impact of the time-varying term on projected
persistency.



We now construct the simulated cashflows from the simulated data.


```{r lapse3_construct_simulated_cashflows}
#| echo: true
#| message: false

lapse3_sim_cashflow_file <- glue("{stan_modeldir}/lapse3_sim_cashflow_tbl.parquet")

# Check if cached simulated cashflow data exists
if (file_exists(lapse3_sim_cashflow_file)) {
  message(glue("Loading cached simulated cashflows from {lapse3_sim_cashflow_file}"))
  lapse3_sim_cashflow_tbl <- read_parquet(lapse3_sim_cashflow_file)
  
} else {
  message(glue("Computing simulated cashflows, will save to {lapse3_sim_cashflow_file}"))
  
  lapse3_sim_cashflow_tbl <- lapse3_sim_policy_data_tbl |>
    group_nest(draw_id) |>
    mutate(
      newdata = map(
        data, \(x) x |> calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date),

        .progress = "joining_data")
        ) |>
    select(-data) |>
    unnest(newdata) |>
    left_join(policy_cutdown_tbl |> select(policy_id, prem_ape), by = "policy_id")
  
  # Save for future use
  lapse3_sim_cashflow_tbl |> write_parquet_compressed(lapse3_sim_cashflow_file)
  message(glue("Simulated cashflows saved to {lapse3_sim_cashflow_file}"))
}

lapse3_sim_cashflow_tbl |> glimpse()
```

Cashflow construction again nests draw-level simulations. We will compare ribbon
plots across models to judge whether dynamic effects tighten credible intervals
for segments like recent or low-premium policies.


```{r lapse3_construct_simulated_monthly_cashflows}
#| echo: true

lapse3_sim_monthly_cashflow_tbl <- lapse3_sim_cashflow_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )

lapse3_sim_monthly_cashflow_tbl |> glimpse()
```


```{r lapse3_construct_comparison_plot}
#| echo: true

ggplot(lapse3_sim_monthly_cashflow_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title = glue("Simulated vs Actual Cashflows for Model {lapse3_stanmodel_name}")
    )
```

We create a similar plot using ribbons using quantiles of the simulated
cashflows.

```{r lapse3_construct_comparison_plot_ribbons}
#| echo: true

lapse3_sim_monthly_bands_tbl <- lapse3_sim_monthly_cashflow_tbl |>
  summarise(
    p10 = quantile(total_cashflow, 0.10),
    p25 = quantile(total_cashflow, 0.25),
    p50 = quantile(total_cashflow, 0.50),
    p75 = quantile(total_cashflow, 0.75),
    p90 = quantile(total_cashflow, 0.90),
    
    .by = c(cashflow_date)
    )

ggplot(lapse3_sim_monthly_bands_tbl) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p10, ymax = p90),
    alpha = 0.5
    ) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p25, ymax = p75),
    alpha = 0.5
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title    = glue("Simulated vs Actual Cashflows for Model {lapse3_stanmodel_name}"),
    subtitle = "Showing 50% and 80% bands"
    )
```


### Recent Policies Only (2010+)

We now repeat both visualizations, but restrict the data to only policies that
have a start date from 2010-01-01 onwards, so that we are focusing solely on
more recently incepted policies.

```{r lapse3_filter_recent_policies}
#| echo: true

# Filter simulated cashflows for recent policies
lapse3_sim_monthly_cashflow_recent_tbl <- lapse3_sim_cashflow_tbl |>
  filter(policy_startdate >= as.Date("2010-01-01")) |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )
```

```{r lapse3_construct_comparison_plot_recent}
#| echo: true

ggplot(lapse3_sim_monthly_cashflow_recent_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_recent_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title = glue("Simulated vs Actual Cashflows for Model {lapse3_stanmodel_name}"),
    subtitle = "Recent policies (start date >= 2010-01-01)"
    )
```

```{r lapse3_construct_comparison_plot_ribbons_recent}
#| echo: true

lapse3_sim_monthly_bands_recent_tbl <- lapse3_sim_monthly_cashflow_recent_tbl |>
  summarise(
    p10 = quantile(total_cashflow, 0.10),
    p25 = quantile(total_cashflow, 0.25),
    p50 = quantile(total_cashflow, 0.50),
    p75 = quantile(total_cashflow, 0.75),
    p90 = quantile(total_cashflow, 0.90),
    
    .by = c(cashflow_date)
    )

ggplot(lapse3_sim_monthly_bands_recent_tbl) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p10, ymax = p90),
    alpha = 0.5
    ) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p25, ymax = p75),
    alpha = 0.5
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_recent_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title    = glue("Simulated vs Actual Cashflows for Model {lapse3_stanmodel_name}"),
    subtitle = "Recent policies (start date >= 2010-01-01), showing 50% and 80% bands"
    )
```


### Low Premium Policies (prem_ape < 12k)

We repeat the low-premium evaluation for the time-varying smoker effect model.
Again, no refit is performed; this is purely an assessment subset.

Because lapse3 introduces a time-varying effect, deviations here can hint at
interactions between premium level and the dynamic covariate structure that are
not explicitly modeled. Persistent bias would be a signal to explore extending
the time-varying component (e.g., premium strata interactions) in future work.

```{r lapse3_filter_low_premium_policies}
#| echo: true

lapse3_sim_monthly_cashflow_lowprem_tbl <- lapse3_sim_cashflow_tbl |>
  semi_join(lowprem_inforce_policy_ids_tbl, by = "policy_id") |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )
```

```{r lapse3_construct_comparison_plot_lowprem}
#| echo: true

ggplot(lapse3_sim_monthly_cashflow_lowprem_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_lowprem_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title = glue("Simulated vs Actual Cashflows (Low Premium) for Model {lapse3_stanmodel_name}"),
    subtitle = "Policies with prem_ape < 12k"
    )
```

```{r lapse3_construct_comparison_plot_ribbons_lowprem}
#| echo: true

lapse3_sim_monthly_bands_lowprem_tbl <- lapse3_sim_monthly_cashflow_lowprem_tbl |>
  summarise(
    p10 = quantile(total_cashflow, 0.10),
    p25 = quantile(total_cashflow, 0.25),
    p50 = quantile(total_cashflow, 0.50),
    p75 = quantile(total_cashflow, 0.75),
    p90 = quantile(total_cashflow, 0.90),
    
    .by = c(cashflow_date)
    )

ggplot(lapse3_sim_monthly_bands_lowprem_tbl) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p10, ymax = p90),
    alpha = 0.5
    ) +
  geom_ribbon(
    aes(x = cashflow_date, ymin = p25, ymax = p75),
    alpha = 0.5
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_lowprem_tbl,
    colour = "red"
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Time",
    y = "Cashflow",
    title    = glue("Simulated vs Actual Cashflows (Low Premium) for Model {lapse3_stanmodel_name}"),
    subtitle = "Policies with prem_ape < 12k, showing 50% and 80% bands"
    )
```

Interpretation parallels the previous model: we look for systematic drift of
the actual (red) path relative to the central band. If lapse3 (with its
time-varying smoker effect) reduces divergence versus lapse2 for the same
low-premium slice, that supports retaining the dynamic term. If performance is
similar, added complexity may not yield incremental calibration benefits here.


Across all three models we now hold comparable visual diagnostics (full, recent,
low-premium). A subsequent comparative summary could compute calibration error
metrics (e.g., actual within central band proportion) to objectively rank
specifications—left for future enhancement.



# Chunk Timing Summary

Execution timing offers a performance profile for incremental optimization. The
slowest chunks typically involve survival extraction and cashflow generation;
these are prime candidates for parallel refinement or reduced draw counts in
iterative development cycles.

This section summarizes the execution time for all chunks in this notebook.

```{r chunk_timing_summary}
#| echo: false

# Create timing summary table
timing_summary_tbl <- tibble(
    chunk_label = names(chunk_times),
    seconds     = unlist(chunk_times)
    ) |>
  arrange(desc(seconds)) |>
  mutate(
    minutes = round(seconds / 60, 2),
    hours = round(seconds / 3600, 2),
    formatted = case_when(
      seconds < 60 ~ glue("{round(seconds, 1)} sec"),
      seconds < 3600 ~ glue("{minutes} min"),
      TRUE ~ glue("{hours} hr")
      )
    )

# Print summary
message("")
message(strrep("=", 80))
message("CHUNK EXECUTION TIMES (slowest first)")
message(strrep("=", 80))
message("")

timing_summary_tbl |>
  select(chunk_label, formatted) |>
  print(n = Inf)

message("")
message(strrep("-", 80))
total_time <- sum(timing_summary_tbl$seconds)
message(glue("Total execution time: {round(total_time/60, 2)} minutes ({round(total_time, 1)} seconds)"))
message(strrep("=", 80))

# Save to file for future reference
dir_create("chunk_timings", recurse = TRUE)
timing_output_file <- glue("chunk_timings/chunk_timings_{format(Sys.time(), '%Y%m%d_%H%M%S')}.parquet")
timing_summary_tbl |> write_parquet_compressed(timing_output_file)
message(glue("Timing data saved to: {timing_output_file}"))
```


