---
title: "Initial Exploration of the P/NBD Model"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    light: superhero
    dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
---


```{r import_libraries}
#| echo: false
#| message: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(arrow)
library(fs)
library(purrr)
library(furrr)
library(glue)
library(cmdstanr)
library(brms)
library(posterior)
library(bayesplot)
library(tidybayes)
library(splines2)
library(rstanarm)


source("lib_utils.R")
source("lib_brms_hazard.R")
source("lib_survival_modelling.R")
source("lib_cashflow_projection.R")


conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "MASS",
    "fitdistrplus")
  )


options(
  width = 80L,
  warn  = 1,
  mc.cores = parallelly::availableCores()
  )

stan_modeldir <- "stan_model"
stan_codedir  <- "stan_code"


set.seed(42)
stanfit_seed <- 4000

theme_set(theme_cowplot())

plan(multisession)
```



# Load Data

We first load up the data we are using to train our dataset, as well as the
'rolled-forward' data that observes the same data a year later.

```{r load_modelling_data}
#| echo: true

model_training_tbl <- read_parquet("data/model_training_small_tbl.parquet")

model_training_tbl |> glimpse()
```

We then load up the 'current' data that we use to validate our model.


```{r load_current_data}
#| echo: true

model_rollforward_tbl <- read_parquet("data/model_rollforward_small_tbl.parquet")

model_rollforward_tbl |> glimpse()
```


## Create Shared Data Objects

These objects are used across all three model iterations and are created once
here to avoid duplication.

```{r create_monthly_dates}
#| echo: true

monthly_tbl <- tibble(
  date = seq(as.Date("2015-01-01"), as.Date("2016-02-01"), by = "month"),
  weeks = difftime(date, as.Date("2015-01-01"), units = "weeks") |> as.numeric()
  )

monthly_tbl |> glimpse()
```


```{r create_inforce_subset}
#| echo: true

inforce_tbl <- model_training_tbl |>
  filter(status_at_reference_date == "inforce")

message(glue("Inforce policies: {nrow(inforce_tbl)}"))
```


```{r create_policy_cutdown}
#| echo: true

policy_cutdown_tbl <- model_training_tbl |>
  select(policy_id, policy_startdate, policy_enddate, prem_ape)

policy_cutdown_tbl |> glimpse()
```


```{r extract_cashflow_week_count}
#| echo: true

cashflow_week_count <- monthly_tbl |> pull(weeks)

message(glue("Number of cashflow periods: {length(cashflow_week_count)}"))
```


```{r calculate_actual_cashflows}
#| echo: true

actual_cashflows_tbl <- model_rollforward_tbl |>
  semi_join(inforce_tbl, by = "policy_id") |>
  calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date)

actual_cashflows_tbl |> glimpse()
```


```{r summarise_actual_cashflows}
#| echo: true

actual_monthly_tbl <- actual_cashflows_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )

actual_monthly_tbl |> glimpse()
```



# Build Initial Cox Model


## brms Model

```{r lapse1_fit_cox_model_brms}
#| echo: true

lapse1_stanmodel_name <- "lapse1_coxph"

lapse1_coxph_brmsfit <- brm(
  policy_lifetime | cens(1-lapsed) ~ gender_life1 + smoker_life1,
  data            = model_training_tbl,
  family          = cox(link = "log"),
  cores           = 4,
  seed            = 421,
  save_warmup     = FALSE,
  backend         = "cmdstanr",
  save_model      = glue("{stan_codedir}/{lapse1_stanmodel_name}.stan"),
  file            = glue("{stan_modeldir}/{lapse1_stanmodel_name}_brmsfit"),
  output_dir      = stan_modeldir,
  output_basename = lapse1_stanmodel_name,

  stan_model_args = list(
    include_paths   =     "stan_code",
    pedantic        =            TRUE,
    dir             =    "stan_model"
    )
  )

summary(lapse1_coxph_brmsfit)
```


## rstanarm Model


```{r lapse1_fit_cox_model_rstanarm}
#| echo: true
#| message: false
#| output: false

lapse1_stanmodel_name <- "lapse1_coxph_stansurv"
lapse1_stanmodel_file <- glue("{stan_modeldir}/{lapse1_stanmodel_name}.rds")

# Check if fitted model exists, load it if so, otherwise fit new model
if (file_exists(lapse1_stanmodel_file)) {
  message(glue("Loading cached model from {lapse1_stanmodel_file}"))
  lapse1_coxph_stansurv <- read_rds(lapse1_stanmodel_file)

} else {
  message(glue("Fitting new model, will save to {lapse1_stanmodel_file}"))
  
  lapse1_coxph_stansurv <- stan_surv(
    Surv(policy_lifetime, lapsed) ~ gender_life1 + smoker_life1,
    data   = model_training_tbl,
    
    # MCMC sampling parameters
    chains = 4,
    iter   = 2000,
    warmup = 1000,
    seed   = 4000,
    cores  = 4,
    
    # Baseline hazard specification
    basehaz     = "ms",  # M-splines (flexible, default)
    basehaz_ops = list(df = 6),  # Degrees of freedom for baseline hazard
    
    # Prior specifications (explicit is better than implicit)
    prior           = rstanarm::normal(location = 0, scale = 2.5), # Weakly informative for coefficients
    prior_intercept = rstanarm::normal(location = 0, scale = 10),  # For baseline hazard
    
    # Enable verbose output for console monitoring (not captured in document)
    refresh = 200  # Print progress every 200 iterations
    )
  
  # Save fitted model for future use
  write_rds(lapse1_coxph_stansurv, lapse1_stanmodel_file)
  message(glue("Model saved to {lapse1_stanmodel_file}"))
}

summary(lapse1_coxph_stansurv)
```


# Construct Posterior Survival Estimates

We now want to use our fitted model to predict the lapse for the portfolio.

We model the monthly paid premia in the portfolio for the final year for which
we have data. We can then compare our modelled cashflows against the observed
data.


## Extract Posterior Estimates


```{r lapse1_construct_posterior_estimates}
#| echo: true

lapse1_post_draws_lst <- posterior_survfit(
    lapse1_coxph_stansurv,
    newdata       = inforce_tbl,
    last_time     = NULL,
    draws         = 2000,
    condition     = TRUE,
    return_matrix = TRUE
    )

lapse1_post_draws_lst[[1]] |> glimpse()
```


```{r lapse1_extract_posterior_survival_estimates}
#| echo: true

lapse1_post_survival_tbl <- lapse1_post_draws_lst |>
  map_dfr(
    extract_stansurv_posterior_survivals,

    .progress = "extract_post_surv"
    ) |>
  filter(time < 5 * 52) |>
  mutate(
    policy_id = inforce_tbl$policy_id[idx],
    
    .before = 1
    ) |>
  select(policy_id, draw_id, time, post_surv_prob = value) |>
  interpolate_survival_data(
    new_times = cashflow_week_count
    ) |>
  group_nest(policy_id, draw_id)

lapse1_post_survival_tbl |> glimpse()
```



```{r lapse1_calculate_sim_lapse_times}
#| echo: true

lapse1_sim_lapse_times_tbl <- lapse1_post_survival_tbl |>
  mutate(
    rnd_draw   = runif(n(), 0, 1),
    lapse_time = map2_dbl(
      data, rnd_draw,
      \(x, y) extract_survival_time(x, y, min_time = 1),

      .progress = "extract_survival_time"
      )
    ) |>
  select(policy_id, draw_id, lapse_time)

lapse1_sim_lapse_times_tbl |> glimpse()
```



```{r lapse1_construct_policy_sim_data}
#| echo: true

lapse1_sim_policy_data_tbl <- lapse1_sim_lapse_times_tbl |>
  inner_join(policy_cutdown_tbl, by = "policy_id") |>
  mutate(
    lapse_date    = as.Date("2015-01-01") + weeks(lapse_time),
    policy_status = case_when(

      # Policy has expired (term ended)
      (policy_enddate <= lapse_date) & (as.Date("2016-01-01") > policy_enddate) ~ "completed",

      # Policy has lapsed and the lapse occurred before or on the reference date
      (lapse_date < policy_enddate) & (lapse_date <= as.Date("2016-01-01")) ~ "lapsed",

      # Policy is within term and either not lapsed or lapsed after reference date
      as.Date("2016-01-01") <= policy_enddate ~ "inforce",

      # Default case (shouldn't occur with proper data)
      TRUE ~ "unknown"
      ),
    policy_statuschangedate = case_when(
      policy_status == "lapsed"    ~ lapse_date,
      policy_status == "completed" ~ policy_enddate,
      policy_status == "inforce"   ~ policy_startdate,
      TRUE                         ~ as.Date(NA)
      )
    )


lapse1_sim_policy_data_tbl |> glimpse()
```


## Construct Simulated Cashflow Data


```{r lapse1_construct_simulated_cashflows}
#| echo: true

lapse1_sim_cashflow_tbl <- lapse1_sim_policy_data_tbl |>
  group_nest(draw_id) |>
  mutate(
    newdata = map(
      data, \(x) x |> calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date),
                  
      .progress = "joining_data")
      ) |>
  select(-data) |>
  unnest(newdata)

lapse1_sim_cashflow_tbl |> glimpse()
```


```{r lapse1_construct_simulated_monthly_cashflows}
#| echo: true

lapse1_sim_monthly_cashflow_tbl <- lapse1_sim_cashflow_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )

lapse1_sim_monthly_cashflow_tbl |> glimpse()
```


```{r lapse1_construct_comparison_plot}
#| echo: true

ggplot(lapse1_sim_monthly_cashflow_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_tbl,
    colour = "red"
    )
```


We now clean up the environment before building the next model.

```{r lapse1_clear_large_objects}
#| echo: false

# Save tibbles before removing
lapse1_post_survival_tbl |> write_parquet(glue("{stan_modeldir}/lapse1_post_survival_tbl.parquet"))
lapse1_sim_cashflow_tbl  |> write_parquet(glue("{stan_modeldir}/lapse1_sim_cashflow_tbl.parquet"))

# Note: lapse1_post_draws_lst is a list and cannot be saved as parquet
# Save as RDS for now - consider alternative storage if needed
lapse1_post_draws_lst |> write_rds(glue("{stan_modeldir}/lapse1_post_draws_lst.rds"))

# Remove large objects
rm(lapse1_post_survival_tbl, lapse1_post_draws_lst, lapse1_sim_cashflow_tbl)

gc(verbose = TRUE)
```


# Build Expanded Cox Model

We now want to add a number of covariates to the model to see if we can improve
it.

## rstanarm Model


```{r lapse2_fit_cox_model_rstanarm}
#| echo: true
#| message: false
#| output: false

lapse2_stanmodel_name <- "lapse2_coxph_stansurv"
lapse2_stanmodel_file <- glue("{stan_modeldir}/{lapse2_stanmodel_name}.rds")

# Check if fitted model exists, load it if so, otherwise fit new model
if (file_exists(lapse2_stanmodel_file)) {
  message(glue("Loading cached model from {lapse2_stanmodel_file}"))
  lapse2_coxph_stansurv <- read_rds(lapse2_stanmodel_file)

} else {
  message(glue("Fitting new model, will save to {lapse2_stanmodel_file}"))

  # Create stan_model directory if it doesn't exist
  dir_create(stan_modeldir, recurse = TRUE)
  
  lapse2_coxph_stansurv <- stan_surv(
    Surv(policy_lifetime, lapsed) ~ gender_life1 + smoker_life1 + cluster_id + prem_sa_ratio,
    data   = model_training_tbl,
    
    # MCMC sampling parameters
    chains = 4,
    iter   = 2000,
    warmup = 1000,
    seed   = 4000,
    cores  = 4,
    
    # Baseline hazard specification
    basehaz     = "ms",  # M-splines (flexible, default)
    basehaz_ops = list(df = 6),  # Degrees of freedom for baseline hazard
    
    # Prior specifications (explicit is better than implicit)
    prior           = rstanarm::normal(location = 0, scale = 2.5), # Weakly informative for coefficients
    prior_intercept = rstanarm::normal(location = 0, scale = 10),  # For baseline hazard
    
    # Enable verbose output for console monitoring (not captured in document)
    refresh = 200  # Print progress every 200 iterations
    )
  
  # Save fitted model for future use
  write_rds(lapse2_coxph_stansurv, lapse2_stanmodel_file)
  message(glue("Model saved to {lapse2_stanmodel_file}"))
}

summary(lapse2_coxph_stansurv)
```


## Construct Posterior Survival Estimates

We now want to use our fitted model to predict the lapse for the portfolio.

We model the monthly paid premia in the portfolio for the final year for which
we have data. We can then compare our modelled cashflows against the observed
data.

```{r lapse2_construct_rstanarm_posterior_estimates}
#| echo: true

lapse2_post_draws_lst <- posterior_survfit(
    lapse2_coxph_stansurv,
    newdata       = inforce_tbl,
    last_time     = NULL,
    draws         = 2000,
    condition     = TRUE,
    return_matrix = TRUE
    )

lapse2_post_draws_lst[[1]] |> glimpse()
```


```{r lapse2_extract_posterior_survival_estimates}
#| echo: true

lapse2_cashflow_week_count <- monthly_tbl |> pull(weeks)

lapse2_post_survival_tbl <- lapse2_post_draws_lst |>
  map_dfr(
    extract_stansurv_posterior_survivals,

    .progress = "extract_post_surv"
    ) |>
  filter(time < 5 * 52) |>
  mutate(
    policy_id = inforce_tbl$policy_id[idx],
    
    .before = 1
    ) |>
  select(policy_id, draw_id, time, post_surv_prob = value) |>
  interpolate_survival_data(
    new_times = lapse2_cashflow_week_count
    ) |>
  group_nest(policy_id, draw_id)

lapse2_post_survival_tbl |> glimpse()
```



```{r lapse2_calculate_sim_lapse_times}
#| echo: true

lapse2_sim_lapse_times_tbl <- lapse2_post_survival_tbl |>
  mutate(
    rnd_draw   = runif(n(), 0, 1),
    lapse_time = map2_dbl(
      data, rnd_draw,
      \(x, y) extract_survival_time(x, y, min_time = 1),

      .progress = "extract_survival_time"
      )
    ) |>
  select(policy_id, draw_id, lapse_time)

lapse2_sim_lapse_times_tbl |> glimpse()
```



```{r lapse2_construct_policy_sim_data}
#| echo: true

lapse2_sim_policy_data_tbl <- lapse2_sim_lapse_times_tbl |>
  inner_join(policy_cutdown_tbl, by = "policy_id") |>
  mutate(
    lapse_date    = as.Date("2015-01-01") + weeks(lapse_time),
    policy_status = case_when(

      # Policy has expired (term ended)
      (policy_enddate <= lapse_date) & (as.Date("2016-01-01") > policy_enddate) ~ "completed",

      # Policy has lapsed and the lapse occurred before or on the reference date
      (lapse_date < policy_enddate) & (lapse_date <= as.Date("2016-01-01")) ~ "lapsed",

      # Policy is within term and either not lapsed or lapsed after reference date
      as.Date("2016-01-01") <= policy_enddate ~ "inforce",

      # Default case (shouldn't occur with proper data)
      TRUE ~ "unknown"
      ),
    policy_statuschangedate = case_when(
      policy_status == "lapsed"    ~ lapse_date,
      policy_status == "completed" ~ policy_enddate,
      policy_status == "inforce"   ~ policy_startdate,
      TRUE                         ~ as.Date(NA)
      )
    )


lapse2_sim_policy_data_tbl |> glimpse()
```



We now construct the simulated cashflows from the simulated data.


```{r lapse2_construct_simulated_cashflows}
#| echo: true

lapse2_sim_cashflow_tbl <- lapse2_sim_policy_data_tbl |>
  group_nest(draw_id) |>
  mutate(
    newdata = map(
      data, \(x) x |> calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date),
                  
      .progress = "joining_data")
      ) |>
  select(-data) |>
  unnest(newdata)

lapse2_sim_cashflow_tbl |> glimpse()
```


```{r lapse2_construct_simulated_monthly_cashflows}
#| echo: true

lapse2_sim_monthly_cashflow_tbl <- lapse2_sim_cashflow_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )

lapse2_sim_monthly_cashflow_tbl |> glimpse()
```


```{r lapse2_construct_comparison_plot}
#| echo: true

ggplot(lapse2_sim_monthly_cashflow_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_tbl,
    colour = "red"
    )
```


We now clean up the environment before building the next model.

```{r lapse2_clear_large_objects}
#| echo: false

# Save tibbles before removing
lapse2_post_survival_tbl |> write_parquet(glue("{stan_modeldir}/lapse2_post_survival_tbl.parquet"))
lapse2_sim_cashflow_tbl  |> write_parquet(glue("{stan_modeldir}/lapse2_sim_cashflow_tbl.parquet"))

# Note: lapse2_post_draws_lst is a list and cannot be saved as parquet
# Save as RDS for now - consider alternative storage if needed
lapse2_post_draws_lst |> write_rds(glue("{stan_modeldir}/lapse2_post_draws_lst.rds"))

# Remove large objects
rm(lapse2_post_survival_tbl, lapse2_post_draws_lst, lapse2_sim_cashflow_tbl)

gc(verbose = TRUE)
```




# Add Time-Dependent Parameters

We now want to add a number of covariates to the model to see if we can improve
it.

## rstanarm Model


```{r lapse3_fit_cox_model_rstanarm}
#| echo: true
#| #| message: false
#| output: false

lapse3_stanmodel_name <- "lapse3_coxph_stansurv"
lapse3_stanmodel_file <- glue("{stan_modeldir}/{lapse3_stanmodel_name}.rds")

# Check if fitted model exists, load it if so, otherwise fit new model
if (file_exists(lapse3_stanmodel_file)) {
  message(glue("Loading cached model from {lapse3_stanmodel_file}"))
  lapse3_coxph_stansurv <- read_rds(lapse3_stanmodel_file)

} else {
  message(glue("Fitting new model, will save to {lapse3_stanmodel_file}"))

  # Create stan_model directory if it doesn't exist
  dir_create(stan_modeldir, recurse = TRUE)
  
  lapse3_coxph_stansurv <- stan_surv(
    Surv(policy_lifetime, lapsed) ~ gender_life1 + tve(smoker_life1, df = 4, degree = 2),
    data   = model_training_tbl,
    
    # MCMC sampling parameters
    chains = 4,
    iter   = 2000,
    warmup = 1000,
    seed   = 4000,
    cores  = 4,
    
    # Baseline hazard specification
    basehaz     = "ms",  # M-splines (flexible, default)
    basehaz_ops = list(df = 6),  # Degrees of freedom for baseline hazard
    
    # Prior specifications (explicit is better than implicit)
    prior           = rstanarm::normal(location = 0, scale = 2.5), # Weakly informative for coefficients
    prior_intercept = rstanarm::normal(location = 0, scale = 10),  # For baseline hazard
    
    # Enable verbose output for console monitoring (not captured in document)
    refresh = 200  # Print progress every 200 iterations
    )
  
  # Save fitted model for future use
  write_rds(lapse3_coxph_stansurv, lapse3_stanmodel_file)
  message(glue("Model saved to {lapse3_stanmodel_file}"))
}

summary(lapse3_coxph_stansurv)
```


## Construct Posterior Survival Estimates

We now want to use our fitted model to predict the lapse for the portfolio.

We model the monthly paid premia in the portfolio for the final year for which
we have data. We can then compare our modelled cashflows against the observed
data.

```{r lapse3_construct_rstanarm_posterior_estimates}
#| echo: true

lapse3_post_draws_lst <- posterior_survfit(
    lapse3_coxph_stansurv,
    newdata       = inforce_tbl,
    last_time     = NULL,
    draws         = 2000,
    condition     = TRUE,
    return_matrix = TRUE
    )

lapse3_post_draws_lst[[1]] |> glimpse()
```


```{r lapse3_extract_posterior_survival_estimates}
#| echo: true

lapse3_cashflow_week_count <- monthly_tbl |> pull(weeks)

lapse3_post_survival_tbl <- lapse3_post_draws_lst |>
  map_dfr(
    extract_stansurv_posterior_survivals,

    .progress = "extract_post_surv"
    ) |>
  filter(time < 5 * 52) |>
  mutate(
    policy_id = inforce_tbl$policy_id[idx],
    
    .before = 1
    ) |>
  select(policy_id, draw_id, time, post_surv_prob = value) |>
  interpolate_survival_data(
    new_times = lapse3_cashflow_week_count
    ) |>
  group_nest(policy_id, draw_id)

lapse3_post_survival_tbl |> glimpse()
```



```{r lapse3_calculate_sim_lapse_times}
#| echo: true

lapse3_sim_lapse_times_tbl <- lapse3_post_survival_tbl |>
  mutate(
    rnd_draw   = runif(n(), 0, 1),
    lapse_time = map2_dbl(
      data, rnd_draw,
      \(x, y) extract_survival_time(x, y, min_time = 1),

      .progress = "extract_survival_time"
      )
    ) |>
  select(policy_id, draw_id, lapse_time)

lapse3_sim_lapse_times_tbl |> glimpse()
```



```{r lapse3_construct_policy_sim_data}
#| echo: true

lapse3_sim_policy_data_tbl <- lapse3_sim_lapse_times_tbl |>
  inner_join(policy_cutdown_tbl, by = "policy_id") |>
  mutate(
    lapse_date    = as.Date("2015-01-01") + weeks(lapse_time),
    policy_status = case_when(

      # Policy has expired (term ended)
      (policy_enddate <= lapse_date) & (as.Date("2016-01-01") > policy_enddate) ~ "completed",

      # Policy has lapsed and the lapse occurred before or on the reference date
      (lapse_date < policy_enddate) & (lapse_date <= as.Date("2016-01-01")) ~ "lapsed",

      # Policy is within term and either not lapsed or lapsed after reference date
      as.Date("2016-01-01") <= policy_enddate ~ "inforce",

      # Default case (shouldn't occur with proper data)
      TRUE ~ "unknown"
      ),
    policy_statuschangedate = case_when(
      policy_status == "lapsed"    ~ lapse_date,
      policy_status == "completed" ~ policy_enddate,
      policy_status == "inforce"   ~ policy_startdate,
      TRUE                         ~ as.Date(NA)
      )
    )


lapse3_sim_policy_data_tbl |> glimpse()
```



We now construct the simulated cashflows from the simulated data.


```{r lapse3_construct_simulated_cashflows}
#| echo: true

lapse3_sim_cashflow_tbl <- lapse3_sim_policy_data_tbl |>
  group_nest(draw_id) |>
  mutate(
    newdata = map(
      data, \(x) x |> calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date),

      .progress = "joining_data")
      ) |>
  select(-data) |>
  unnest(newdata)

lapse3_sim_cashflow_tbl |> glimpse()
```


```{r lapse3_construct_simulated_monthly_cashflows}
#| echo: true

lapse3_sim_monthly_cashflow_tbl <- lapse3_sim_cashflow_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )

lapse3_sim_monthly_cashflow_tbl |> glimpse()
```


```{r lapse3_construct_comparison_plot}
#| echo: true

ggplot(lapse3_sim_monthly_cashflow_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_tbl,
    colour = "red"
    )
```

