---
title: "Initial Exploration of the P/NBD Model"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    light: superhero
    dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
---


```{r import_libraries}
#| echo: false
#| message: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(arrow)
library(fs)
library(purrr)
library(furrr)
library(glue)
library(cmdstanr)
library(brms)
library(posterior)
library(bayesplot)
library(tidybayes)
library(splines2)
library(rstanarm)


source("lib_utils.R")
source("lib_brms_hazard.R")
source("lib_survival_modelling.R")
source("lib_cashflow_projection.R")


conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "MASS",
    "fitdistrplus")
  )


options(
  width = 80L,
  warn  = 1,
  mc.cores = parallelly::availableCores()
  )

stan_modeldir <- "stan_model"
stan_codedir  <- "stan_code"


set.seed(42)
stanfit_seed <- 4000

theme_set(theme_cowplot())

plan(multisession)
```



# Load Data

We first load up the data we are using to train our dataset, as well as the
'rolled-forward' data that observes the same data a year later.

```{r load_modelling_data}
#| echo: true

model_training_tbl <- read_parquet("data/model_training_small_tbl.parquet")

model_training_tbl |> glimpse()
```

We then load up the 'current' data that we use to validate our model.


```{r load_current_data}
#| echo: true

model_rollforward_tbl <- read_parquet("data/model_rollforward_small_tbl.parquet")

model_rollforward_tbl |> glimpse()
```



# Build Initial Cox Model


## brms Model

```{r build_initial_cox_model_brms}
#| echo: true

stanmodel_name <- "lapse1_coxph"

lapse1_coxph_brmsfit <- brm(
  policy_lifetime | cens(1-lapsed) ~ gender_life1 + smoker_life1,
  data            = model_training_tbl,
  family          = cox(link = "log"),
  cores           = 4,
  seed            = 421,
  save_warmup     = FALSE,
  backend         = "cmdstanr",
  save_model      = glue("{stan_codedir}/{stanmodel_name}.stan"),
  file            = glue("{stan_modeldir}/{stanmodel_name}_brmsfit"),
  output_dir      = stan_modeldir,
  output_basename = stanmodel_name,

  stan_model_args = list(
    include_paths   =     "stan_code",
    pedantic        =            TRUE,
    dir             =    "stan_model"
    )
  )

summary(lapse1_coxph_brmsfit)
```


## rstanarm Model


```{r build_initial_cox_model_rstanarm}
#| echo: true

stanmodel_name <- "lapse1_coxph_stansurv"
stanmodel_file <- glue("{stan_modeldir}/{stanmodel_name}.rds")

# Check if fitted model exists, load it if so, otherwise fit new model
if (file_exists(stanmodel_file)) {
  write_lines(glue("Loading cached model from {stanmodel_file}"), stdout())
  lapse1_coxph_stansurv <- read_rds(stanmodel_file)
} else {
  write_lines(glue("Fitting new model, will save to {stanmodel_file}"), stdout())
  
  # Create stan_model directory if it doesn't exist
  dir_create(stan_modeldir, recurse = TRUE)
  
  lapse1_coxph_stansurv <- stan_surv(
    Surv(policy_lifetime, lapsed) ~ gender_life1 + smoker_life1,
    data   = model_training_tbl,
    
    # MCMC sampling parameters
    chains = 4,
    iter   = 2000,
    warmup = 1000,
    seed   = 4000,
    cores  = 4,
    
    # Baseline hazard specification
    basehaz     = "ms",  # M-splines (flexible, default)
    basehaz_ops = list(df = 6),  # Degrees of freedom for baseline hazard
    
    # Prior specifications (explicit is better than implicit)
    prior           = rstanarm::normal(location = 0, scale = 2.5),  # Weakly informative for coefficients
    prior_intercept = rstanarm::normal(location = 0, scale = 10),  # For baseline hazard
    )
  
  # Save fitted model for future use
  write_rds(lapse1_coxph_stansurv, stanmodel_file)
  write_lines(glue("Model saved to {stanmodel_file}"), stdout())
}

summary(lapse1_coxph_stansurv)
```


# Construct Posterior Survival Estimates

We now want to use our fitted model to predict the lapse for the portfolio.

We model the monthly paid premia in the portfolio for the final year for which
we have data. We can then compare our modelled cashflows against the observed
data.

## Set Up Monthly Dates

```{r create_monthly_estimates}
#| echo: true

monthly_tbl <- tibble(
  date = seq(as.Date("2015-01-01"), as.Date("2016-02-01"), by = "month"),
  weeks = difftime(date, as.Date("2015-01-01"), units = "weeks") |> as.numeric()
  )

monthly_tbl |> glimpse()
```





## Extract Posterior Estimates


```{r construct_brms_posterior_estimates}
#| echo: true

inforce_tbl <- model_training_tbl |>
  filter(status_at_reference_date  == "inforce")

post_draws_lst <- posterior_survfit(
    lapse1_coxph_stansurv,
    newdata       = inforce_tbl,
    last_time     = NULL,
    draws         = 2000,
    condition     = TRUE,
    return_matrix = TRUE
    )

post_draws_lst[[1]] |> glimpse()
```


```{r extract_posterior_survival_estimates}
#| echo: true

cashflow_week_count <- monthly_tbl |> pull(weeks)

post_survival_tbl <- post_draws_lst |>
  map_dfr(
    extract_stansurv_posterior_survivals,

    .progress = "extract_post_surv"
    ) |>
  filter(time < 5 * 52) |>
  mutate(
    policy_id = inforce_tbl$policy_id[idx],
    
    .before = 1
    ) |>
  select(policy_id, draw_id, time, post_surv_prob = value) |>
  interpolate_survival_data(
    new_times = cashflow_week_count
    ) |>
  group_nest(policy_id, draw_id)

post_survival_tbl |> glimpse()
```



```{r calculate_sim_lapse_times}
#| echo: true

sim_lapse_times_tbl <- post_survival_tbl |>
  mutate(
    rnd_draw   = runif(n(), 0, 1),
    lapse_time = map2_dbl(
      data, rnd_draw,
      \(x, y) extract_survival_time(x, y, min_time = 1),

      .progress = "extract_survival_time"
      )
    ) |>
  select(policy_id, draw_id, lapse_time)

sim_lapse_times_tbl |> glimpse()
```



```{r construct_policy_sim_data}
#| echo: true

policy_cutdown_tbl <- model_training_tbl |>
  select(policy_id, policy_startdate, policy_enddate, prem_ape)

sim_policy_data_tbl <- sim_lapse_times_tbl |>
  inner_join(policy_cutdown_tbl, by = "policy_id") |>
  mutate(
    lapse_date    = as.Date("2015-01-01") + weeks(lapse_time),
    policy_status = case_when(

      # Policy has expired (term ended)
      (policy_enddate <= lapse_date) & (as.Date("2016-01-01") > policy_enddate) ~ "completed",

      # Policy has lapsed and the lapse occurred before or on the reference date
      (lapse_date < policy_enddate) & (lapse_date <= as.Date("2016-01-01")) ~ "lapsed",

      # Policy is within term and either not lapsed or lapsed after reference date
      as.Date("2016-01-01") <= policy_enddate ~ "inforce",

      # Default case (shouldn't occur with proper data)
      TRUE ~ "unknown"
      ),
    policy_statuschangedate = case_when(
      policy_status == "lapsed"    ~ lapse_date,
      policy_status == "completed" ~ policy_enddate,
      policy_status == "inforce"   ~ policy_startdate,
      TRUE                         ~ as.Date(NA)
      )
    )


sim_policy_data_tbl |> glimpse()
```


## Construct Actual Cashflow Data

```{r construct_monthly_cashflows}
#| echo: true

actual_cashflows_tbl <- model_rollforward_tbl |>
  semi_join(inforce_tbl, by = "policy_id") |>
  calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date)

actual_cashflows_tbl |> glimpse()
```

We now add the cashflows across all the policies to get the total monthly
cashflow observed in the data.

```{r construct_monthly_actual_cashflow}
#| echo: true

actual_monthly_tbl <- actual_cashflows_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )


actual_monthly_tbl |> glimpse()
```


We now construct the simulated cashflows from the simulated data.


```{r construct_simulated_cashflows}
#| echo: true

sim_cashflow_tbl <- sim_policy_data_tbl |>
  group_nest(draw_id) |>
  mutate(
    newdata = map(
      data, \(x) x |> calculate_monthly_cashflows(cashflow_dates = monthly_tbl$date),
                  
      .progress = "joining_data")
      ) |>
  select(-data) |>
  unnest(newdata)

sim_cashflow_tbl |> glimpse()
```


```{r construct_simulated_monthly_cashflows}
#| echo: true

sim_monthly_cashflow_tbl <- sim_cashflow_tbl |>
  summarise(
    total_cashflow = sum(cashflow),
    
    .by = c(draw_id, cashflow_date)
    ) |>
  filter(
    cashflow_date <= as.Date("2016-01-01")
    )

sim_monthly_cashflow_tbl |> glimpse()
```


```{r construct_comparison_plot}
#| echo: true

ggplot(sim_monthly_cashflow_tbl) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow, group = draw_id),
    alpha = 0.01
    ) +
  geom_line(
    aes(x = cashflow_date, y = total_cashflow),
    data = actual_monthly_tbl,
    colour = "red"
    )


```



